---
title: "Bivariate analysis of enteric fever and leptospirosis"
author: "Hazlienor"
date: "2023-04-14"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## prepare environment

load package

```{r}
library(sf)
library(tidyverse)
library(here)
library(janitor)
library(gtsummary)
library(DT)
library(stringr)
library(readxl)
library(broom)
library(tmap)
library(mapview)
library(lubridate)
library(maptools)
library(spatstat)
library(spdep)
library(gridExtra)
library(grid)
library(sparr)
```


## prepare data


read polygon data

```{r}
kel <- st_read(here("Map",
                    "kelantan.shp"))
```


read population data per mukim per year

```{r}
kel_mukim <- read_xlsx(here ("mukim.xlsx"))
```


merge population data to polygon

```{r}
kel_map <- merge(kel,kel_mukim,by.x="MUKIM", by.y="MUKIM", all.x=T, sort=F)
dim(kel_map)
class(kel_map)
st_crs(kel_map)
```


load linelisting 

```{r}
linelist2 <- read_xlsx(here ("linelist.xlsx"))
```


merge population data to polygon

```{r}
kel_map <- merge(kel,kel_mukim,by.x="MUKIM", by.y="MUKIM", all.x=T, sort=F)
dim(kel_map)
class(kel_map)
st_crs(kel_map)
```


list cases by year

```{r}
#extract enteric fever 2016-2022
linelist2 <- read_xlsx(here ("linelist.xlsx"))
listENT_kel <- filter(linelist2, Diagnosis=="Enteric_fever")  
listENT_kel16 <- filter(listENT_kel, `Tahun Daftar` =="2016")
listENT_kel17 <- filter(listENT_kel, `Tahun Daftar` =="2017")
listENT_kel18 <- filter(listENT_kel, `Tahun Daftar` =="2018")
listENT_kel19 <- filter(listENT_kel, `Tahun Daftar` =="2019")
listENT_kel20 <- filter(listENT_kel, `Tahun Daftar` =="2020")
listENT_kel21 <- filter(listENT_kel, `Tahun Daftar` =="2021")
listENT_kel22 <- filter(listENT_kel, `Tahun Daftar` =="2022")

#extract leptospirosis 2016-2022
linelist2 <- read_xlsx(here ("linelist.xlsx"))
listLEP_kel <- filter(linelist2, Diagnosis=="Leptospirosis")
LEP_kel16 <- filter(listLEP_kel, `Tahun Daftar` =="2016")
LEP_kel17 <- filter(listLEP_kel, `Tahun Daftar` =="2017")
LEP_kel18 <- filter(listLEP_kel, `Tahun Daftar` =="2018")
LEP_kel19 <- filter(listLEP_kel, `Tahun Daftar` =="2019")
LEP_kel20 <- filter(listLEP_kel, `Tahun Daftar` =="2020")
LEP_kel21 <- filter(listLEP_kel, `Tahun Daftar` =="2021")
LEP_kel22 <- filter(listLEP_kel, `Tahun Daftar` =="2022")
```


count cases per mukim

```{r}
# join point to polygon
polyENT_kel <- merge(kel_map, listENT_kel, by = c("MUKIM"))

# count cumulative enteric fever per mukim
count_ent <- polyENT_kel%>% 
  count(Daerah, MUKIM, AVR) %>% 
  ungroup
combined_dataent <- st_join(kel_map, count_ent, by = "MUKIM")

# count cumulative lep fever per mukim
polyLEP_kel <- merge(kel_map, listLEP_kel, by = c("MUKIM"))
count_lep <- polyLEP_kel%>% 
  count(Daerah, MUKIM, AVR) %>% 
  ungroup
combined_datalep <- st_join(kel_map, count_lep, by = "MUKIM")
```


convert data to spatial format

```{r}
kel_map.sp <- as(kel_map, "Spatial")
class(kel_map.sp)
```


transform to owin format

```{r}
kel_map.owin <- as(kel_map.sp, "owin")
class(kel_map.owin)
```


mark data with diagnosis (factor)

```{r}
#| cache = TRUE
# transform to sp
loc_both <- st_as_sf(linelist2, 
                     coords = c("Longitude (WGS)", "Latitude (WGS)"), 
                     crs = 4326)
loc_both <-st_transform(loc_both, 3168)
loc_both.sp <- as(loc_both, "Spatial")

# transform data sp to ppp - save categorical data as factor
all_diag.ppp <- as(loc_both.sp, "ppp")
marks(all_diag.ppp) <- as.factor(loc_both.sp$Diagnosis)
Window(all_diag.ppp) <- kel_map.owin
```

```{r}
#summary
summary(all_diag.ppp)
```


plot split data

```{r}
#| cache = TRUE
plot(split(all_diag.ppp))
plot(density(split(all_diag.ppp)), ribbon = FALSE)
```


## correlation of cases and population (visual)

plot by tmap 

```{r}
tmap_mode("plot")

cases_mapent <- tm_shape(combined_dataent) + tm_polygons("n") + tm_layout(main.title="Enteric Fever Cases")
pop_map <- tm_shape(kel_map) + tm_polygons("AVR") + tm_layout(main.title="Population")
cases_maplep <- tm_shape(combined_datalep) + tm_polygons("n") + tm_layout(main.title="Leptospirosis Cases")

tmap_arrange(cases_mapent, pop_map, cases_maplep,  ncol=3)   # arrange into 3x1 facets
```


plot by ggplot

```{r name-of-chunk8, fig.width=16, fig.height=9}
pop_map <- ggplot() +
    geom_sf(data = kel_map, aes(fill = AVR), na.rm = FALSE) +
    scale_fill_gradient(low = "white", high = "red", na.value = "grey") +
  labs(fill = "Population") +
  theme_void()

ENT_map <- ggplot() +
    geom_sf(data = combined_dataent, aes(fill = n), na.rm = FALSE) +
    scale_fill_gradient(low = "white", high = "red", na.value = "grey") +
  labs(fill = "Enteric Fever Cases") +
  theme_void()

LEP_map <- ggplot() +
    geom_sf(data = combined_datalep, aes(fill = n), na.rm = FALSE) +
    scale_fill_gradient(low = "white", high = "red", na.value = "grey") +
  labs(fill = "Leptospirosis Cases") +
  theme_void()

```

```{r name-of-chunk7, fig.width=16, fig.height=9}
grid.arrange(ENT_map, pop_map, LEP_map, nrow=1, top=textGrob("Comparison of Enteric Fever and Leptospirosis Cases to Population in Kelantan 2016-2022", gp = gpar(fontsize = 20)))
```


# spatial relative risk (sparr package)

cumulative cases

```{r}
#| cache = TRUE
# split data
diag_lep <- split(all_diag.ppp)$Leptospirosis
diag_ent <- split(all_diag.ppp)$Enteric_fever

# Compute a symmetric (pooled) adaptive relative risk estimate
# with tolerance contours
h0 <- OS(all_diag.ppp, nstar="geometric")
diag_rr <- risk(diag_lep, diag_ent, h0=h0, adapt=TRUE, tolerate=TRUE,
                hp=OS(all_diag.ppp)/2, pilot.symmetry="pooled", davies.baddeley=0.05)

plot(diag_rr)
```


prepare file for spatial relative risk by year

```{r}
#| cache = TRUE
# extract list contain both diagnosis by year
list2016 <- linelist2 %>% 
  filter(`Tahun Daftar` == "2016")
list2017 <- linelist2 %>% 
  filter(`Tahun Daftar` == "2017")
list2018 <- linelist2 %>% 
  filter(`Tahun Daftar` == "2018")
list2019 <- linelist2 %>% 
  filter(`Tahun Daftar` == "2019")
list2020 <- linelist2 %>% 
  filter(`Tahun Daftar` == "2020")
list2021 <- linelist2 %>% 
  filter(`Tahun Daftar` == "2021")
list2022 <- linelist2 %>% 
  filter(`Tahun Daftar` == "2022")

# transform to sf
list2016.sf <- st_as_sf(list2017, 
                     coords = c("Longitude (WGS)", "Latitude (WGS)"), 
                     crs = 4326)
list2017.sf <- st_as_sf(list2017, 
                     coords = c("Longitude (WGS)", "Latitude (WGS)"), 
                     crs = 4326)
list2018.sf <- st_as_sf(list2018, 
                     coords = c("Longitude (WGS)", "Latitude (WGS)"), 
                     crs = 4326)
list2019.sf <- st_as_sf(list2019, 
                     coords = c("Longitude (WGS)", "Latitude (WGS)"), 
                     crs = 4326)
list2020.sf <- st_as_sf(list2020, 
                     coords = c("Longitude (WGS)", "Latitude (WGS)"), 
                     crs = 4326)
list2021.sf <- st_as_sf(list2021, 
                     coords = c("Longitude (WGS)", "Latitude (WGS)"), 
                     crs = 4326)
list2022.sf <- st_as_sf(list2022, 
                     coords = c("Longitude (WGS)", "Latitude (WGS)"), 
                     crs = 4326)

# transform to RSO
loc_list2016.sf <-st_transform(list2016.sf, 3168)
loc_list2017.sf <-st_transform(list2017.sf, 3168)
loc_list2018.sf <-st_transform(list2018.sf, 3168)
loc_list2019.sf <-st_transform(list2019.sf, 3168)
loc_list2020.sf <-st_transform(list2020.sf, 3168)
loc_list2021.sf <-st_transform(list2021.sf, 3168)
loc_list2022.sf <-st_transform(list2022.sf, 3168)

# transform to sp
loc_list2016.sp <- as(loc_list2016.sf, "Spatial")
loc_list2017.sp <- as(loc_list2017.sf, "Spatial")
loc_list2018.sp <- as(loc_list2018.sf, "Spatial")
loc_list2019.sp <- as(loc_list2019.sf, "Spatial")
loc_list2020.sp <- as(loc_list2020.sf, "Spatial")
loc_list2021.sp <- as(loc_list2021.sf, "Spatial")
loc_list2022.sp <- as(loc_list2022.sf, "Spatial")

# transform data sp to ppp 
loc_list2016.ppp <- as(loc_list2016.sp, "ppp")
loc_list2017.ppp <- as(loc_list2017.sp, "ppp")
loc_list2018.ppp <- as(loc_list2018.sp, "ppp")
loc_list2019.ppp <- as(loc_list2019.sp, "ppp")
loc_list2020.ppp <- as(loc_list2020.sp, "ppp")
loc_list2021.ppp <- as(loc_list2021.sp, "ppp")
loc_list2022.ppp <- as(loc_list2022.sp, "ppp")

# save categorical data as factor (marked data)
marks(loc_list2016.ppp) <- as.factor(loc_list2016.sp$Diagnosis)
marks(loc_list2017.ppp) <- as.factor(loc_list2017.sp$Diagnosis)
marks(loc_list2018.ppp) <- as.factor(loc_list2018.sp$Diagnosis)
marks(loc_list2019.ppp) <- as.factor(loc_list2019.sp$Diagnosis)
marks(loc_list2020.ppp) <- as.factor(loc_list2020.sp$Diagnosis)
marks(loc_list2021.ppp) <- as.factor(loc_list2021.sp$Diagnosis)
marks(loc_list2022.ppp) <- as.factor(loc_list2022.sp$Diagnosis)

# set window
Window(loc_list2016.ppp) <- kel_map.owin
Window(loc_list2017.ppp) <- kel_map.owin
Window(loc_list2018.ppp) <- kel_map.owin
Window(loc_list2019.ppp) <- kel_map.owin
Window(loc_list2020.ppp) <- kel_map.owin
Window(loc_list2021.ppp) <- kel_map.owin
Window(loc_list2022.ppp) <- kel_map.owin
```


calculate spatial relative risk by year

```{r}
#| cache = TRUE
# 2016
# split into diagnosis
diag_lep16 <- split(loc_list2016.ppp)$Leptospirosis
diag_ent16 <- split(loc_list2016.ppp)$Enteric_fever

# Compute a symmetric (pooled) adaptive relative risk estimate
# with tolerance contours
h0_16 <- OS(loc_list2016.ppp, nstar="geometric")
diag_rr16 <- risk(diag_lep16, diag_ent16, h0=h0_16, adapt=TRUE, tolerate=TRUE,
                hp=OS(loc_list2016.ppp)/2, pilot.symmetry="pooled", davies.baddeley=0.05)

# 2017
diag_lep17 <- split(loc_list2017.ppp)$Leptospirosis
diag_ent17 <- split(loc_list2017.ppp)$Enteric_fever
h0_17 <- OS(loc_list2017.ppp, nstar="geometric")
diag_rr17 <- risk(diag_lep17, diag_ent17, h0=h0_17, adapt=TRUE, tolerate=TRUE,
                hp=OS(loc_list2017.ppp)/2, pilot.symmetry="pooled", davies.baddeley=0.05)

# 2018
diag_lep18 <- split(loc_list2018.ppp)$Leptospirosis
diag_ent18 <- split(loc_list2018.ppp)$Enteric_fever
h0_18 <- OS(loc_list2018.ppp, nstar="geometric")
diag_rr18 <- risk(diag_lep18, diag_ent18, h0=h0_18, adapt=TRUE, tolerate=TRUE,
                hp=OS(loc_list2018.ppp)/2, pilot.symmetry="pooled", davies.baddeley=0.05)

# 2019
diag_lep19 <- split(loc_list2019.ppp)$Leptospirosis
diag_ent19 <- split(loc_list2019.ppp)$Enteric_fever
h0_19 <- OS(loc_list2019.ppp, nstar="geometric")
diag_rr19 <- risk(diag_lep19, diag_ent19, h0=h0_19, adapt=TRUE, tolerate=TRUE,
                hp=OS(loc_list2019.ppp)/2, pilot.symmetry="pooled", davies.baddeley=0.05)

# 2020
diag_lep20 <- split(loc_list2020.ppp)$Leptospirosis
diag_ent20 <- split(loc_list2020.ppp)$Enteric_fever
h0_20 <- OS(loc_list2020.ppp, nstar="geometric")
diag_rr20 <- risk(diag_lep20, diag_ent20, h0=h0_20, adapt=TRUE, tolerate=TRUE,
                hp=OS(loc_list2020.ppp)/2, pilot.symmetry="pooled", davies.baddeley=0.05)

# 2021
diag_lep21 <- split(loc_list2021.ppp)$Leptospirosis
diag_ent21 <- split(loc_list2021.ppp)$Enteric_fever
h0_21 <- OS(loc_list2021.ppp, nstar="geometric")
diag_rr21 <- risk(diag_lep21, diag_ent21, h0=h0_21, adapt=TRUE, tolerate=TRUE,
                hp=OS(loc_list2021.ppp)/2, pilot.symmetry="pooled", davies.baddeley=0.05)

# 2022
diag_lep22 <- split(loc_list2022.ppp)$Leptospirosis
diag_ent22 <- split(loc_list2022.ppp)$Enteric_fever
h0_22 <- OS(loc_list2022.ppp, nstar="geometric")
diag_rr22 <- risk(diag_lep22, diag_ent22, h0=h0_22, adapt=TRUE, tolerate=TRUE,
                hp=OS(loc_list2022.ppp)/2, pilot.symmetry="pooled", davies.baddeley=0.05)
```


plot spatial relative risk by year

```{r name-of-chunk32, fig.width=15, fig.height=15}
#| cache = TRUE
par( mfrow= c(3,3) )
plot(diag_rr16, main = "2016", cex.main = 8, cex.lab = 6)
box(lwd = 2)
plot(diag_rr17, main = "2017", cex.main = 8, cex.lab = 6)
box(lwd = 2)
plot(diag_rr18, main = "2018", cex.main = 8, cex.lab = 6)
box(lwd = 2)
plot(diag_rr19, main = "2019", cex.main = 8, cex.lab = 6)
box(lwd = 2)
plot(diag_rr20, main = "2020", cex.main = 8, cex.lab = 6)
box(lwd = 2)
plot(diag_rr21, main = "2021", cex.main = 8, cex.lab = 6)
box(lwd = 2)
plot(diag_rr22, main = "2022", cex.main = 8, cex.lab = 6)
box(lwd = 2)
plot(diag_rr, main = "2016-2022", cex.main = 8, cex.lab = 6)
box(lwd = 2)
```


# spatially-varying probability (spatstat)

Estimates the spatially-varying probability of each type of point, using kernel smoothing
- A value of 0 for a particular label suggests no occurrence of that label at that location, while a value of 1 suggests a definite occurrence.
- the probability of the case to be leptospirosis case in that area as opposed to others, case = leptospirosis, control = enteric fever

```{r name-of-chunk32a, fig.width=15, fig.height=15}
# calculate risk
relrisk_all <- relrisk.ppp(all_diag.ppp, casecontrol=TRUE, at="pixels")
relrisk_16 <- relrisk.ppp(loc_list2016.ppp, casecontrol=TRUE, at="pixels")
relrisk_17 <- relrisk.ppp(loc_list2017.ppp, casecontrol=TRUE, at="pixels")
relrisk_18 <- relrisk.ppp(loc_list2018.ppp, casecontrol=TRUE, at="pixels")
relrisk_19 <- relrisk.ppp(loc_list2019.ppp, casecontrol=TRUE, at="pixels")
relrisk_20 <- relrisk.ppp(loc_list2020.ppp, casecontrol=TRUE, at="pixels")
relrisk_21 <- relrisk.ppp(loc_list2021.ppp, casecontrol=TRUE, at="pixels")
relrisk_22 <- relrisk.ppp(loc_list2022.ppp, casecontrol=TRUE, at="pixels")

par(mfrow=c(3,3))
plot(relrisk_16, main = "2016", cex.main = 8, cex.lab = 6)
box(lwd = 2)
plot(relrisk_17, main = "2017", cex.main = 8, cex.lab = 6)
box(lwd = 2)
plot(relrisk_18, main = "2018", cex.main = 8, cex.lab = 6)
box(lwd = 2)
plot(relrisk_19, main = "2019", cex.main = 8, cex.lab = 6)
box(lwd = 2)
plot(relrisk_20, main = "2020", cex.main = 8, cex.lab = 6)
box(lwd = 2)
plot(relrisk_21, main = "2021", cex.main = 8, cex.lab = 6)
box(lwd = 2)
plot(relrisk_22, main = "2022", cex.main = 8, cex.lab = 6)
box(lwd = 2)
plot(relrisk_all, main = "2016-2022", cex.main = 8, cex.lab = 6)
box(lwd = 2)
```

# bivariate L function 

calculate bivariate L function 

```{r}
#| cache = TRUE
Lboth <- alltypes(all_diag.ppp, "L")
```

```{r}
#| cache = TRUE
Lboth16 <- alltypes(loc_list2016.ppp, "L")
Lboth17 <- alltypes(loc_list2017.ppp, "L")
Lboth18 <- alltypes(loc_list2018.ppp, "L")
Lboth19 <- alltypes(loc_list2019.ppp, "L")
Lboth20 <- alltypes(loc_list2020.ppp, "L")
Lboth21 <- alltypes(loc_list2021.ppp, "L")
Lboth22 <- alltypes(loc_list2022.ppp, "L")
```


plot bivariate L graph by year

```{r name-of-chunk33, fig.width=20, fig.height=9}
plot(Lboth16, main = "2016", cex.main = 3, cex.lab = 1.5)
plot(Lboth17, main = "2017", cex.main = 3, cex.lab = 1.5)
plot(Lboth18, main = "2018", cex.main = 3, cex.lab = 1.5)
plot(Lboth19, main = "2019", cex.main = 3, cex.lab = 1.5)
plot(Lboth20, main = "2020", cex.main = 3, cex.lab = 1.5)
plot(Lboth21, main = "2021", cex.main = 3, cex.lab = 1.5)
plot(Lboth22, main = "2022", cex.main = 3, cex.lab = 1.5)
plot(Lboth, main = "2016-2022", cex.main = 3, cex.lab = 1.5)
```


```{r}
#| cache = TRUE
#plot estimates of the bivariate pair correlation functions
Lbothest <- plot(alltypes(all_diag.ppp, pcfcross))
```


```{r}
#| cache = TRUE
# bivariate G-functions
Gboth16 <- alltypes(loc_list2016.ppp, Gcross)
Gboth17 <- alltypes(loc_list2017.ppp, Gcross)
Gboth18 <- alltypes(loc_list2018.ppp, Gcross)
Gboth19 <- alltypes(loc_list2019.ppp, Gcross)
Gboth20 <- alltypes(loc_list2020.ppp, Gcross)
Gboth21 <- alltypes(loc_list2021.ppp, Gcross)
Gboth22 <- alltypes(loc_list2022.ppp, Gcross)
Gboth <- alltypes(all_diag.ppp, Gcross)
```


```{r name-of-chunk, fig.width=20, fig.height=9}
# plot bivariate G-functions
plot(Gboth16)
plot(Gboth17)
plot(Gboth18)
plot(Gboth19)
plot(Gboth20)
plot(Gboth21)
plot(Gboth22)
plot(Gboth)
```


# calculate Gdot function

```{r}
#| cache = TRUE
Gdotall <- alltypes(all_diag.ppp, Gdot)
Gdot16 <- alltypes(loc_list2016.ppp, Gdot)
Gdot17 <- alltypes(loc_list2017.ppp, Gdot)
Gdot18 <- alltypes(loc_list2018.ppp, Gdot)
Gdot19 <- alltypes(loc_list2019.ppp, Gdot)
Gdot20 <- alltypes(loc_list2020.ppp, Gdot)
Gdot21 <- alltypes(loc_list2021.ppp, Gdot)
Gdot22 <- alltypes(loc_list2022.ppp, Gdot)

```


# plot Gdot

```{r}
plot(Gdot16)
plot(Gdot17)
plot(Gdot18)
plot(Gdot19)
plot(Gdot20)
plot(Gdot21)
plot(Gdot22)
plot(Gdotall)
```


# Ripley's K cross function

calculate K cross

```{r}
#| cache = TRUE
K16 <- Kcross(loc_list2016.ppp, "Leptospirosis", "Enteric_fever") 
K17 <- Kcross(loc_list2017.ppp, "Leptospirosis", "Enteric_fever") 
K18 <- Kcross(loc_list2018.ppp, "Leptospirosis", "Enteric_fever") 
K19 <- Kcross(loc_list2019.ppp, "Leptospirosis", "Enteric_fever") 
K20 <- Kcross(loc_list2020.ppp, "Leptospirosis", "Enteric_fever") 
K21 <- Kcross(loc_list2021.ppp, "Leptospirosis", "Enteric_fever") 
K22 <- Kcross(loc_list2022.ppp, "Leptospirosis", "Enteric_fever") 
Kall <- Kcross(all_diag.ppp, "Leptospirosis", "Enteric_fever") 

```


plot Ripley's cross K function
- Departures from the reference line indicate clustering of the two types of points relative to each other
- When the observed K value is larger than the expected K value for a particular distance, the distribution is more clustered than a random distribution at that distance (scale of analysis). When the observed K value is smaller than the expected K value, the distribution is more dispersed than a random distribution at that distance

```{r name-of-chunk34, fig.width=20, fig.height=9}
par(mfrow=c(2,4))
plot(K16, main = 2016, cex.main = 3, cex.lab = 1.5)
plot(K17, main = 2017, cex.main = 3, cex.lab = 1.5)
plot(K18, main = 2018, cex.main = 3, cex.lab = 1.5)
plot(K19, main = 2019, cex.main = 3, cex.lab = 1.5)
plot(K20, main = 2020, cex.main = 3, cex.lab = 1.5)
plot(K21, main = 2021, cex.main = 3, cex.lab = 1.5)
plot(K22, main = 2022, cex.main = 3, cex.lab = 1.5)
plot(Kall, main = 2016-2022, cex.main = 3, cex.lab = 1.5)
```

# Kdot

Calculate Kdot

```{r}
#| cache = TRUE
Kdot16 <- Kdot(loc_list2016.ppp, "Leptospirosis")
Kdot17 <- Kdot(loc_list2017.ppp, "Leptospirosis")
Kdot18 <- Kdot(loc_list2018.ppp, "Leptospirosis")
Kdot19 <- Kdot(loc_list2019.ppp, "Leptospirosis")
Kdot20 <- Kdot(loc_list2020.ppp, "Leptospirosis")
Kdot21 <- Kdot(loc_list2021.ppp, "Leptospirosis")
Kdot22 <- Kdot(loc_list2022.ppp, "Leptospirosis")
Kdotall <- Kdot(all_diag.ppp, "Leptospirosis")
```


Plot Kdot
- Departures from the reference line indicate clustering or repulsion of the leptospirosis points relative to others

```{r name-of-chunk35, fig.width=20, fig.height=9}
par(mfrow=c(2,4))
plot(Kdot16, main = "2016", cex.main = 3, cex.lab = 1.5)
plot(Kdot17, main = "2017", cex.main = 3, cex.lab = 1.5)
plot(Kdot18, main = "2018", cex.main = 3, cex.lab = 1.5)
plot(Kdot19, main = "2019", cex.main = 3, cex.lab = 1.5)
plot(Kdot20, main = "2020", cex.main = 3, cex.lab = 1.5)
plot(Kdot21, main = "2021", cex.main = 3, cex.lab = 1.5)
plot(Kdot22, main = "2022", cex.main = 3, cex.lab = 1.5)
plot(Kdotall, main = "2016-2022", cex.main = 3, cex.lab = 1.5)
```


# Pair correlation function

pcf for K cross

```{r}
#| cache = TRUE
K16pcf<- pcf(K16, spar=1, method="b")
K17pcf<- pcf(K17, spar=1, method="b")
K18pcf<- pcf(K18, spar=1, method="b")
K19pcf<- pcf(K19, spar=1, method="b")
K20pcf<- pcf(K20, spar=1, method="b")
K21pcf<- pcf(K21, spar=1, method="b")
K22pcf<- pcf(K22, spar=1, method="b")
Kallpcf<- pcf(Kall, spar=1, method="b")
```



Plot pcf for K cross

```{r name-of-chunk36, fig.width=20, fig.height=9}
par(mfrow=c(2,4))
plot(K16pcf, main = "2016", cex.main = 3, cex.lab = 1.5)
plot(K17pcf, main = "2017", cex.main = 3, cex.lab = 1.5)
plot(K18pcf, main = "2018", cex.main = 3, cex.lab = 1.5)
plot(K19pcf, main = "2019", cex.main = 3, cex.lab = 1.5)
plot(K20pcf, main = "2020", cex.main = 3, cex.lab = 1.5)
plot(K21pcf, main = "2021", cex.main = 3, cex.lab = 1.5)
plot(K22pcf, main = "2022", cex.main = 3, cex.lab = 1.5)
plot(Kallpcf, main = "2016-2022", cex.main = 3, cex.lab = 1.5)
```

```{r}
#| cache = TRUE
# pcf cross
cross16pcf <- pcfcross(loc_list2016.ppp, "Leptospirosis", "Enteric_fever", stoyan=0.1)
cross17pcf <- pcfcross(loc_list2017.ppp, "Leptospirosis", "Enteric_fever", stoyan=0.1)
cross18pcf <- pcfcross(loc_list2018.ppp, "Leptospirosis", "Enteric_fever", stoyan=0.1)
cross19pcf <- pcfcross(loc_list2019.ppp, "Leptospirosis", "Enteric_fever", stoyan=0.1)
cross20pcf <- pcfcross(loc_list2020.ppp, "Leptospirosis", "Enteric_fever", stoyan=0.1)
cross21pcf <- pcfcross(loc_list2021.ppp, "Leptospirosis", "Enteric_fever", stoyan=0.1)
cross22pcf <- pcfcross(loc_list2022.ppp, "Leptospirosis", "Enteric_fever", stoyan=0.1)
crosspcf <- pcfcross(all_diag.ppp, "Leptospirosis", "Enteric_fever", stoyan=0.1)
```

```{r}
crosspcf <- pcfcross(all_diag.ppp, "Leptospirosis", "Enteric_fever", stoyan=0.1)
```


plot cross pcf
- A value g(r) < 1 indicates that interpoint distances equal to r are less frequent than would be expected for a completely random process, so this suggests regularity. A value g(r) > 1 indicates that this interpoint distance is more frequent than expected for a completely random pattern, which suggests clustering.

- If the value of gr,d obs(r) is larger than the value of the gr,d high(r), then leptospirosis and enteric fever suggest a positive relationship. If the value of gr,d obs(r) is smaller than the value of the gr,d low(r), then the leptospirosis and enteric fever suggest a negative relationship. In the grey zone, we cannot reject the null hypothesis of independency of components. The variable “r” refers to distance from leptospirosis to enteric fever.

```{r name-of-chunk37, fig.width=20, fig.height=9}
#| cache = TRUE
par(mfrow=c(2,4))
plot(cross16pcf, main = "2016", cex.main = 3, cex.lab = 1.5)
plot(cross17pcf, main = "2017", cex.main = 3, cex.lab = 1.5)
plot(cross18pcf, main = "2018", cex.main = 3, cex.lab = 1.5)
plot(cross19pcf, main = "2019", cex.main = 3, cex.lab = 1.5)
plot(cross20pcf, main = "2020", cex.main = 3, cex.lab = 1.5)
plot(cross21pcf, main = "2021", cex.main = 3, cex.lab = 1.5)
plot(cross22pcf, main = "2022", cex.main = 3, cex.lab = 1.5)
plot(crosspcf, main = "2016-2022", cex.main = 3, cex.lab = 1.5)
```


enveloped ripley cross K
- When the observed K value is larger than the upper confidence envelope (HiConfEnv) value, spatial clustering for that distance is statistically significant. When the observed K value is smaller than the lower confidence envelope (LwConfEnv) value, spatial dispersion for that distance is statistically significant.

```{r}
#| cache = TRUE
env_K01 <- plot(envelope(all_diag.ppp, Kcross, nsim = 99))
```

```{r}
#| cache = TRUE
env_K16 <- envelope(loc_list2016.ppp, Kcross, nsim = 99)
```

```{r}
#| cache = TRUE
env_K17 <- envelope(loc_list2017.ppp, Kcross, nsim = 99)
env_K18 <- envelope(loc_list2018.ppp, Kcross, nsim = 99)
env_K19 <- envelope(loc_list2019.ppp, Kcross, nsim = 99)
env_K20 <- envelope(loc_list2020.ppp, Kcross, nsim = 99)
env_K21 <- envelope(loc_list2021.ppp, Kcross, nsim = 99)
env_K22 <- envelope(loc_list2022.ppp, Kcross, nsim = 99)

```


Plot enveloped Ripley's (cross) K by year

```{r name-of-chunk38, fig.width=20, fig.height=9}
par(mfrow=c(2,4))
plot(env_K16, main = "2016", cex.main = 3)
plot(env_K17, main = "2017", cex.main = 3)
plot(env_K18, main = "2018", cex.main = 3)
plot(env_K19, main = "2019", cex.main = 3)
plot(env_K20, main = "2020", cex.main = 3)
plot(env_K21, main = "2021", cex.main = 3)
plot(env_K22, main = "2022", cex.main = 3)
```



