---
title: "Point Pattern Analysis - density & intensity of Enteric Fever and Leptospirosis in Kelantan"
author: "Hazlienor"
date: "2023-04-12"
output: 
  html_document: 
    keep_md: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load library

```{r}
library(sf)
library(tidyverse)
library(here)
library(janitor)
library(gtsummary)
library(DT)
library(stringr)
library(readxl)
library(broom)
library(tmap)
library(mapview)
library(lubridate)
library(maptools)
library(spatstat)
library(spdep)
library(gridExtra)
library(grid)
```


## Prepare data

read polygon data

```{r}
kel <- st_read(here("Map",
                    "kelantan.shp"))
```


read population data per mukim per year

```{r}
kel_mukim <- read_xlsx(here ("mukim.xlsx"))
```


merge population data to polygon

```{r}
kel_map <- merge(kel,kel_mukim,by.x="MUKIM", by.y="MUKIM", all.x=T, sort=F)
dim(kel_map)
class(kel_map)
st_crs(kel_map)
```

load disease data in xlxs

```{r}
linelist <- read_xlsx(here ("linelist.xlsx")) %>% clean_names()
glimpse(linelist)
```


remove na coordinate

```{r}
#disease data all - list all is all leptospirosis and enteric fever cases
listALL <- linelist %>% 
  filter(!is.na(latitude_wgs),
         !is.na(longitude_wgs))
```


convert all disease data to spatial data

```{r}
loc_ALL <- st_as_sf(listALL, 
                    coords = c("longitude_wgs", "latitude_wgs"), 
                    crs = 4326)
```


convert CRS to RSO

```{r}
loc_ALL2 <- st_transform(loc_ALL, 3168)
```



## Point Pattern Analysis

# Transform data

convert data to spatial format

```{r}
kel_map.sp <- as(kel_map, "Spatial")
class(kel_map.sp)
```

```{r}
plot(kel_map.sp)
```

convert point to spatial

```{r}
loc_ALL2.sp <- as(loc_ALL2, "Spatial")
class(loc_ALL2.sp)
```

select point only in Kelantan

```{r}
all_kel <- loc_ALL2 %>% 
  mutate(within_kel_map = lengths(st_within(loc_ALL2, kel_map)))
all_kel2 <- all_kel %>% 
  filter(within_kel_map == 1)
```


split data by diagnosis

```{r}
Enteric_kel <- all_kel2 %>% 
  filter(diagnosis == "Enteric_fever") 
lepto_kel <- all_kel2 %>% 
  filter(diagnosis == "Leptospirosis") 
```



## point pattern analysis for enteric fever

extract enteric fever by year

```{r}
loc_ent16 <- subset(Enteric_kel, tahun_daftar=="2016")
loc_ent17 <- subset(Enteric_kel, tahun_daftar=="2017")
loc_ent18 <- subset(Enteric_kel, tahun_daftar=="2018")
loc_ent19 <- subset(Enteric_kel, tahun_daftar=="2019")
loc_ent20 <- subset(Enteric_kel, tahun_daftar=="2020")
loc_ent21 <- subset(Enteric_kel, tahun_daftar=="2021")
loc_ent22 <- subset(Enteric_kel, tahun_daftar=="2022")
```


convert point to spatial

```{r}
loc_ent16.sp <- as(loc_ent16, "Spatial")
loc_ent17.sp <- as(loc_ent17, "Spatial")
loc_ent18.sp <- as(loc_ent18, "Spatial")
loc_ent19.sp <- as(loc_ent19, "Spatial")
loc_ent20.sp <- as(loc_ent20, "Spatial")
loc_ent21.sp <- as(loc_ent21, "Spatial")
loc_ent22.sp <- as(loc_ent22, "Spatial")
```


convert point spatial data to ppp format

```{r}
loc_ALL2.ppp <- as(loc_ALL2.sp, 'ppp')
class(loc_ALL2.ppp)
loc_ent16.ppp <- as(loc_ent16.sp, 'ppp')
loc_ent17.ppp <- as(loc_ent17.sp, 'ppp')
loc_ent18.ppp <- as(loc_ent18.sp, 'ppp')
loc_ent19.ppp <- as(loc_ent19.sp, 'ppp')
loc_ent20.ppp <- as(loc_ent20.sp, 'ppp')
loc_ent21.ppp <- as(loc_ent21.sp, 'ppp')
loc_ent22.ppp <- as(loc_ent22.sp, 'ppp')
```

transform to owin format

```{r}
kel_map.owin <- as(kel_map.sp, "owin")
class(kel_map.owin)
```


plot ppp object with window 

```{r}
# plot ALL
plot(kel_map.owin)
points(loc_ALL2.ppp)
```

convert enteric fever and lepto pint data to spatial to ppp

```{r}
Enteric_kel.sp <- as(Enteric_kel, "Spatial")
Enteric_kel.ppp <- as(Enteric_kel.sp, 'ppp')
lepto_kel.sp <- as(lepto_kel, "Spatial")
lepto_kel.ppp <- as(lepto_kel.sp, 'ppp')
```


plot enteric fever with window

```{r}
# plot Enteric Fever 2016-2022
plot(kel_map.owin)
points(Enteric_kel.ppp)
```

Remove marks

```{r}
# remove marks enteric fever 2016-2022
Enteric_kel.ppp2 <- Enteric_kel.ppp
marks(Enteric_kel.ppp2) <- NULL

# remove marks (year)
loc_ent16.ppp2 <- loc_ent16.ppp
marks(loc_ent16.ppp2) <- NULL
loc_ent17.ppp2 <- loc_ent17.ppp
marks(loc_ent17.ppp2) <- NULL
loc_ent18.ppp2 <- loc_ent18.ppp
marks(loc_ent18.ppp2) <- NULL
loc_ent19.ppp2 <- loc_ent19.ppp
marks(loc_ent19.ppp2) <- NULL
loc_ent20.ppp2 <- loc_ent20.ppp
marks(loc_ent20.ppp2) <- NULL
loc_ent21.ppp2 <- loc_ent21.ppp
marks(loc_ent21.ppp2) <- NULL
loc_ent22.ppp2 <- loc_ent22.ppp
marks(loc_ent22.ppp2) <- NULL
```

```{r}
# generate window
Window(Enteric_kel.ppp2) <- kel_map.owin

# generate window (year)
Window(loc_ent16.ppp2) <- kel_map.owin
Window(loc_ent17.ppp2) <- kel_map.owin
Window(loc_ent18.ppp2) <- kel_map.owin
Window(loc_ent19.ppp2) <- kel_map.owin
Window(loc_ent20.ppp2) <- kel_map.owin
Window(loc_ent21.ppp2) <- kel_map.owin
Window(loc_ent22.ppp2) <- kel_map.owin
```


Plot enteric fever by year

```{r name-of-chunk9, fig.width=20, fig.height=11}
par(mfrow=c(2,4))
plot(loc_ent16.ppp2, main = 2016, cols=rgb(0,0,0,.2), pch=20)
plot(loc_ent17.ppp2, main = 2017, cols=rgb(0,0,0,.2), pch=20)
plot(loc_ent18.ppp2, main = 2018, cols=rgb(0,0,0,.2), pch=20)
plot(loc_ent19.ppp2, main = 2019, cols=rgb(0,0,0,.2), pch=20)
plot(loc_ent20.ppp2, main = 2020, cols=rgb(0,0,0,.2), pch=20)
plot(loc_ent21.ppp2, main = 2021, cols=rgb(0,0,0,.2), pch=20)
plot(loc_ent22.ppp2, main = 2022, cols=rgb(0,0,0,.2), pch=20)
plot(Enteric_kel.ppp2, main = "2016-2022", cols=rgb(0,0,0,.2), pch=20)
```

# density analysis for enteric fever

quadrat count + plot

```{r name-of-chunk10, fig.width=28, fig.height=16}
par( mfrow= c(2,4) )
# 2016
quadr_count_ent16 <- quadratcount(loc_ent16.ppp2, 
                                  nx = 10,
                                  ny = 14)
plot(loc_ent16.ppp2, pch = 20, cex = 0.1, main = "2016")
plot(quadr_count_ent16, add = TRUE, cex = 2)

# 2017
quadr_count_ent17 <- quadratcount(loc_ent17.ppp2, 
                                  nx = 10,
                                  ny = 14)
plot(loc_ent17.ppp2, pch = 20, cex = 0.1, main = "2017")
plot(quadr_count_ent17, add = TRUE, cex = 2)

# 2018
quadr_count_ent18 <- quadratcount(loc_ent18.ppp2, 
                                  nx = 10,
                                  ny = 14)
plot(loc_ent18.ppp2, pch = 20, cex = 0.1, main = "2018")
plot(quadr_count_ent18, add = TRUE, cex = 2)

# 2019
quadr_count_ent19 <- quadratcount(loc_ent19.ppp2, 
                                  nx = 10,
                                  ny = 14)
plot(loc_ent19.ppp2, pch = 20, cex = 0.1, main = "2019")
plot(quadr_count_ent19, add = TRUE, cex = 2)

# 2020
quadr_count_ent20 <- quadratcount(loc_ent20.ppp2, 
                                  nx = 10,
                                  ny = 14)
plot(loc_ent20.ppp2, pch = 20, cex = 0.1, main = "2020")
plot(quadr_count_ent20, add = TRUE, cex = 2)

# 2021
quadr_count_ent21 <- quadratcount(loc_ent21.ppp2, 
                                  nx = 10,
                                  ny = 14)
plot(loc_ent21.ppp2, pch = 20, cex = 0.1, main = "2021")
plot(quadr_count_ent21, add = TRUE, cex = 2)

# 2022
quadr_count_ent22 <- quadratcount(loc_ent22.ppp2, 
                                  nx = 10,
                                  ny = 14)
plot(loc_ent22.ppp2, pch = 20, cex = 0.1, main = "2022")
plot(quadr_count_ent22, add = TRUE, cex = 2)

# cumulative cases 2016-2022
quadr_count_ent <- quadratcount(Enteric_kel.ppp2, 
                                nx = 10,
                                ny = 14)
plot(Enteric_kel.ppp2, pch = 20, cex = 0.1, main = "2016-2022")
plot(quadr_count_ent, add = TRUE, cex = 2)

mtext("Point Density Maps of Enteric Fever Cases in Kelantan 2016-2022", side = 1, line = -1, cex = 2, outer = TRUE)
```


calculate VMR (Variance-to-Mean Ratio) enteric fever

- VMR greater than 1 indicates spatial clustering.
- Enteric Fever not uniformly distributed and may be clustered in certain areas.

```{r}
#2016-2022
Qcountent<-data.frame(quadr_count_ent)
var(Qcountent$Freq)/mean(Qcountent$Freq)

#2016
Qcountent16<-data.frame(quadr_count_ent16)
var(Qcountent16$Freq)/mean(Qcountent16$Freq)

#2017
Qcountent17<-data.frame(quadr_count_ent17)
var(Qcountent17$Freq)/mean(Qcountent17$Freq)

#2018
Qcountent18<-data.frame(quadr_count_ent18)
var(Qcountent18$Freq)/mean(Qcountent18$Freq)

#2019
Qcountent19<-data.frame(quadr_count_ent19)
var(Qcountent19$Freq)/mean(Qcountent19$Freq)

#2020
Qcountent20<-data.frame(quadr_count_ent20)
var(Qcountent20$Freq)/mean(Qcountent20$Freq)

#2021
Qcountent21<-data.frame(quadr_count_ent21)
var(Qcountent21$Freq)/mean(Qcountent21$Freq)

#2022
Qcountent22<-data.frame(quadr_count_ent22)
var(Qcountent22$Freq)/mean(Qcountent22$Freq)
```
calculate Chi-square test for spatial randomnes - enteric fever

- a high VMR score and a significant chi-square test result would indicate that the points are clustered and non-randomly distributed, suggesting the presence of spatial processes such as spatial contagion, spatial dependence, or spatial interaction.

```{r}
#| cache = TRUE
chi_ent <- quadrat.test(Enteric_kel.ppp2, nx= 30, ny=15)
chi_ent16 <- quadrat.test(loc_ent16.ppp2, nx= 30, ny=15)
chi_ent17 <- quadrat.test(loc_ent17.ppp2, nx= 30, ny=15)
chi_ent18 <- quadrat.test(loc_ent18.ppp2, nx= 30, ny=15)
chi_ent19 <- quadrat.test(loc_ent19.ppp2, nx= 30, ny=15)
chi_ent20 <- quadrat.test(loc_ent20.ppp2, nx= 30, ny=15)
chi_ent21 <- quadrat.test(loc_ent21.ppp2, nx= 30, ny=15)
chi_ent22 <- quadrat.test(loc_ent22.ppp2, nx= 30, ny=15)
```
result chi square for enteric fever

```{r}
chi_ent.df <- data.frame(Dataset = "Enteric_kel.ppp2",
                       TestStatistic = chi_ent$statistic,
                       PValue = chi_ent$p.value)
chi_ent16.df <- data.frame(Dataset = "loc_ent16.ppp2",
                       TestStatistic = chi_ent16$statistic,
                       PValue = chi_ent16$p.value)
chi_ent17.df <- data.frame(Dataset = "loc_ent17.ppp2",
                       TestStatistic = chi_ent17$statistic,
                       PValue = chi_ent17$p.value)
chi_ent18.df <- data.frame(Dataset = "loc_ent18.ppp2",
                       TestStatistic = chi_ent18$statistic,
                       PValue = chi_ent18$p.value)
chi_ent19.df <- data.frame(Dataset = "loc_ent19.ppp2",
                       TestStatistic = chi_ent19$statistic,
                       PValue = chi_ent19$p.value)
chi_ent20.df <- data.frame(Dataset = "loc_ent20.ppp2",
                       TestStatistic = chi_ent20$statistic,
                       PValue = chi_ent20$p.value)
chi_ent21.df <- data.frame(Dataset = "loc_ent21.ppp2",
                       TestStatistic = chi_ent21$statistic,
                       PValue = chi_ent21$p.value)
chi_ent22.df <- data.frame(Dataset = "loc_ent22.ppp2",
                       TestStatistic = chi_ent22$statistic,
                       PValue = chi_ent22$p.value)
chi_quadENT <- bind_rows(
  data.frame(Dataset = "2016", chi_ent16.df), 
  data.frame(Dataset = "2017", chi_ent17.df), 
  data.frame(Dataset = "2018", chi_ent18.df), 
  data.frame(Dataset = "2019", chi_ent19.df), 
  data.frame(Dataset = "2020", chi_ent20.df), 
  data.frame(Dataset = "2021", chi_ent21.df), 
  data.frame(Dataset = "2022", chi_ent22.df), 
  data.frame(Dataset = "2016-2022", chi_ent.df))
chi_quadENT

```


Monte Carlo permutation for spatial randomnes - enteric fever (chi square assumption not met)

```{r}
#| cache = TRUE
# calculate monte carlo
mc_qent <- quadrat.test(Enteric_kel.ppp2, nx= 30, ny=15, method = "MonteCarlo")
mc_qent16 <- quadrat.test(loc_ent16.ppp2, nx= 30, ny=15, method = "MonteCarlo")
mc_qent17 <- quadrat.test(loc_ent17.ppp2, nx= 30, ny=15, method = "MonteCarlo")
mc_qent18 <- quadrat.test(loc_ent18.ppp2, nx= 30, ny=15, method = "MonteCarlo")
mc_qent19 <- quadrat.test(loc_ent19.ppp2, nx= 30, ny=15, method = "MonteCarlo")
mc_qent20 <- quadrat.test(loc_ent20.ppp2, nx= 30, ny=15, method = "MonteCarlo")
mc_qent21 <- quadrat.test(loc_ent21.ppp2, nx= 30, ny=15, method = "MonteCarlo")
mc_qent22 <- quadrat.test(loc_ent22.ppp2, nx= 30, ny=15, method = "MonteCarlo")

# convert to data frame
mc_qent.df <- data.frame(Dataset = "Enteric_kel.ppp2",
                       TestStatistic = mc_qent$statistic,
                       PValue = mc_qent$p.value)
mc_qent16.df <- data.frame(Dataset = "loc_ent16.ppp2",
                       TestStatistic = mc_qent16$statistic,
                       PValue = mc_qent16$p.value)
mc_qent17.df <- data.frame(Dataset = "loc_ent17.ppp2",
                       TestStatistic = mc_qent17$statistic,
                       PValue = mc_qent17$p.value)
mc_qent18.df <- data.frame(Dataset = "loc_ent18.ppp2",
                       TestStatistic = mc_qent18$statistic,
                       PValue = mc_qent18$p.value)
mc_qent19.df <- data.frame(Dataset = "loc_ent19.ppp2",
                       TestStatistic = mc_qent19$statistic,
                       PValue = mc_qent19$p.value)
mc_qent20.df <- data.frame(Dataset = "loc_ent20.ppp2",
                       TestStatistic = mc_qent20$statistic,
                       PValue = mc_qent20$p.value)
mc_qent21.df <- data.frame(Dataset = "loc_ent21.ppp2",
                       TestStatistic = mc_qent21$statistic,
                       PValue = mc_qent21$p.value)
mc_qent22.df <- data.frame(Dataset = "loc_ent22.ppp2",
                       TestStatistic = mc_qent22$statistic,
                       PValue = mc_qent22$p.value)

# combine rows
mc_qENT <- bind_rows(
  data.frame(Dataset = "2016", mc_qent16.df), 
  data.frame(Dataset = "2017", mc_qent17.df), 
  data.frame(Dataset = "2018", mc_qent18.df), 
  data.frame(Dataset = "2019", mc_qent19.df), 
  data.frame(Dataset = "2020", mc_qent20.df), 
  data.frame(Dataset = "2021", mc_qent21.df), 
  data.frame(Dataset = "2022", mc_qent22.df), 
  data.frame(Dataset = "2016-2022", mc_qent.df))
mc_qENT
```

# Intensity analysis for enteric fever

```{r name-of-chunk11, fig.width=16, fig.height=8}
par( mfrow= c(2,4) )

inten_ent16 <-intensity(quadr_count_ent16)
plot(intensity(quadr_count_ent16, image = TRUE), main = "2016", las = 1)
plot(loc_ent16, pch = 20, cex = 0.6, add = TRUE)

inten_ent17 <-intensity(quadr_count_ent17)
plot(intensity(quadr_count_ent17, image = TRUE), main = "2017", las = 1)
plot(loc_ent17, pch = 20, cex = 0.6, add = TRUE)

inten_ent18 <-intensity(quadr_count_ent18)
plot(intensity(quadr_count_ent18, image = TRUE), main = "2018", las = 1)
plot(loc_ent18, pch = 20, cex = 0.6, add = TRUE)

inten_ent19 <-intensity(quadr_count_ent19)
plot(intensity(quadr_count_ent19, image = TRUE), main = "2019", las = 1)
plot(loc_ent19, pch = 20, cex = 0.6, add = TRUE)

inten_ent20 <-intensity(quadr_count_ent20)
plot(intensity(quadr_count_ent20, image = TRUE), main = "2020", las = 1)
plot(loc_ent20, pch = 20, cex = 0.6, add = TRUE)

inten_ent21 <-intensity(quadr_count_ent21)
plot(intensity(quadr_count_ent21, image = TRUE), main = "2021", las = 1)
plot(loc_ent21, pch = 20, cex = 0.6, add = TRUE)

inten_ent22 <-intensity(quadr_count_ent22)
plot(intensity(quadr_count_ent22, image = TRUE), main = "2022", las = 1)
plot(loc_ent22, pch = 20, cex = 0.6, add = TRUE)

inten_ent <-intensity(quadr_count_ent)
plot(intensity(quadr_count_ent, image = TRUE), main = "2016-2022", las = 1)
plot(Enteric_kel.ppp2, pch = 20, cex = 0.6, add = TRUE)

mtext("Intensity Maps of Enteric Fever Cases in Kelantan 2016-2022", side = 1, line = -1, cex = 1.5, outer = TRUE)
```

# Kernel density raster for enteric fever

rescale to km

```{r}
# point
Enteric_kel.ppp2.km <- rescale(Enteric_kel.ppp2, 1000, 'km') #cumulative
loc_ent16.ppp2.km <- rescale(loc_ent16.ppp2, 1000, 'km') #2016
loc_ent17.ppp2.km <- rescale(loc_ent17.ppp2, 1000, 'km') #2017
loc_ent18.ppp2.km <- rescale(loc_ent18.ppp2, 1000, 'km') #2018
loc_ent19.ppp2.km <- rescale(loc_ent19.ppp2, 1000, 'km') #2019
loc_ent20.ppp2.km <- rescale(loc_ent20.ppp2, 1000, 'km') #2020
loc_ent21.ppp2.km <- rescale(loc_ent21.ppp2, 1000, 'km') #2021
loc_ent22.ppp2.km <- rescale(loc_ent22.ppp2, 1000, 'km') #2022

# map
kel_map.owin.km <- rescale(kel_map.owin, 1000, 'km') 
```


calculate kernel density estimate (kde) for Enteric Fever

```{r}
kde.ent <- density(Enteric_kel.ppp2.km) #2016 - 2022
kde.ent16 <- density(loc_ent16.ppp2.km) #2016
kde.ent17 <- density(loc_ent17.ppp2.km) #2017
kde.ent18 <- density(loc_ent18.ppp2.km) #2018
kde.ent19 <- density(loc_ent19.ppp2.km) #2019
kde.ent20 <- density(loc_ent20.ppp2.km) #2020
kde.ent21 <- density(loc_ent21.ppp2.km) #2021
kde.ent22 <- density(loc_ent22.ppp2.km) #2022
```


plot kde Enteric fever

```{r name-of-chunk12, fig.width=16, fig.height=8}
par(mfrow=c(2,4))
plot(kde.ent16, main = 2016, las = 1) #2016
plot(kde.ent17, main = 2017, las = 1) #2017
plot(kde.ent18, main = 2018, las = 1) #2018
plot(kde.ent19, main = 2019, las = 1) #2019
plot(kde.ent20, main = 2020, las = 1) #2020
plot(kde.ent21, main = 2021, las = 1) #2021
plot(kde.ent22, main = 2022, las = 1) #2022
plot(kde.ent, main = "2016-2022", las = 1) #2016-2022
```

kde 20km bandwith for enteric fever

```{r name-of-chunk13, fig.width=16, fig.height=8}
kde.ent.20km <- density(Enteric_kel.ppp2.km, sigma = 20) #2016-2022
kde.ent16.20km <- density(loc_ent16.ppp2.km, sigma = 20) #2016
kde.ent17.20km <- density(loc_ent17.ppp2.km, sigma = 20) #2017
kde.ent18.20km <- density(loc_ent18.ppp2.km, sigma = 20) #2018
kde.ent19.20km <- density(loc_ent19.ppp2.km, sigma = 20) #2019
kde.ent20.20km <- density(loc_ent20.ppp2.km, sigma = 20) #2020
kde.ent21.20km <- density(loc_ent21.ppp2.km, sigma = 20) #2021
kde.ent22.20km <- density(loc_ent22.ppp2.km, sigma = 20) #2022

# plot kde 20km

par(mfrow=c(2,4))
# 2016
plot(kde.ent16.20km, main = 2016, las = 1)
contour(kde.ent16.20km, add = TRUE)
# 2017
plot(kde.ent17.20km, main = 2017, las = 1)
contour(kde.ent17.20km, add = TRUE)
# 2018
plot(kde.ent18.20km, main = 2018, las = 1)
contour(kde.ent18.20km, add = TRUE)
# 2019
plot(kde.ent19.20km, main = 2019, las = 1)
contour(kde.ent19.20km, add = TRUE)
# 2020
plot(kde.ent20.20km, main = 2020, las = 1)
contour(kde.ent20.20km, add = TRUE)
# 2021
plot(kde.ent21.20km, main = 2021, las = 1)
contour(kde.ent21.20km, add = TRUE)
# 2022
plot(kde.ent22.20km, main = 2022, las = 1)
contour(kde.ent22.20km, add = TRUE)
# 2016-2022
plot(kde.ent.20km, main = "2016-2022", las = 1)
contour(kde.ent.20km, add = TRUE)
```


calculate kde 5km bandwith enteric fever

```{r}
#| cache = TRUE
kde.ent.5km <- density(Enteric_kel.ppp2.km, sigma = 5) #2016-2022
kde.ent16.5km <- density(loc_ent16.ppp2.km, sigma = 5) #2016
kde.ent17.5km <- density(loc_ent17.ppp2.km, sigma = 5) #2017
kde.ent18.5km <- density(loc_ent18.ppp2.km, sigma = 5) #2018
kde.ent19.5km <- density(loc_ent19.ppp2.km, sigma = 5) #2019
kde.ent20.5km <- density(loc_ent20.ppp2.km, sigma = 5) #2020
kde.ent21.5km <- density(loc_ent21.ppp2.km, sigma = 5) #2021
kde.ent22.5km <- density(loc_ent22.ppp2.km, sigma = 5) #2022
```


plot kde 5km bandwith enteric fever

```{r name-of-chunk14, fig.width=20, fig.height=11}
# plot all KDE enteric fever in one view
par( mfrow= c(2,4) )
plot(kde.ent16.5km, main = "2016", cex.main = 1.5, font.main = 2, las = 1)
contour(kde.ent16.5km, add = TRUE)
plot(kde.ent17.5km, main = "2017", cex.main = 1.5, font.main = 2, las = 1)
contour(kde.ent17.5km, add = TRUE)
plot(kde.ent18.5km, main = "2018", cex.main = 1.5, font.main = 2, las = 1)
contour(kde.ent18.5km, add = TRUE)
plot(kde.ent19.5km, main = "2019", cex.main = 1.5, font.main = 2, las = 1)
contour(kde.ent19.5km, add = TRUE)
plot(kde.ent20.5km, main = "2020", cex.main = 1.5, font.main = 2, las = 1)
contour(kde.ent20.5km, add = TRUE)
plot(kde.ent21.5km, main = "2021", cex.main = 1.5, font.main = 2, las = 1)
contour(kde.ent21.5km, add = TRUE)
plot(kde.ent22.5km, main = "2022", cex.main = 1.5, font.main = 2, las = 1)
contour(kde.ent22.5km, add = TRUE)
plot(kde.ent.5km, main = "2016-2022", cex.main = 1.5, font.main = 2, las = 1)
contour(kde.ent.5km, add = TRUE)

mtext("Kernel Density Estimate (KDE) Heatmaps of Enteric Fever Cases in Kelantan 2016-2022", side = 1, line = -1, cex = 2, outer = TRUE)
```


## point pattern analysis for Leptospirosis

prepare data for leptospirosis

```{r}
plot(kel_map.owin)
points(lepto_kel.ppp)
```


extract leptospirosis by year, convert to ppp

```{r}
loc_lep16 <- subset(lepto_kel, tahun_daftar=="2016")
loc_lep17 <- subset(lepto_kel, tahun_daftar=="2017")
loc_lep18 <- subset(lepto_kel, tahun_daftar=="2018")
loc_lep19 <- subset(lepto_kel, tahun_daftar=="2019")
loc_lep20 <- subset(lepto_kel, tahun_daftar=="2020")
loc_lep21 <- subset(lepto_kel, tahun_daftar=="2021")
loc_lep22 <- subset(lepto_kel, tahun_daftar=="2022")

```


convert point to spatial

```{r}
loc_lep16.sp <- as(loc_lep16, "Spatial")
```

```{r}
loc_lep17.sp <- as(loc_lep17, "Spatial")
```

```{r}
loc_lep18.sp <- as(loc_lep18, "Spatial")
```

```{r}
loc_lep19.sp <- as(loc_lep19, "Spatial")
```

```{r}
loc_lep20.sp <- as(loc_lep20, "Spatial")
```

```{r}
loc_lep21.sp <- as(loc_lep21, "Spatial")
```

```{r}
loc_lep22.sp <- as(loc_lep22, "Spatial")
```


convert point spatial data to ppp format

```{r}
loc_lep16.ppp <- as(loc_lep16.sp, 'ppp')
loc_lep17.ppp <- as(loc_lep17.sp, 'ppp')
loc_lep18.ppp <- as(loc_lep18.sp, 'ppp')
loc_lep19.ppp <- as(loc_lep19.sp, 'ppp')
loc_lep20.ppp <- as(loc_lep20.sp, 'ppp')
loc_lep21.ppp <- as(loc_lep21.sp, 'ppp')
loc_lep22.ppp <- as(loc_lep22.sp, 'ppp')
```


remove marks

```{r}
#remove marks leptospirosis 2016-2022
lepto_kel.ppp2 <- lepto_kel.ppp
marks(lepto_kel.ppp2) <- NULL

#remove marks (year)
loc_lep16.ppp2 <- loc_lep16.ppp
marks(loc_lep16.ppp2) <- NULL
loc_lep17.ppp2 <- loc_lep17.ppp
marks(loc_lep17.ppp2) <- NULL
loc_lep18.ppp2 <- loc_lep18.ppp
marks(loc_lep18.ppp2) <- NULL
loc_lep19.ppp2 <- loc_lep19.ppp
marks(loc_lep19.ppp2) <- NULL
loc_lep20.ppp2 <- loc_lep20.ppp
marks(loc_lep20.ppp2) <- NULL
loc_lep21.ppp2 <- loc_lep21.ppp
marks(loc_lep21.ppp2) <- NULL
loc_lep22.ppp2 <- loc_lep22.ppp
marks(loc_lep22.ppp2) <- NULL
```


generate window

```{r}
Window(lepto_kel.ppp2) <- kel_map.owin
Window(loc_lep16.ppp2) <- kel_map.owin
Window(loc_lep17.ppp2) <- kel_map.owin
Window(loc_lep18.ppp2) <- kel_map.owin
Window(loc_lep19.ppp2) <- kel_map.owin
Window(loc_lep20.ppp2) <- kel_map.owin
Window(loc_lep21.ppp2) <- kel_map.owin
Window(loc_lep22.ppp2) <- kel_map.owin
```


plot leptospirosis cases

```{r name-of-chunk15, fig.width=20, fig.height=11}
par(mfrow=c(2,4))
plot(loc_lep16.ppp2, main = 2016, cols=rgb(0,0,0,.2), pch=20) #2016
plot(loc_lep17.ppp2, main = 2017, cols=rgb(0,0,0,.2), pch=20) #2017
plot(loc_lep18.ppp2, main = 2018, cols=rgb(0,0,0,.2), pch=20) #2018
plot(loc_lep19.ppp2, main = 2019, cols=rgb(0,0,0,.2), pch=20) #2019
plot(loc_lep20.ppp2, main = 2020, cols=rgb(0,0,0,.2), pch=20) #2020
plot(loc_lep21.ppp2, main = 2021, cols=rgb(0,0,0,.2), pch=20) #2021
plot(loc_lep22.ppp2, main = 2022, cols=rgb(0,0,0,.2), pch=20) #2022
plot(lepto_kel.ppp2, main = "2016-2022", cols=rgb(0,0,0,.2), pch=20) #2016-2022
```


# density analysis for leptospirosis

calculate quadrat count 

```{r}
#| cache = TRUE
quadr_count_lep <- quadratcount(lepto_kel.ppp2, nx = 10,ny = 14) #2016-2022
quadr_count_lep16 <- quadratcount(loc_lep16.ppp2, nx = 10,ny = 14) #2016
quadr_count_lep17 <- quadratcount(loc_lep17.ppp2, nx = 10,ny = 14) #2017
quadr_count_lep18 <- quadratcount(loc_lep18.ppp2, nx = 10,ny = 14) #2018
quadr_count_lep19 <- quadratcount(loc_lep19.ppp2, nx = 10,ny = 14) #2019
quadr_count_lep20 <- quadratcount(loc_lep20.ppp2, nx = 10,ny = 14) #2020
quadr_count_lep21 <- quadratcount(loc_lep21.ppp2, nx = 10,ny = 14) #2021
quadr_count_lep22 <- quadratcount(loc_lep22.ppp2, nx = 10,ny = 14) #2022
```


plot quadrat count

```{r name-of-chunk16, fig.width=28, fig.height=16}
par( mfrow= c(2,4) )
#2016
plot(loc_lep16.ppp2, pch = 20, cex = 0.1, main = "2016")
plot(quadr_count_lep16, add = TRUE, cex = 2)
#2017
plot(loc_lep17.ppp2, pch = 20, cex = 0.1, main = "2017")
plot(quadr_count_lep17, add = TRUE, cex = 2)
#2018
plot(loc_lep18.ppp2, pch = 20, cex = 0.1, main = "2018")
plot(quadr_count_lep18, add = TRUE, cex = 2)
#2019
plot(loc_lep19.ppp2, pch = 20, cex = 0.1, main = "2019")
plot(quadr_count_lep19, add = TRUE, cex = 2)
#2020
plot(loc_lep20.ppp2, pch = 20, cex = 0.1, main = "2020")
plot(quadr_count_lep20, add = TRUE, cex = 2)
#2021
plot(loc_lep21.ppp2, pch = 20, cex = 0.1, main = "2021")
plot(quadr_count_lep21, add = TRUE, cex = 2)
#2022
plot(loc_lep22.ppp2, pch = 20, cex = 0.1, main = "2022")
plot(quadr_count_lep22, add = TRUE, cex = 2)
#2016-2022
plot(lepto_kel.ppp2, pch = 20, cex = 0.1, main = "2016-2022")
plot(quadr_count_lep, add = TRUE, cex = 2) 

mtext("Point Density Maps of Leptospirosis Cases in Kelantan 2016-2022", side = 1, line = -1, cex = 2, outer = TRUE)
```


calculate VMR (Variance-to-Mean Ratio) Leptospirosis
- VMR greater than 1 indicates spatial clustering.
- leptospirosis not uniformly distributed and may be clustered in certain areas.

```{r}
#| cache = TRUE
#2016-2022
Qcountlep<-data.frame(quadr_count_lep)
var(Qcountlep$Freq)/mean(Qcountlep$Freq)

#2016
Qcountlep16<-data.frame(quadr_count_lep16)
var(Qcountlep16$Freq)/mean(Qcountlep16$Freq)

#2017
Qcountlep17<-data.frame(quadr_count_lep17)
var(Qcountlep17$Freq)/mean(Qcountlep17$Freq)

#2018
Qcountlep18<-data.frame(quadr_count_lep18)
var(Qcountlep18$Freq)/mean(Qcountlep18$Freq)

#2019
Qcountlep19<-data.frame(quadr_count_lep19)
var(Qcountlep19$Freq)/mean(Qcountlep19$Freq)

#2020
Qcountlep20<-data.frame(quadr_count_lep20)
var(Qcountlep20$Freq)/mean(Qcountlep20$Freq)

#2021
Qcountlep21<-data.frame(quadr_count_lep21)
var(Qcountlep21$Freq)/mean(Qcountlep21$Freq)

#2022
Qcountlep22<-data.frame(quadr_count_lep22)
var(Qcountlep22$Freq)/mean(Qcountlep22$Freq)
```

calculate Chi-square test for spatial randomnes - leptospirosis

- a high VMR score and a significant chi-square test result would indicate that the points are clustered and non-randomly distributed, suggesting the presence of spatial processes such as spatial contagion, spatial dependence, or spatial interaction.

```{r}
chi_lep <- quadrat.test(lepto_kel.ppp2, nx= 30, ny=15)
chi_lep16 <- quadrat.test(loc_lep16.ppp2, nx= 30, ny=15)
chi_lep17 <- quadrat.test(loc_lep17.ppp2, nx= 30, ny=15)
chi_lep18 <- quadrat.test(loc_lep18.ppp2, nx= 30, ny=15)
chi_lep19 <- quadrat.test(loc_lep19.ppp2, nx= 30, ny=15)
chi_lep20 <- quadrat.test(loc_lep20.ppp2, nx= 30, ny=15)
chi_lep21 <- quadrat.test(loc_lep21.ppp2, nx= 30, ny=15)
chi_lep22 <- quadrat.test(loc_lep22.ppp2, nx= 30, ny=15)
```

result chi square for enteric fever

```{r}
chi_lep.df <- data.frame(Dataset = "lepto_kel.ppp2",
                       TestStatistic = chi_lep$statistic,
                       PValue = chi_lep$p.value)
chi_lep16.df <- data.frame(Dataset = "loc_lep16.ppp2",
                       TestStatistic = chi_lep16$statistic,
                       PValue = chi_lep16$p.value)
chi_lep17.df <- data.frame(Dataset = "loc_lep17.ppp2",
                       TestStatistic = chi_lep17$statistic,
                       PValue = chi_lep17$p.value)
chi_lep18.df <- data.frame(Dataset = "loc_lep18.ppp2",
                       TestStatistic = chi_lep18$statistic,
                       PValue = chi_lep18$p.value)
chi_lep19.df <- data.frame(Dataset = "loc_lep19.ppp2",
                       TestStatistic = chi_lep19$statistic,
                       PValue = chi_lep19$p.value)
chi_lep20.df <- data.frame(Dataset = "loc_lep20.ppp2",
                       TestStatistic = chi_lep20$statistic,
                       PValue = chi_lep20$p.value)
chi_lep21.df <- data.frame(Dataset = "loc_lep21.ppp2",
                       TestStatistic = chi_lep21$statistic,
                       PValue = chi_lep21$p.value)
chi_lep22.df <- data.frame(Dataset = "loc_lep22.ppp2",
                       TestStatistic = chi_lep22$statistic,
                       PValue = chi_lep22$p.value)
chi_quadLEP <- bind_rows(
  data.frame(Dataset = "2016", chi_lep16.df), 
  data.frame(Dataset = "2017", chi_lep17.df), 
  data.frame(Dataset = "2018", chi_lep18.df), 
  data.frame(Dataset = "2019", chi_lep19.df), 
  data.frame(Dataset = "2020", chi_lep20.df), 
  data.frame(Dataset = "2021", chi_lep21.df), 
  data.frame(Dataset = "2022", chi_lep22.df), 
  data.frame(Dataset = "2016-2022", chi_lep.df))
chi_quadLEP

```


Monte Carlo permutation for spatial randomnes - leptospirosis (chi square assumption not met)

```{r}
# calculate monte carlo
#| cache = TRUE
mc_qlep <- quadrat.test(lepto_kel.ppp2, nx= 30, ny=15, method = "MonteCarlo")
mc_qlep16 <- quadrat.test(loc_lep16.ppp2, nx= 30, ny=15, method = "MonteCarlo")
mc_qlep17 <- quadrat.test(loc_lep17.ppp2, nx= 30, ny=15, method = "MonteCarlo")
mc_qlep18 <- quadrat.test(loc_lep18.ppp2, nx= 30, ny=15, method = "MonteCarlo")
mc_qlep19 <- quadrat.test(loc_lep19.ppp2, nx= 30, ny=15, method = "MonteCarlo")
mc_qlep20 <- quadrat.test(loc_lep20.ppp2, nx= 30, ny=15, method = "MonteCarlo")
mc_qlep21 <- quadrat.test(loc_lep21.ppp2, nx= 30, ny=15, method = "MonteCarlo")
mc_qlep22 <- quadrat.test(loc_lep22.ppp2, nx= 30, ny=15, method = "MonteCarlo")
```


convert to data frame

```{r}
mc_qlep.df <- data.frame(Dataset = "lepto_kel.ppp2",
                       TestStatistic = mc_qlep$statistic,
                       PValue = mc_qlep$p.value)
mc_qlep16.df <- data.frame(Dataset = "loc_lep16.ppp2",
                       TestStatistic = mc_qlep16$statistic,
                       PValue = mc_qlep16$p.value)
mc_qlep17.df <- data.frame(Dataset = "loc_lep17.ppp2",
                       TestStatistic = mc_qlep17$statistic,
                       PValue = mc_qlep17$p.value)
mc_qlep18.df <- data.frame(Dataset = "loc_lep18.ppp2",
                       TestStatistic = mc_qlep18$statistic,
                       PValue = mc_qlep18$p.value)
mc_qlep19.df <- data.frame(Dataset = "loc_lep19.ppp2",
                       TestStatistic = mc_qlep19$statistic,
                       PValue = mc_qlep19$p.value)
mc_qlep20.df <- data.frame(Dataset = "loc_lep20.ppp2",
                       TestStatistic = mc_qlep20$statistic,
                       PValue = mc_qlep20$p.value)
mc_qlep21.df <- data.frame(Dataset = "loc_lep21.ppp2",
                       TestStatistic = mc_qlep21$statistic,
                       PValue = mc_qlep21$p.value)
mc_qlep22.df <- data.frame(Dataset = "loc_lep22.ppp2",
                       TestStatistic = mc_qlep22$statistic,
                       PValue = mc_qlep22$p.value)
```


Display quadrat test with Monte Carlo permutation

```{r}
mc_qLEP <- bind_rows(
  data.frame(Dataset = "2016", mc_qlep16.df), 
  data.frame(Dataset = "2017", mc_qlep17.df), 
  data.frame(Dataset = "2018", mc_qlep18.df), 
  data.frame(Dataset = "2019", mc_qlep19.df), 
  data.frame(Dataset = "2020", mc_qlep20.df), 
  data.frame(Dataset = "2021", mc_qlep21.df), 
  data.frame(Dataset = "2022", mc_qlep22.df), 
  data.frame(Dataset = "2016-2022", mc_qlep.df))
mc_qLEP
```


# intensity analysis for leptospirosis

Plot intensity

```{r name-of-chunk17, fig.width=16, fig.height=8}
par( mfrow= c(2,4) )
inten_lep16 <-intensity(quadr_count_lep16)
plot(intensity(quadr_count_lep16, image = TRUE), main = 2016, las = 1)
plot(loc_lep16, pch = 20, cex = 0.6, add = TRUE)

inten_lep17 <-intensity(quadr_count_lep17)
plot(intensity(quadr_count_lep17, image = TRUE), main = 2017, las = 1)
plot(loc_lep17, pch = 20, cex = 0.6, add = TRUE)

inten_lep18 <-intensity(quadr_count_lep18)
plot(intensity(quadr_count_lep18, image = TRUE), main = 2018, las = 1)
plot(loc_lep18, pch = 20, cex = 0.6, add = TRUE)

inten_lep19 <-intensity(quadr_count_lep19)
plot(intensity(quadr_count_lep19, image = TRUE), main = 2019, las = 1)
plot(loc_lep19, pch = 20, cex = 0.6, add = TRUE)

inten_lep20 <-intensity(quadr_count_lep20)
plot(intensity(quadr_count_lep20, image = TRUE), main = 2020, las = 1)
plot(loc_lep20, pch = 20, cex = 0.6, add = TRUE)

inten_lep21 <-intensity(quadr_count_lep21)
plot(intensity(quadr_count_lep21, image = TRUE), main = 2021, las = 1)
plot(loc_lep21, pch = 20, cex = 0.6, add = TRUE)

inten_lep22 <-intensity(quadr_count_lep22)
plot(intensity(quadr_count_lep22, image = TRUE), main = 2022, las = 1)
plot(loc_lep22, pch = 20, cex = 0.6, add = TRUE)

inten_lep <-intensity(quadr_count_lep)
plot(intensity(quadr_count_lep, image = TRUE), main = "2016-2022" , las = 1)
plot(lepto_kel.ppp2, pch = 20, cex = 0.6, add = TRUE)

mtext("Intensity Maps of Leptospirosis Cases in Kelantan 2016-2022", side = 1, line = -1, cex = 1.5, outer = TRUE)
```

# Kernel density raster for leptospirosis

rescale to km

```{r}
#point
lepto_kel.ppp2.km <- rescale(lepto_kel.ppp2, 1000, 'km') #cumulative
loc_lep16.ppp2.km <- rescale(loc_lep16.ppp2, 1000, 'km') #2016
loc_lep17.ppp2.km <- rescale(loc_lep17.ppp2, 1000, 'km') #2017
loc_lep18.ppp2.km <- rescale(loc_lep18.ppp2, 1000, 'km') #2018
loc_lep19.ppp2.km <- rescale(loc_lep19.ppp2, 1000, 'km') #2019
loc_lep20.ppp2.km <- rescale(loc_lep20.ppp2, 1000, 'km') #2020
loc_lep21.ppp2.km <- rescale(loc_lep21.ppp2, 1000, 'km') #2021
loc_lep22.ppp2.km <- rescale(loc_lep22.ppp2, 1000, 'km') #2022

#map
kel_map.owin.km <- rescale(kel_map.owin, 1000, 'km') 
```


calculate kernel density raster (kde) for Leptospirosis 2016 - 2022

```{r}
#| cache = TRUE
kde.lep <- density(lepto_kel.ppp2.km) #2016 - 2022
kde.lep16 <- density(loc_lep16.ppp2.km) #2016
kde.lep17 <- density(loc_lep17.ppp2.km) #2017
kde.lep18 <- density(loc_lep18.ppp2.km) #2018
kde.lep19 <- density(loc_lep19.ppp2.km) #2019
kde.lep20 <- density(loc_lep20.ppp2.km) #2020
kde.lep21 <- density(loc_lep21.ppp2.km) #2021
kde.lep22 <- density(loc_lep22.ppp2.km) #2022
```


plot KDE leptospirosis

```{r name-of-chunk18, fig.width=20, fig.height=11}
par(mfrow=c(2,4))
plot(kde.lep16, main = 2016, las = 1) 
plot(kde.lep17, main = 2017, las = 1) 
plot(kde.lep18, main = 2018, las = 1) 
plot(kde.lep19, main = 2019, las = 1) 
plot(kde.lep20, main = 2020, las = 1) 
plot(kde.lep21, main = 2021, las = 1) 
plot(kde.lep22, main = 2022, las = 1) 
plot(kde.lep, main = "2016-2022", las = 1)
```

plot kde 20km bandwith leptospirosis

```{r name-of-chunk19, fig.width=20, fig.height=11}
kde.lep.20km <- density(lepto_kel.ppp2.km, sigma = 20) #2016 - 2022
kde.lep16.20km <- density(loc_lep16.ppp2.km, sigma = 20) #2016
kde.lep17.20km <- density(loc_lep17.ppp2.km, sigma = 20) #2017
kde.lep18.20km <- density(loc_lep18.ppp2.km, sigma = 20) #2018
kde.lep19.20km <- density(loc_lep19.ppp2.km, sigma = 20) #2019
kde.lep20.20km <- density(loc_lep20.ppp2.km, sigma = 20) #2020
kde.lep21.20km <- density(loc_lep21.ppp2.km, sigma = 20) #2021
kde.lep22.20km <- density(loc_lep22.ppp2.km, sigma = 20) #2022

#plot KDE leptospirosis 20km bandwith
par(mfrow=c(2,4))
#2016
plot(kde.lep16.20km, main = 2016, las = 1)
contour(kde.lep16.20km, add = TRUE)
#2017
plot(kde.lep17.20km, main = 2017, las = 1)
contour(kde.lep17.20km, add = TRUE)
#2018
plot(kde.lep18.20km, main = 2018, las = 1)
contour(kde.lep18.20km, add = TRUE)
#2019
plot(kde.lep19.20km, main = 2019, las = 1)
contour(kde.lep19.20km, add = TRUE)
#2020
plot(kde.lep20.20km, main = 2020, las = 1)
contour(kde.lep20.20km, add = TRUE)
#2021
plot(kde.lep21.20km, main = 2021, las = 1)
contour(kde.lep21.20km, add = TRUE)
#2022
plot(kde.lep22.20km, main = 2022, las = 1)
contour(kde.lep22.20km, add = TRUE)
#2016-2022
plot(kde.lep.20km, main = "2016-2022", las = 1)
contour(kde.lep.20km, add = TRUE)
```

kde 5km bandwith leptospirosis

```{r}
#| cache = TRUE
kde.lep.5km <- density(lepto_kel.ppp2.km, sigma = 5) #2016 - 2022
kde.lep16.5km <- density(loc_lep16.ppp2.km, sigma = 5) #2016
kde.lep17.5km <- density(loc_lep17.ppp2.km, sigma = 5) #2017
kde.lep18.5km <- density(loc_lep18.ppp2.km, sigma = 5) #2018
kde.lep19.5km <- density(loc_lep19.ppp2.km, sigma = 5) #2019
kde.lep20.5km <- density(loc_lep20.ppp2.km, sigma = 5) #2020
kde.lep21.5km <- density(loc_lep21.ppp2.km, sigma = 5) #2021
kde.lep22.5km <- density(loc_lep22.ppp2.km, sigma = 5) #2022
```


plot kde 5km bandwith leptospirosis

```{r name-of-chunk20, fig.width=20, fig.height=11}
par( mfrow= c(2,4) )
plot(kde.lep16.5km, main = "2016", cex.main = 1.5, font.main = 2, las = 1)
contour(kde.lep16.5km, add = TRUE)
plot(kde.lep17.5km, main = "2017", cex.main = 1.5, font.main = 2, las = 1)
contour(kde.lep17.5km, add = TRUE)
plot(kde.lep18.5km, main = "2018", cex.main = 1.5, font.main = 2, las = 1)
contour(kde.lep18.5km, add = TRUE)
plot(kde.lep19.5km, main = "2019", cex.main = 1.5, font.main = 2, las = 1)
contour(kde.lep19.5km, add = TRUE)
plot(kde.lep20.5km, main = "2020", cex.main = 1.5, font.main = 2, las = 1)
contour(kde.lep20.5km, add = TRUE)
plot(kde.lep21.5km, main = "2021", cex.main = 1.5, font.main = 2, las = 1)
contour(kde.lep21.5km, add = TRUE)
plot(kde.lep22.5km, main = "2022", cex.main = 1.5, font.main = 2, las = 1)
contour(kde.lep22.5km, add = TRUE)
plot(kde.lep.5km, main = "2016-2022", cex.main = 1.5, font.main = 2, las = 1)
contour(kde.lep.5km, add = TRUE)

#add title to the plot
mtext("Kernel Density Estimate (KDE) Heatmaps of Leptospirosis Cases in Kelantan 2016-2022", side = 1, line = -1, cex = 2, outer = TRUE)
```


