---
title: "SpatialMPH"
author: "Hazlienor"
date: "2023-03-24"
output: 
  html_document: 
    keep_md: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Prepare environment

Required packages 

```{r}
# install.packages("sf")
# install.packages("tidyverse")
# install.packages("here") #working directory
# install.packages("janitor")
# install.packages("gtsummary")
# install.packages("DT")
# install.packages("stringr")
# install.packages("readxl")
# install.packages("broom")
# install.packages("tmap")
# install.packages("mapview")
# install.packages("lubridate")
# install.packages("vctrs")
# install.packages("spatialECO") for NNI

```

Load packages 

```{r}
library(sf)
library(tidyverse)
library(here)
library(janitor)
library(gtsummary)
library(DT)
library(stringr)
library(readxl)
library(broom)
library(tmap)
library(mapview)
library(lubridate)
library(maptools)
library(spatstat)
library(spdep)
library(gridExtra)
library(sparr)
library(grid)
library(spatialEco)
```

## Load polygon data

read polygon data - kelantan map

```{r}
kel <- st_read(here("Map",
                    "kelantan.shp"))
```


```{r}
st_geometry(kel)
```

read population data per mukim per year

```{r}
kel_mukim <- read_xlsx(here ("mukim.xlsx"))
kel_mukim %>% datatable()
```

merge population data to polygon

```{r}
kel_map <- merge(kel,kel_mukim,by.x="MUKIM", by.y="MUKIM", all.x=T, sort=F)
dim(kel_map)
class(kel_map)
st_crs(kel_map)
```

Make plot

```{r}
st_geometry(kel_map)
plot(kel_map[,2]) # negeri
plot(kel_map[,3]) # daerah
plot(kel_map[,1]) # mukim
```


Adding facet 

```{r}
kel_map %>% ggplot() + geom_sf(aes(fill = DAERAH)) +ggtitle('Population by district') + theme_bw()
kel_map %>% ggplot() + geom_sf(aes(fill = MUKIM)) +ggtitle('Population by subdistrict') + theme_bw() + theme(legend.position = "none")
tm_shape(kel_map) + tm_polygons("AVR_FEMALE") + tm_layout(legend.width = 0.50)
tm_shape(kel_map) + tm_polygons("AVR")
tm_shape(kel_map) +
  tm_polygons(c("MUKIM", "DAERAH", "AVR_MALE")) +
  tm_facets(ncol = 3)
tm_shape(kel_map) +
  tm_polygons(c("MUKIM", "DAERAH", "AVR_MALE")) +
  tm_facets(nrow = 3)
```


## load disease data

read linelisting in .xlsx format

```{r}
linelist <- read_xlsx(here ("linelist.xlsx")) %>% clean_names()
glimpse(linelist)
```


remove na coordinate

```{r}
#disease data all - list all is all leptospirosis and enteric fever cases
listALL <- linelist %>% 
  filter(!is.na(latitude_wgs),
         !is.na(longitude_wgs))
glimpse(listALL)
```


## convert all disease data to spatial data

```{r}
loc_ALL <- st_as_sf(listALL, 
                    coords = c("longitude_wgs", "latitude_wgs"), 
                    crs = 4326)
loc_ALL %>% datatable()
```


confirm CRS is wgs84

```{r}
st_crs(loc_ALL)
```


plot map to see outlier

```{r}
ggplot() +
  geom_sf(data = loc_ALL) +
  ggtitle("Map of Leptospirosis and Enteric Fever") +
  theme_bw()
```


## convert shapefile to RSO

```{r}
loc_ALL2 <- st_transform(loc_ALL, 3168)
loc_ALL2 %>% datatable()
```


select point only in Kelantan (all_kel)

```{r}
all_kel <- loc_ALL2 %>% 
  mutate(within_kel_map = lengths(st_within(loc_ALL2, kel_map)))
all_kel2 <- all_kel %>% 
  filter(within_kel_map == 1)
```

glimpse polygon and point data

```{r}
glimpse(all_kel2)
glimpse (loc_ALL2)
```

## Descriptive 

Descriptive for all cases

```{r}
desc_all <- listALL%>%
  select(diagnosis, age, gender, race, nationality, klasifikasi_kes)
tbl_summary(desc_all)
```

Categorized age group

```{r}
desc_all$age_group <- cut(desc_all$age, breaks=c(0, 6, 9, 18, 60, Inf), 
                      labels=c("Under 5", "Children (6-9)", "Adolescent (10-19)" , "Adult (20-64)", "Elderly (65 and above)"))
```


Descriptive by diagnosis

```{r}
tbl_summary(desc_all, 
            by = diagnosis,
            label = list(age ~ "Age in years", age_group ~ "Age Category", gender ~ "Gender", race ~ "Race", nationality ~"Nationality", klasifikasi_kes ~ "Case Classification"), 
            digits = age ~ 1) %>%
  add_overall()
```


## overall plot for all cases

```{r name-of-chunk1, fig.width=18, fig.height=8}
all_plot <- ggplot() +
  geom_sf(data = kel) +
  geom_sf(data = all_kel2) +
  ggtitle("Map of Enteric Fever and Leptospirosis Cases in Kelantan for 2016-2022") +
  theme_bw()
all_plot + facet_grid(diagnosis ~ tahun_daftar) +
  theme(plot.title = element_text(size = 20),  strip.text = element_text(size = 20)) 
```


add facet

```{r}
# overall plot
all_plot + facet_wrap(~diagnosis)

# split data by diagnosis
Enteric_kel <- all_kel2 %>% 
  filter(diagnosis == "Enteric_fever") 
lepto_kel <- all_kel2 %>% 
  filter(diagnosis == "Leptospirosis") 
```


plot cumulative enteric fever cases


```{r}
Enteric_plot <- ggplot() +
  geom_sf(data = kel) +
  geom_sf(data = Enteric_kel) +
  ggtitle("Map of Enteric Fever Cases for 2016-2022") +
  theme_bw()
Enteric_plot
```


```{r name-of-chunk2, fig.width=14, fig.height=14}
Enteric_plot + 
  facet_wrap(~tahun_daftar) +
  geom_point(size=0.1) +
  theme(plot.title = element_text(size = 20),  strip.text = element_text(size = 20)) +
  ggtitle("Map of Enteric Fever Cases in Kelantan for 2016-2022")
```


plot cumulative leptospirosis cases


```{r}
lepto_kel <- all_kel2 %>% 
  filter(diagnosis == "Leptospirosis") 
lepto_plot <- ggplot() +
  geom_sf(data = kel) +
  geom_sf(data = lepto_kel) +
  ggtitle("Map of Leptospirosis Cases for 2016-2022") +
  theme_bw()
lepto_plot
```


```{r name-of-chunk3, fig.width=14, fig.height=14}
lepto_plot + 
  facet_wrap(~tahun_daftar) +
  geom_point(size=0.05) +
  theme(axis.text.x = element_text(angle = 0, hjust = 0.5, vjust = 0.5)) +
  theme(plot.title = element_text(size = 20),  strip.text = element_text(size = 20)) +
  ggtitle("Map of Leptospirosis Cases in Kelantan for 2016-2022")

```


split cumulative enteric fever cases by district

```{r name-of-chunk4, fig.width=16, fig.height=8}
# Cumulative Enteric fever cases by district
Enteric_plot + 
  facet_wrap(~DAERAH, nrow = 2) +
  theme(plot.title = element_text(size = 20),  strip.text = element_text(size = 18))
```


split cumulative leptospirosis cases by district

```{r name-of-chunk5, fig.width=16, fig.height=8}
lepto_plot + 
  facet_wrap(~DAERAH, nrow = 2) +
  theme(plot.title = element_text(size = 20),  strip.text = element_text(size = 18))
```


plot cumulative enteric fever cases by district using tmap

```{r}
tm_shape(kel) +
  tm_polygons("DAERAH", palette = rev(hcl.colors(7, "Teal"))) + 
  tm_shape(Enteric_kel) +
  tm_dots(size = 0.1) +
  tm_facets("tahun_daftar", nrow = 3, ncol = 3, as.layers = TRUE) +
  tm_layout(legend.outside = TRUE) +
  tm_layout(title = "Enteric Fever Cases in Kelantan")

```


plot cumulative leptospirosis cases by district using tmap

```{r}
#lepto plots by tmap
tm_shape(kel) +
  tm_polygons("DAERAH") + 
  tm_shape(lepto_kel) +
  tm_dots(size = 0.05) +
  tm_facets("tahun_daftar", nrow = 3, ncol = 3, as.layers = TRUE) +
  tm_layout(legend.outside = TRUE) +
  tm_layout(title = "Leptospirosis Cases in Kelantan")
```


plot combined enteric fever and leptospirosis case

```{r}
#all plot by disease
tm_shape(kel) +
  tm_polygons("NEGERI") +
  tm_facets('DAERAH') + 
  tm_shape(all_kel2) +
  tm_dots(shape = 'diagnosis', size = 0.1) 
```


## trend by year and epid week

Plot trend by year

```{r}
all_year <- all_kel2 %>% 
  group_by(diagnosis, tahun_daftar) %>% 
  count() %>% 
  print(n = 7)

ggplot(all_year, aes(x=tahun_daftar, y=n, group=diagnosis, color=as.factor(diagnosis))) +
  geom_line(size = 1) +
  labs(x="Year", y="Number of Cases", title="Enteric Fever & Leptospirosis Cases per by Year", color="Diagnosis") +
  scale_y_continuous(labels=function(x) format(x, digits=1, nsmall=0)) +
  scale_x_continuous(breaks=unique(all_year$tahun_daftar)) +
 theme_bw()
```


Plot line graph Enteric fever by epid week

```{r}
ent_week <- Enteric_kel %>% 
  group_by(tahun_daftar, epid_daftar) %>% 
  count() %>% 
  print(n = 52)

ggplot(ent_week, aes(x=epid_daftar, y=n, group=tahun_daftar, color=as.factor(tahun_daftar))) +
  geom_line(size = 1) +
  labs(x="Epid Week", y="Number of Cases", title="Enteric Fever Cases per Epid Week Separated by Year", color="Year") +
  scale_color_brewer(palette="Dark2") + 
  scale_y_continuous(labels=function(x) format(x, digits=1, nsmall=0)) +
 theme_bw()
  
```

```{r}
lep_week <- lepto_kel %>% 
  group_by(tahun_daftar, epid_daftar) %>% 
  count() %>% 
  print(n = 52)

ggplot(lep_week, aes(x=epid_daftar, y=n, group=tahun_daftar, color=as.factor(tahun_daftar))) +
  geom_line(size = 0.8) +
  labs(x="Epid Week", y="Number of Cases", title="Enteric Fever Cases per Epid Week Separated by Year", color="Year") +
  scale_color_brewer(palette="Set2") + 
  scale_y_continuous(labels=function(x) format(x, digits=1, nsmall=0)) +
 theme_bw()
```


## Density of cases

# Enteric fever density per subdistrict (mukim)

Joint point to polygon

```{r}
ent_in_muk <- st_join(Enteric_kel, kel_map, 
                      join = st_within)
glimpse(ent_in_muk)
```


count all enteric fever in mukim

```{r}
count_ent_mukim_yr <- ent_in_muk %>% 
  count(DAERAH, MUKIM, tahun_daftar, JUMLAH_2016, JUMLAH_2017, JUMLAH_2018, JUMLAH_2019, JUMLAH_2020, JUMLAH_2021, JUMLAH_2022, AVR) %>% 
  ungroup()
count_ent_mukim_yr %>% datatable()
```


Calculate incidence of enteric fever per 1000 population for mukim

```{r}
count_ent_muk_y_1000 <- count_ent_mukim_yr %>% 
  mutate(incidence_ent = (n/AVR)*1000)
count_ent_muk_y_1000 %>% datatable()
```


Calculate incidence of enteric fever per 1000 population for mukim per year +
join polygon to point

```{r}
# plot incidence enteric fever 
count_ent_mukim <- ent_in_muk %>% 
  count(DAERAH, MUKIM, AVR) %>% 
  ungroup()
count_ent_mukim_1000 <- count_ent_mukim %>% 
  mutate(inc_1000 = (n/AVR)*1000)
kelmap_with_ent <- st_join(kel_map, count_ent_mukim_1000)
glimpse(kelmap_with_ent)

# Incidence Enteric fever in 2016
count_ent_mukim_y16 <- subset(count_ent_mukim_yr, tahun_daftar=='2016') 
count_ent_mukim_y16_1000 <- count_ent_mukim_y16 %>% 
  mutate(inc_1000 = (n/JUMLAH_2016)*1000)
kelmap_with_ent_16 <- st_join(kel_map, count_ent_mukim_y16_1000)

# Incidence Enteric fever in 2017
count_ent_mukim_y17 <- subset(count_ent_mukim_yr, tahun_daftar=='2017') 
count_ent_mukim_y17_1000 <- count_ent_mukim_y17 %>% 
  mutate(inc_1000 = (n/JUMLAH_2017)*1000)
kelmap_with_ent_17 <- st_join(kel_map, count_ent_mukim_y17_1000)

# Incidence Enteric fever in 2018
count_ent_mukim_y18 <- subset(count_ent_mukim_yr, tahun_daftar=='2018') 
count_ent_mukim_y18_1000 <- count_ent_mukim_y18 %>% 
  mutate(inc_1000 = (n/JUMLAH_2018)*1000)
kelmap_with_ent_18 <- st_join(kel_map, count_ent_mukim_y18_1000)

# Incidence Enteric fever in 2019
count_ent_mukim_y19 <- subset(count_ent_mukim_yr, tahun_daftar=='2019') 
count_ent_mukim_y19_1000 <- count_ent_mukim_y19 %>% 
  mutate(inc_1000 = (n/JUMLAH_2019)*1000)
kelmap_with_ent_19 <- st_join(kel_map, count_ent_mukim_y19_1000)

# Incidence Enteric fever in 2020
count_ent_mukim_y20 <- subset(count_ent_mukim_yr, tahun_daftar=='2020') 
count_ent_mukim_y20_1000 <- count_ent_mukim_y20 %>% 
  mutate(inc_1000 = (n/JUMLAH_2020)*1000)
kelmap_with_ent_20 <- st_join(kel_map, count_ent_mukim_y20_1000)

# Incidence Enteric fever in 2021
count_ent_mukim_y21 <- subset(count_ent_mukim_yr, tahun_daftar=='2021') 
count_ent_mukim_y21_1000 <- count_ent_mukim_y21 %>% 
  mutate(inc_1000 = (n/JUMLAH_2021)*1000)
kelmap_with_ent_21 <- st_join(kel_map, count_ent_mukim_y21_1000)

# Incidence Enteric fever in 2022
count_ent_mukim_y22 <- subset(count_ent_mukim_yr, tahun_daftar=='2022') 
count_ent_mukim_y22_1000 <- count_ent_mukim_y22 %>% 
  mutate(inc_1000 = (n/JUMLAH_2021)*1000)
kelmap_with_ent_22 <- st_join(kel_map, count_ent_mukim_y22_1000)
```


plot incidence map Enteric fever 2016-2022

```{r}
#| cache = TRUE
incplot_ent <- ggplot() + 
  geom_sf(data = kelmap_with_ent, aes(fill = inc_1000)) +
  scale_fill_gradientn(colors = sf.colors(20)) +
  ggtitle('2016-2022') +
  theme_bw() 
incplot_ent

#plot incidence map Enteric fever 2016
incplot_ent16 <- ggplot() + 
  geom_sf(data = kelmap_with_ent_16, aes(fill = inc_1000)) +
  scale_fill_gradientn(colors = sf.colors(20)) +
  ggtitle('2016') +
  theme_bw() 
incplot_ent16

#plot incidence map Enteric fever 2017
incplot_ent17 <- ggplot() + 
  geom_sf(data = kelmap_with_ent_17, aes(fill = inc_1000)) +
  scale_fill_gradientn(colors = sf.colors(20)) +
  ggtitle('2017') +
  theme_bw()
incplot_ent17

#plot incidence map Enteric fever 2018
incplot_ent18 <- ggplot() + 
  geom_sf(data = kelmap_with_ent_18, aes(fill = inc_1000)) +
  scale_fill_gradientn(colors = sf.colors(20)) +
  ggtitle('2018') +
  theme_bw()
incplot_ent18

#plot incidence map Enteric fever 2019
incplot_ent19 <- ggplot() + 
  geom_sf(data = kelmap_with_ent_19, aes(fill = inc_1000)) +
  scale_fill_gradientn(colors = sf.colors(20)) +
  ggtitle('2019') +
  theme_bw()
incplot_ent19

#plot incidence map Enteric fever 2020
incplot_ent20 <- ggplot() + 
  geom_sf(data = kelmap_with_ent_20, aes(fill = inc_1000)) +
  scale_fill_gradientn(colors = sf.colors(20)) +
  ggtitle('2020') +
  theme_bw()
incplot_ent20

#plot incidence map Enteric fever 2021
incplot_ent21 <- ggplot() + 
  geom_sf(data = kelmap_with_ent_21, aes(fill = inc_1000)) +
  scale_fill_gradientn(colors = sf.colors(20)) +
  ggtitle('2021') +
  theme_bw()
incplot_ent21

#plot incidence map Enteric fever 2022
incplot_ent22 <- ggplot() + 
  geom_sf(data = kelmap_with_ent_22, aes(fill = inc_1000)) +
  scale_fill_gradientn(colors = sf.colors(20)) +
  ggtitle('2022') +
  theme_bw()
incplot_ent22
```


combine incidence map Enteric fever 2016-2022 in one view 

```{r name-of-chunk7, fig.width=16, fig.height=9}
grid.arrange(incplot_ent16, incplot_ent17, incplot_ent18, incplot_ent19, incplot_ent20, incplot_ent21, incplot_ent22,incplot_ent, nrow=2, top=textGrob("Incidence Map of Enteric Fever in Kelantan per 1000 population 2016-2022", gp = gpar(fontsize = 20)))
```


# Leptospirosis density per subdistrict (mukim)

Joint point to polygon
```{r}
#lepto density per mukim
lep_in_muk <- st_join(lepto_kel, kel_map, 
                      join = st_within)
glimpse(lep_in_muk)
```


count all lepto cases in mukim

```{r}
count_lep_mukim_yr <- lep_in_muk %>% 
  count(DAERAH, MUKIM, tahun_daftar, JUMLAH_2016, JUMLAH_2017, JUMLAH_2018, JUMLAH_2019, JUMLAH_2020, JUMLAH_2021, JUMLAH_2022, AVR) %>% 
  ungroup()
count_lep_mukim_yr %>% datatable()
```


Calculate incidence of leptospirosis per 1000 population for mukim+year

```{r}
count_lep_muk_y_1000 <- count_lep_mukim_yr %>% 
  mutate(incidence_lep = (n/AVR)*1000)
count_lep_muk_y_1000 %>% datatable()
```


Calculate incidence of enteric fever per 1000 population for mukim per year +
join polygon to point

```{r}
#| cache = TRUE
count_lep_mukim <- lep_in_muk %>% 
  count(DAERAH, MUKIM, tahun_daftar, JUMLAH_2016, JUMLAH_2017, JUMLAH_2018, JUMLAH_2019, JUMLAH_2020, JUMLAH_2021, JUMLAH_2022, AVR) %>% 
  ungroup()
count_lep_mukim_1000 <- count_lep_mukim %>% 
  mutate(inc_1000 = (n/AVR)*1000)
kelmap_with_lep <- st_join(kel_map, count_lep_mukim_1000)
glimpse(kelmap_with_lep)

# Incidence Leptospirosis in 2016
count_lep_mukim_y16 <- subset(count_lep_mukim_yr, tahun_daftar=='2016') 
count_lep_mukim_y16_1000 <- count_lep_mukim_y16 %>% 
  mutate(inc_1000 = (n/JUMLAH_2016)*1000)
kelmap_with_lep_16 <- st_join(kel_map, count_lep_mukim_y16_1000)

# Incidence Leptospirosis in 2017
count_lep_mukim_y17 <- subset(count_lep_mukim_yr, tahun_daftar=='2017') 
count_lep_mukim_y17_1000 <- count_lep_mukim_y17 %>% 
  mutate(inc_1000 = (n/JUMLAH_2017)*1000)
kelmap_with_lep_17 <- st_join(kel_map, count_lep_mukim_y17_1000)

# Incidence Leptospirosis in 2018
count_lep_mukim_y18 <- subset(count_lep_mukim_yr, tahun_daftar=='2018') 
count_lep_mukim_y18_1000 <- count_lep_mukim_y18 %>% 
  mutate(inc_1000 = (n/JUMLAH_2018)*1000)
kelmap_with_lep_18 <- st_join(kel_map, count_lep_mukim_y18_1000)

# Incidence Leptospirosis in 2019
count_lep_mukim_y19 <- subset(count_lep_mukim_yr, tahun_daftar=='2019') 
count_lep_mukim_y19_1000 <- count_lep_mukim_y19 %>% 
  mutate(inc_1000 = (n/JUMLAH_2019)*1000)
kelmap_with_lep_19 <- st_join(kel_map, count_lep_mukim_y19_1000)

# Incidence Leptospirosis in 2020
count_lep_mukim_y20 <- subset(count_lep_mukim_yr, tahun_daftar=='2020') 
count_lep_mukim_y20_1000 <- count_lep_mukim_y20 %>% 
  mutate(inc_1000 = (n/JUMLAH_2020)*1000)
kelmap_with_lep_20 <- st_join(kel_map, count_lep_mukim_y20_1000)

# Incidence Leptospirosis in 2021
count_lep_mukim_y21 <- subset(count_lep_mukim_yr, tahun_daftar=='2021') 
count_lep_mukim_y21_1000 <- count_lep_mukim_y21 %>% 
  mutate(inc_1000 = (n/JUMLAH_2021)*1000)
kelmap_with_lep_21 <- st_join(kel_map, count_lep_mukim_y21_1000)

# Incidence Leptospirosis in 2022
count_lep_mukim_y22 <- subset(count_lep_mukim_yr, tahun_daftar=='2022') 
count_lep_mukim_y22_1000 <- count_lep_mukim_y22 %>% 
  mutate(inc_1000 = (n/JUMLAH_2020)*1000)
kelmap_with_lep_22 <- st_join(kel_map, count_lep_mukim_y22_1000)
```


plot incidence map Leptospirosis 2016-2022

```{r}
#| cache = TRUE
incplot_lep <- ggplot() + 
  geom_sf(data = kelmap_with_lep, aes(fill = inc_1000)) +
  scale_fill_gradientn(colors = sf.colors(20)) +
  ggtitle('2016-2022') +
  theme_bw()
incplot_lep 

# plot incidence map Leptospirosis 2016
incplot_lep16 <- ggplot() + 
  geom_sf(data = kelmap_with_lep_16, aes(fill = inc_1000)) +
  scale_fill_gradientn(colors = sf.colors(20)) +
  ggtitle('2016') +
  theme_bw()
incplot_lep16

#plot incidence map Leptospirosis 2017
incplot_lep17 <- ggplot() + 
  geom_sf(data = kelmap_with_lep_17, aes(fill = inc_1000)) +
  scale_fill_gradientn(colors = sf.colors(20)) +
  ggtitle('2017') +
  theme_bw()
incplot_lep17

#plot incidence map Leptospirosis 2018
incplot_lep18 <- ggplot() + 
  geom_sf(data = kelmap_with_lep_18, aes(fill = inc_1000)) +
  scale_fill_gradientn(colors = sf.colors(20)) +
  ggtitle('2018') +
  theme_bw()
incplot_lep18

#plot incidence map Leptospirosis 2019
incplot_lep19 <- ggplot() + 
  geom_sf(data = kelmap_with_lep_19, aes(fill = inc_1000)) +
  scale_fill_gradientn(colors = sf.colors(20)) +
  ggtitle('2019') +
  theme_bw()
incplot_lep19

#plot incidence map Leptospirosis 2020
incplot_lep20 <- ggplot() + 
  geom_sf(data = kelmap_with_lep_20, aes(fill = inc_1000)) +
  scale_fill_gradientn(colors = sf.colors(20)) +
  ggtitle('2020') +
  theme_bw()
incplot_lep20

#plot incidence map Leptospirosis 2021
incplot_lep21 <- ggplot() + 
  geom_sf(data = kelmap_with_lep_21, aes(fill = inc_1000)) +
  scale_fill_gradientn(colors = sf.colors(20)) +
  ggtitle('2021') +
  theme_bw()
incplot_lep21

#plot incidence map Leptospirosis 2022
incplot_lep22 <- ggplot() + 
  geom_sf(data = kelmap_with_lep_22, aes(fill = inc_1000)) +
  scale_fill_gradientn(colors = sf.colors(20)) +
  ggtitle('2022') +
  theme_bw()
incplot_lep22
```


combine incidence maps Leptospirosis 2016-2022 in one view

```{r name-of-chunk8, fig.width=16, fig.height=9}
# combine all leptospirosis incidence plot  in one view
grid.arrange(incplot_lep16, incplot_lep17, incplot_lep18, incplot_lep19, incplot_lep20, incplot_lep21, incplot_lep22,incplot_lep, nrow=2, top=textGrob("Incidence Map of Leptospirosis in Kelantan per 1000 population 2016-2022", gp = gpar(fontsize = 20)))

```

## Point Pattern Analysis

# Prepare data

convert data to spatial format

```{r}
kel_map.sp <- as(kel_map, "Spatial")
class(kel_map.sp)
```

```{r}
plot(kel_map.sp)
```

convert point to spatial

```{r}
loc_ALL2.sp <- as(loc_ALL2, "Spatial")
class(loc_ALL2.sp)
```

# point pattern analysis for enteric fever

extract enteric fever by year

```{r}
loc_ent16 <- subset(Enteric_kel, tahun_daftar=="2016")
loc_ent17 <- subset(Enteric_kel, tahun_daftar=="2017")
loc_ent18 <- subset(Enteric_kel, tahun_daftar=="2018")
loc_ent19 <- subset(Enteric_kel, tahun_daftar=="2019")
loc_ent20 <- subset(Enteric_kel, tahun_daftar=="2020")
loc_ent21 <- subset(Enteric_kel, tahun_daftar=="2021")
loc_ent22 <- subset(Enteric_kel, tahun_daftar=="2022")
```


convert point to spatial

```{r}
loc_ent16.sp <- as(loc_ent16, "Spatial")
loc_ent17.sp <- as(loc_ent17, "Spatial")
loc_ent18.sp <- as(loc_ent18, "Spatial")
loc_ent19.sp <- as(loc_ent19, "Spatial")
loc_ent20.sp <- as(loc_ent20, "Spatial")
loc_ent21.sp <- as(loc_ent21, "Spatial")
loc_ent22.sp <- as(loc_ent22, "Spatial")
```


convert point spatial data to ppp format

```{r}
loc_ALL2.ppp <- as(loc_ALL2.sp, 'ppp')
class(loc_ALL2.ppp)
loc_ent16.ppp <- as(loc_ent16.sp, 'ppp')
loc_ent17.ppp <- as(loc_ent17.sp, 'ppp')
loc_ent18.ppp <- as(loc_ent18.sp, 'ppp')
loc_ent19.ppp <- as(loc_ent19.sp, 'ppp')
loc_ent20.ppp <- as(loc_ent20.sp, 'ppp')
loc_ent21.ppp <- as(loc_ent21.sp, 'ppp')
loc_ent22.ppp <- as(loc_ent22.sp, 'ppp')
```

transform to owin format

```{r}
kel_map.owin <- as(kel_map.sp, "owin")
class(kel_map.owin)
```


plot ppp object with window 

```{r}
# plot ALL
plot(kel_map.owin)
points(loc_ALL2.ppp)
```

convert enteric fever and lepto pint data to spatial to ppp

```{r}
Enteric_kel.sp <- as(Enteric_kel, "Spatial")
Enteric_kel.ppp <- as(Enteric_kel.sp, 'ppp')
lepto_kel.sp <- as(lepto_kel, "Spatial")
lepto_kel.ppp <- as(lepto_kel.sp, 'ppp')
```


plot enteric fever with window

```{r}
# plot Enteric Fever 2016-2022
plot(kel_map.owin)
points(Enteric_kel.ppp)
```

Remove marks

```{r}
# remove marks enteric fever 2016-2022
Enteric_kel.ppp2 <- Enteric_kel.ppp
marks(Enteric_kel.ppp2) <- NULL

# remove marks (year)
loc_ent16.ppp2 <- loc_ent16.ppp
marks(loc_ent16.ppp2) <- NULL
loc_ent17.ppp2 <- loc_ent17.ppp
marks(loc_ent17.ppp2) <- NULL
loc_ent18.ppp2 <- loc_ent18.ppp
marks(loc_ent18.ppp2) <- NULL
loc_ent19.ppp2 <- loc_ent19.ppp
marks(loc_ent19.ppp2) <- NULL
loc_ent20.ppp2 <- loc_ent20.ppp
marks(loc_ent20.ppp2) <- NULL
loc_ent21.ppp2 <- loc_ent21.ppp
marks(loc_ent21.ppp2) <- NULL
loc_ent22.ppp2 <- loc_ent22.ppp
marks(loc_ent22.ppp2) <- NULL
```

```{r}
# generate window
Window(Enteric_kel.ppp2) <- kel_map.owin

# generate window (year)
Window(loc_ent16.ppp2) <- kel_map.owin
Window(loc_ent17.ppp2) <- kel_map.owin
Window(loc_ent18.ppp2) <- kel_map.owin
Window(loc_ent19.ppp2) <- kel_map.owin
Window(loc_ent20.ppp2) <- kel_map.owin
Window(loc_ent21.ppp2) <- kel_map.owin
Window(loc_ent22.ppp2) <- kel_map.owin
```


Plot enteric fever by year

```{r name-of-chunk9, fig.width=20, fig.height=11}
par(mfrow=c(2,4))
plot(loc_ent16.ppp2, main = 2016, cols=rgb(0,0,0,.2), pch=20)
plot(loc_ent17.ppp2, main = 2017, cols=rgb(0,0,0,.2), pch=20)
plot(loc_ent18.ppp2, main = 2018, cols=rgb(0,0,0,.2), pch=20)
plot(loc_ent19.ppp2, main = 2019, cols=rgb(0,0,0,.2), pch=20)
plot(loc_ent20.ppp2, main = 2020, cols=rgb(0,0,0,.2), pch=20)
plot(loc_ent21.ppp2, main = 2021, cols=rgb(0,0,0,.2), pch=20)
plot(loc_ent22.ppp2, main = 2022, cols=rgb(0,0,0,.2), pch=20)
plot(Enteric_kel.ppp2, main = "2016-2022", cols=rgb(0,0,0,.2), pch=20)
```

# density analysis for enteric fever

quadrat count + plot

```{r name-of-chunk10, fig.width=28, fig.height=16}
par( mfrow= c(2,4) )
# 2016
quadr_count_ent16 <- quadratcount(loc_ent16.ppp2, 
                                  nx = 10,
                                  ny = 14)
plot(loc_ent16.ppp2, pch = 20, cex = 0.1, main = "2016")
plot(quadr_count_ent16, add = TRUE, cex = 2)

# 2017
quadr_count_ent17 <- quadratcount(loc_ent17.ppp2, 
                                  nx = 10,
                                  ny = 14)
plot(loc_ent17.ppp2, pch = 20, cex = 0.1, main = "2017")
plot(quadr_count_ent17, add = TRUE, cex = 2)

# 2018
quadr_count_ent18 <- quadratcount(loc_ent18.ppp2, 
                                  nx = 10,
                                  ny = 14)
plot(loc_ent18.ppp2, pch = 20, cex = 0.1, main = "2018")
plot(quadr_count_ent18, add = TRUE, cex = 2)

# 2019
quadr_count_ent19 <- quadratcount(loc_ent19.ppp2, 
                                  nx = 10,
                                  ny = 14)
plot(loc_ent19.ppp2, pch = 20, cex = 0.1, main = "2019")
plot(quadr_count_ent19, add = TRUE, cex = 2)

# 2020
quadr_count_ent20 <- quadratcount(loc_ent20.ppp2, 
                                  nx = 10,
                                  ny = 14)
plot(loc_ent20.ppp2, pch = 20, cex = 0.1, main = "2020")
plot(quadr_count_ent20, add = TRUE, cex = 2)

# 2021
quadr_count_ent21 <- quadratcount(loc_ent21.ppp2, 
                                  nx = 10,
                                  ny = 14)
plot(loc_ent21.ppp2, pch = 20, cex = 0.1, main = "2021")
plot(quadr_count_ent21, add = TRUE, cex = 2)

# 2022
quadr_count_ent22 <- quadratcount(loc_ent22.ppp2, 
                                  nx = 10,
                                  ny = 14)
plot(loc_ent22.ppp2, pch = 20, cex = 0.1, main = "2022")
plot(quadr_count_ent22, add = TRUE, cex = 2)

# cumulative cases 2016-2022
quadr_count_ent <- quadratcount(Enteric_kel.ppp2, 
                                nx = 10,
                                ny = 14)
plot(Enteric_kel.ppp2, pch = 20, cex = 0.1, main = "2016-2022")
plot(quadr_count_ent, add = TRUE, cex = 2)

mtext("Point Density Maps of Enteric Fever Cases in Kelantan 2016-2022", side = 1, line = -1, cex = 2, outer = TRUE)
```


calculate VMR (Variance-to-Mean Ratio) enteric fever

- VMR greater than 1 indicates spatial clustering.
- Enteric Fever not uniformly distributed and may be clustered in certain areas.

```{r}
#2016-2022
Qcountent<-data.frame(quadr_count_ent)
var(Qcountent$Freq)/mean(Qcountent$Freq)

#2016
Qcountent16<-data.frame(quadr_count_ent16)
var(Qcountent16$Freq)/mean(Qcountent16$Freq)

#2017
Qcountent17<-data.frame(quadr_count_ent17)
var(Qcountent17$Freq)/mean(Qcountent17$Freq)

#2018
Qcountent18<-data.frame(quadr_count_ent18)
var(Qcountent18$Freq)/mean(Qcountent18$Freq)

#2019
Qcountent19<-data.frame(quadr_count_ent19)
var(Qcountent19$Freq)/mean(Qcountent19$Freq)

#2020
Qcountent20<-data.frame(quadr_count_ent20)
var(Qcountent20$Freq)/mean(Qcountent20$Freq)

#2021
Qcountent21<-data.frame(quadr_count_ent21)
var(Qcountent21$Freq)/mean(Qcountent21$Freq)

#2022
Qcountent22<-data.frame(quadr_count_ent22)
var(Qcountent22$Freq)/mean(Qcountent22$Freq)
```
calculate Chi-square test for spatial randomnes - enteric fever

- a high VMR score and a significant chi-square test result would indicate that the points are clustered and non-randomly distributed, suggesting the presence of spatial processes such as spatial contagion, spatial dependence, or spatial interaction.

```{r}
#| cache = TRUE
chi_ent <- quadrat.test(Enteric_kel.ppp2, nx= 30, ny=15)
chi_ent16 <- quadrat.test(loc_ent16.ppp2, nx= 30, ny=15)
chi_ent17 <- quadrat.test(loc_ent17.ppp2, nx= 30, ny=15)
chi_ent18 <- quadrat.test(loc_ent18.ppp2, nx= 30, ny=15)
chi_ent19 <- quadrat.test(loc_ent19.ppp2, nx= 30, ny=15)
chi_ent20 <- quadrat.test(loc_ent20.ppp2, nx= 30, ny=15)
chi_ent21 <- quadrat.test(loc_ent21.ppp2, nx= 30, ny=15)
chi_ent22 <- quadrat.test(loc_ent22.ppp2, nx= 30, ny=15)
```
result chi square for enteric fever

```{r}
chi_ent.df <- data.frame(Dataset = "Enteric_kel.ppp2",
                       TestStatistic = chi_ent$statistic,
                       PValue = chi_ent$p.value)
chi_ent16.df <- data.frame(Dataset = "loc_ent16.ppp2",
                       TestStatistic = chi_ent16$statistic,
                       PValue = chi_ent16$p.value)
chi_ent17.df <- data.frame(Dataset = "loc_ent17.ppp2",
                       TestStatistic = chi_ent17$statistic,
                       PValue = chi_ent17$p.value)
chi_ent18.df <- data.frame(Dataset = "loc_ent18.ppp2",
                       TestStatistic = chi_ent18$statistic,
                       PValue = chi_ent18$p.value)
chi_ent19.df <- data.frame(Dataset = "loc_ent19.ppp2",
                       TestStatistic = chi_ent19$statistic,
                       PValue = chi_ent19$p.value)
chi_ent20.df <- data.frame(Dataset = "loc_ent20.ppp2",
                       TestStatistic = chi_ent20$statistic,
                       PValue = chi_ent20$p.value)
chi_ent21.df <- data.frame(Dataset = "loc_ent21.ppp2",
                       TestStatistic = chi_ent21$statistic,
                       PValue = chi_ent21$p.value)
chi_ent22.df <- data.frame(Dataset = "loc_ent22.ppp2",
                       TestStatistic = chi_ent22$statistic,
                       PValue = chi_ent22$p.value)
chi_quadENT <- bind_rows(
  data.frame(Dataset = "2016", chi_ent16.df), 
  data.frame(Dataset = "2017", chi_ent17.df), 
  data.frame(Dataset = "2018", chi_ent18.df), 
  data.frame(Dataset = "2019", chi_ent19.df), 
  data.frame(Dataset = "2020", chi_ent20.df), 
  data.frame(Dataset = "2021", chi_ent21.df), 
  data.frame(Dataset = "2022", chi_ent22.df), 
  data.frame(Dataset = "2016-2022", chi_ent.df))
chi_quadENT

```


Monte Carlo permutation for spatial randomnes - enteric fever (chi square assumption not met)

```{r}
#| cache = TRUE
# calculate monte carlo
mc_qent <- quadrat.test(Enteric_kel.ppp2, nx= 30, ny=15, method = "MonteCarlo")
mc_qent16 <- quadrat.test(loc_ent16.ppp2, nx= 30, ny=15, method = "MonteCarlo")
mc_qent17 <- quadrat.test(loc_ent17.ppp2, nx= 30, ny=15, method = "MonteCarlo")
mc_qent18 <- quadrat.test(loc_ent18.ppp2, nx= 30, ny=15, method = "MonteCarlo")
mc_qent19 <- quadrat.test(loc_ent19.ppp2, nx= 30, ny=15, method = "MonteCarlo")
mc_qent20 <- quadrat.test(loc_ent20.ppp2, nx= 30, ny=15, method = "MonteCarlo")
mc_qent21 <- quadrat.test(loc_ent21.ppp2, nx= 30, ny=15, method = "MonteCarlo")
mc_qent22 <- quadrat.test(loc_ent22.ppp2, nx= 30, ny=15, method = "MonteCarlo")

# convert to data frame
mc_qent.df <- data.frame(Dataset = "Enteric_kel.ppp2",
                       TestStatistic = mc_qent$statistic,
                       PValue = mc_qent$p.value)
mc_qent16.df <- data.frame(Dataset = "loc_ent16.ppp2",
                       TestStatistic = mc_qent16$statistic,
                       PValue = mc_qent16$p.value)
mc_qent17.df <- data.frame(Dataset = "loc_ent17.ppp2",
                       TestStatistic = mc_qent17$statistic,
                       PValue = mc_qent17$p.value)
mc_qent18.df <- data.frame(Dataset = "loc_ent18.ppp2",
                       TestStatistic = mc_qent18$statistic,
                       PValue = mc_qent18$p.value)
mc_qent19.df <- data.frame(Dataset = "loc_ent19.ppp2",
                       TestStatistic = mc_qent19$statistic,
                       PValue = mc_qent19$p.value)
mc_qent20.df <- data.frame(Dataset = "loc_ent20.ppp2",
                       TestStatistic = mc_qent20$statistic,
                       PValue = mc_qent20$p.value)
mc_qent21.df <- data.frame(Dataset = "loc_ent21.ppp2",
                       TestStatistic = mc_qent21$statistic,
                       PValue = mc_qent21$p.value)
mc_qent22.df <- data.frame(Dataset = "loc_ent22.ppp2",
                       TestStatistic = mc_qent22$statistic,
                       PValue = mc_qent22$p.value)

# combine rows
mc_qENT <- bind_rows(
  data.frame(Dataset = "2016", mc_qent16.df), 
  data.frame(Dataset = "2017", mc_qent17.df), 
  data.frame(Dataset = "2018", mc_qent18.df), 
  data.frame(Dataset = "2019", mc_qent19.df), 
  data.frame(Dataset = "2020", mc_qent20.df), 
  data.frame(Dataset = "2021", mc_qent21.df), 
  data.frame(Dataset = "2022", mc_qent22.df), 
  data.frame(Dataset = "2016-2022", mc_qent.df))
mc_qENT
```

# Intensity analysis for enteric fever

```{r name-of-chunk11, fig.width=16, fig.height=8}
par( mfrow= c(2,4) )

inten_ent16 <-intensity(quadr_count_ent16)
plot(intensity(quadr_count_ent16, image = TRUE), main = "2016", las = 1)
plot(loc_ent16, pch = 20, cex = 0.6, add = TRUE)

inten_ent17 <-intensity(quadr_count_ent17)
plot(intensity(quadr_count_ent17, image = TRUE), main = "2017", las = 1)
plot(loc_ent17, pch = 20, cex = 0.6, add = TRUE)

inten_ent18 <-intensity(quadr_count_ent18)
plot(intensity(quadr_count_ent18, image = TRUE), main = "2018", las = 1)
plot(loc_ent18, pch = 20, cex = 0.6, add = TRUE)

inten_ent19 <-intensity(quadr_count_ent19)
plot(intensity(quadr_count_ent19, image = TRUE), main = "2019", las = 1)
plot(loc_ent19, pch = 20, cex = 0.6, add = TRUE)

inten_ent20 <-intensity(quadr_count_ent20)
plot(intensity(quadr_count_ent20, image = TRUE), main = "2020", las = 1)
plot(loc_ent20, pch = 20, cex = 0.6, add = TRUE)

inten_ent21 <-intensity(quadr_count_ent21)
plot(intensity(quadr_count_ent21, image = TRUE), main = "2021", las = 1)
plot(loc_ent21, pch = 20, cex = 0.6, add = TRUE)

inten_ent22 <-intensity(quadr_count_ent22)
plot(intensity(quadr_count_ent22, image = TRUE), main = "2022", las = 1)
plot(loc_ent22, pch = 20, cex = 0.6, add = TRUE)

inten_ent <-intensity(quadr_count_ent)
plot(intensity(quadr_count_ent, image = TRUE), main = "2016-2022", las = 1)
plot(Enteric_kel.ppp2, pch = 20, cex = 0.6, add = TRUE)

mtext("Intensity Maps of Enteric Fever Cases in Kelantan 2016-2022", side = 1, line = -1, cex = 1.5, outer = TRUE)
```

# Kernel density raster for enteric fever

rescale to km

```{r}
# point
Enteric_kel.ppp2.km <- rescale(Enteric_kel.ppp2, 1000, 'km') #cumulative
loc_ent16.ppp2.km <- rescale(loc_ent16.ppp2, 1000, 'km') #2016
loc_ent17.ppp2.km <- rescale(loc_ent17.ppp2, 1000, 'km') #2017
loc_ent18.ppp2.km <- rescale(loc_ent18.ppp2, 1000, 'km') #2018
loc_ent19.ppp2.km <- rescale(loc_ent19.ppp2, 1000, 'km') #2019
loc_ent20.ppp2.km <- rescale(loc_ent20.ppp2, 1000, 'km') #2020
loc_ent21.ppp2.km <- rescale(loc_ent21.ppp2, 1000, 'km') #2021
loc_ent22.ppp2.km <- rescale(loc_ent22.ppp2, 1000, 'km') #2022

# map
kel_map.owin.km <- rescale(kel_map.owin, 1000, 'km') 
```


calculate kernel density estimate (kde) for Enteric Fever

```{r}
kde.ent <- density(Enteric_kel.ppp2.km) #2016 - 2022
kde.ent16 <- density(loc_ent16.ppp2.km) #2016
kde.ent17 <- density(loc_ent17.ppp2.km) #2017
kde.ent18 <- density(loc_ent18.ppp2.km) #2018
kde.ent19 <- density(loc_ent19.ppp2.km) #2019
kde.ent20 <- density(loc_ent20.ppp2.km) #2020
kde.ent21 <- density(loc_ent21.ppp2.km) #2021
kde.ent22 <- density(loc_ent22.ppp2.km) #2022
```


plot kde Enteric fever

```{r name-of-chunk12, fig.width=16, fig.height=8}
par(mfrow=c(2,4))
plot(kde.ent16, main = 2016, las = 1) #2016
plot(kde.ent17, main = 2017, las = 1) #2017
plot(kde.ent18, main = 2018, las = 1) #2018
plot(kde.ent19, main = 2019, las = 1) #2019
plot(kde.ent20, main = 2020, las = 1) #2020
plot(kde.ent21, main = 2021, las = 1) #2021
plot(kde.ent22, main = 2022, las = 1) #2022
plot(kde.ent, main = "2016-2022", las = 1) #2016-2022
```

kde 20km bandwith for enteric fever

```{r name-of-chunk13, fig.width=16, fig.height=8}
kde.ent.20km <- density(Enteric_kel.ppp2.km, sigma = 20) #2016-2022
kde.ent16.20km <- density(loc_ent16.ppp2.km, sigma = 20) #2016
kde.ent17.20km <- density(loc_ent17.ppp2.km, sigma = 20) #2017
kde.ent18.20km <- density(loc_ent18.ppp2.km, sigma = 20) #2018
kde.ent19.20km <- density(loc_ent19.ppp2.km, sigma = 20) #2019
kde.ent20.20km <- density(loc_ent20.ppp2.km, sigma = 20) #2020
kde.ent21.20km <- density(loc_ent21.ppp2.km, sigma = 20) #2021
kde.ent22.20km <- density(loc_ent22.ppp2.km, sigma = 20) #2022

# plot kde 20km

par(mfrow=c(2,4))
# 2016
plot(kde.ent16.20km, main = 2016, las = 1)
contour(kde.ent16.20km, add = TRUE)
# 2017
plot(kde.ent17.20km, main = 2017, las = 1)
contour(kde.ent17.20km, add = TRUE)
# 2018
plot(kde.ent18.20km, main = 2018, las = 1)
contour(kde.ent18.20km, add = TRUE)
# 2019
plot(kde.ent19.20km, main = 2019, las = 1)
contour(kde.ent19.20km, add = TRUE)
# 2020
plot(kde.ent20.20km, main = 2020, las = 1)
contour(kde.ent20.20km, add = TRUE)
# 2021
plot(kde.ent21.20km, main = 2021, las = 1)
contour(kde.ent21.20km, add = TRUE)
# 2022
plot(kde.ent22.20km, main = 2022, las = 1)
contour(kde.ent22.20km, add = TRUE)
# 2016-2022
plot(kde.ent.20km, main = "2016-2022", las = 1)
contour(kde.ent.20km, add = TRUE)
```


calculate kde 5km bandwith enteric fever

```{r}
#| cache = TRUE
kde.ent.5km <- density(Enteric_kel.ppp2.km, sigma = 5) #2016-2022
kde.ent16.5km <- density(loc_ent16.ppp2.km, sigma = 5) #2016
kde.ent17.5km <- density(loc_ent17.ppp2.km, sigma = 5) #2017
kde.ent18.5km <- density(loc_ent18.ppp2.km, sigma = 5) #2018
kde.ent19.5km <- density(loc_ent19.ppp2.km, sigma = 5) #2019
kde.ent20.5km <- density(loc_ent20.ppp2.km, sigma = 5) #2020
kde.ent21.5km <- density(loc_ent21.ppp2.km, sigma = 5) #2021
kde.ent22.5km <- density(loc_ent22.ppp2.km, sigma = 5) #2022
```


plot kde 5km bandwith enteric fever

```{r name-of-chunk14, fig.width=20, fig.height=11}
# plot all KDE enteric fever in one view
par( mfrow= c(2,4) )
plot(kde.ent16.5km, main = "2016", cex.main = 1.5, font.main = 2, las = 1)
contour(kde.ent16.5km, add = TRUE)
plot(kde.ent17.5km, main = "2017", cex.main = 1.5, font.main = 2, las = 1)
contour(kde.ent17.5km, add = TRUE)
plot(kde.ent18.5km, main = "2018", cex.main = 1.5, font.main = 2, las = 1)
contour(kde.ent18.5km, add = TRUE)
plot(kde.ent19.5km, main = "2019", cex.main = 1.5, font.main = 2, las = 1)
contour(kde.ent19.5km, add = TRUE)
plot(kde.ent20.5km, main = "2020", cex.main = 1.5, font.main = 2, las = 1)
contour(kde.ent20.5km, add = TRUE)
plot(kde.ent21.5km, main = "2021", cex.main = 1.5, font.main = 2, las = 1)
contour(kde.ent21.5km, add = TRUE)
plot(kde.ent22.5km, main = "2022", cex.main = 1.5, font.main = 2, las = 1)
contour(kde.ent22.5km, add = TRUE)
plot(kde.ent.5km, main = "2016-2022", cex.main = 1.5, font.main = 2, las = 1)
contour(kde.ent.5km, add = TRUE)

mtext("Kernel Density Estimate (KDE) Heatmaps of Enteric Fever Cases in Kelantan 2016-2022", side = 1, line = -1, cex = 2, outer = TRUE)
```

# point pattern analysis for Leptospirosis

prepare data for leptospirosis

```{r}
plot(kel_map.owin)
points(lepto_kel.ppp)
```


extract leptospirosis by year, convert to ppp

```{r}
loc_lep16 <- subset(lepto_kel, tahun_daftar=="2016")
loc_lep17 <- subset(lepto_kel, tahun_daftar=="2017")
loc_lep18 <- subset(lepto_kel, tahun_daftar=="2018")
loc_lep19 <- subset(lepto_kel, tahun_daftar=="2019")
loc_lep20 <- subset(lepto_kel, tahun_daftar=="2020")
loc_lep21 <- subset(lepto_kel, tahun_daftar=="2021")
loc_lep22 <- subset(lepto_kel, tahun_daftar=="2022")

```


convert point to spatial

```{r}
loc_lep16.sp <- as(loc_lep16, "Spatial")
```

```{r}
loc_lep17.sp <- as(loc_lep17, "Spatial")
```

```{r}
loc_lep18.sp <- as(loc_lep18, "Spatial")
```

```{r}
loc_lep19.sp <- as(loc_lep19, "Spatial")
```

```{r}
loc_lep20.sp <- as(loc_lep20, "Spatial")
```

```{r}
loc_lep21.sp <- as(loc_lep21, "Spatial")
```

```{r}
loc_lep22.sp <- as(loc_lep22, "Spatial")
```


convert point spatial data to ppp format

```{r}
loc_lep16.ppp <- as(loc_lep16.sp, 'ppp')
loc_lep17.ppp <- as(loc_lep17.sp, 'ppp')
loc_lep18.ppp <- as(loc_lep18.sp, 'ppp')
loc_lep19.ppp <- as(loc_lep19.sp, 'ppp')
loc_lep20.ppp <- as(loc_lep20.sp, 'ppp')
loc_lep21.ppp <- as(loc_lep21.sp, 'ppp')
loc_lep22.ppp <- as(loc_lep22.sp, 'ppp')
```


remove marks

```{r}
#remove marks leptospirosis 2016-2022
lepto_kel.ppp2 <- lepto_kel.ppp
marks(lepto_kel.ppp2) <- NULL

#remove marks (year)
loc_lep16.ppp2 <- loc_lep16.ppp
marks(loc_lep16.ppp2) <- NULL
loc_lep17.ppp2 <- loc_lep17.ppp
marks(loc_lep17.ppp2) <- NULL
loc_lep18.ppp2 <- loc_lep18.ppp
marks(loc_lep18.ppp2) <- NULL
loc_lep19.ppp2 <- loc_lep19.ppp
marks(loc_lep19.ppp2) <- NULL
loc_lep20.ppp2 <- loc_lep20.ppp
marks(loc_lep20.ppp2) <- NULL
loc_lep21.ppp2 <- loc_lep21.ppp
marks(loc_lep21.ppp2) <- NULL
loc_lep22.ppp2 <- loc_lep22.ppp
marks(loc_lep22.ppp2) <- NULL
```


generate window

```{r}
Window(lepto_kel.ppp2) <- kel_map.owin
Window(loc_lep16.ppp2) <- kel_map.owin
Window(loc_lep17.ppp2) <- kel_map.owin
Window(loc_lep18.ppp2) <- kel_map.owin
Window(loc_lep19.ppp2) <- kel_map.owin
Window(loc_lep20.ppp2) <- kel_map.owin
Window(loc_lep21.ppp2) <- kel_map.owin
Window(loc_lep22.ppp2) <- kel_map.owin
```


plot leptospirosis cases

```{r name-of-chunk15, fig.width=20, fig.height=11}
par(mfrow=c(2,4))
plot(loc_lep16.ppp2, main = 2016, cols=rgb(0,0,0,.2), pch=20) #2016
plot(loc_lep17.ppp2, main = 2017, cols=rgb(0,0,0,.2), pch=20) #2017
plot(loc_lep18.ppp2, main = 2018, cols=rgb(0,0,0,.2), pch=20) #2018
plot(loc_lep19.ppp2, main = 2019, cols=rgb(0,0,0,.2), pch=20) #2019
plot(loc_lep20.ppp2, main = 2020, cols=rgb(0,0,0,.2), pch=20) #2020
plot(loc_lep21.ppp2, main = 2021, cols=rgb(0,0,0,.2), pch=20) #2021
plot(loc_lep22.ppp2, main = 2022, cols=rgb(0,0,0,.2), pch=20) #2022
plot(lepto_kel.ppp2, main = "2016-2022", cols=rgb(0,0,0,.2), pch=20) #2016-2022
```


# density analysis for leptospirosis

calculate quadrat count 

```{r}
#| cache = TRUE
quadr_count_lep <- quadratcount(lepto_kel.ppp2, nx = 10,ny = 14) #2016-2022
quadr_count_lep16 <- quadratcount(loc_lep16.ppp2, nx = 10,ny = 14) #2016
quadr_count_lep17 <- quadratcount(loc_lep17.ppp2, nx = 10,ny = 14) #2017
quadr_count_lep18 <- quadratcount(loc_lep18.ppp2, nx = 10,ny = 14) #2018
quadr_count_lep19 <- quadratcount(loc_lep19.ppp2, nx = 10,ny = 14) #2019
quadr_count_lep20 <- quadratcount(loc_lep20.ppp2, nx = 10,ny = 14) #2020
quadr_count_lep21 <- quadratcount(loc_lep21.ppp2, nx = 10,ny = 14) #2021
quadr_count_lep22 <- quadratcount(loc_lep22.ppp2, nx = 10,ny = 14) #2022
```


plot quadrat count

```{r name-of-chunk16, fig.width=28, fig.height=16}
#| cache = TRUE
par( mfrow= c(2,4) )
#2016
plot(loc_lep16.ppp2, pch = 20, cex = 0.1, main = "2016")
plot(quadr_count_lep16, add = TRUE, cex = 2)
#2017
plot(loc_lep17.ppp2, pch = 20, cex = 0.1, main = "2017")
plot(quadr_count_lep17, add = TRUE, cex = 2)
#2018
plot(loc_lep18.ppp2, pch = 20, cex = 0.1, main = "2018")
plot(quadr_count_lep18, add = TRUE, cex = 2)
#2019
plot(loc_lep19.ppp2, pch = 20, cex = 0.1, main = "2019")
plot(quadr_count_lep19, add = TRUE, cex = 2)
#2020
plot(loc_lep20.ppp2, pch = 20, cex = 0.1, main = "2020")
plot(quadr_count_lep20, add = TRUE, cex = 2)
#2021
plot(loc_lep21.ppp2, pch = 20, cex = 0.1, main = "2021")
plot(quadr_count_lep21, add = TRUE, cex = 2)
#2022
plot(loc_lep22.ppp2, pch = 20, cex = 0.1, main = "2022")
plot(quadr_count_lep22, add = TRUE, cex = 2)
#2016-2022
plot(lepto_kel.ppp2, pch = 20, cex = 0.1, main = "2016-2022")
plot(quadr_count_lep, add = TRUE, cex = 2) 

mtext("Point Density Maps of Leptospirosis Cases in Kelantan 2016-2022", side = 1, line = -1, cex = 2, outer = TRUE)
```


calculate VMR (Variance-to-Mean Ratio) Leptospirosis
- VMR greater than 1 indicates spatial clustering.
- leptospirosis not uniformly distributed and may be clustered in certain areas.

```{r}
#| cache = TRUE
#2016-2022
Qcountlep<-data.frame(quadr_count_lep)
var(Qcountlep$Freq)/mean(Qcountlep$Freq)

#2016
Qcountlep16<-data.frame(quadr_count_lep16)
var(Qcountlep16$Freq)/mean(Qcountlep16$Freq)

#2017
Qcountlep17<-data.frame(quadr_count_lep17)
var(Qcountlep17$Freq)/mean(Qcountlep17$Freq)

#2018
Qcountlep18<-data.frame(quadr_count_lep18)
var(Qcountlep18$Freq)/mean(Qcountlep18$Freq)

#2019
Qcountlep19<-data.frame(quadr_count_lep19)
var(Qcountlep19$Freq)/mean(Qcountlep19$Freq)

#2020
Qcountlep20<-data.frame(quadr_count_lep20)
var(Qcountlep20$Freq)/mean(Qcountlep20$Freq)

#2021
Qcountlep21<-data.frame(quadr_count_lep21)
var(Qcountlep21$Freq)/mean(Qcountlep21$Freq)

#2022
Qcountlep22<-data.frame(quadr_count_lep22)
var(Qcountlep22$Freq)/mean(Qcountlep22$Freq)
```

calculate Chi-square test for spatial randomnes - leptospirosis

- a high VMR score and a significant chi-square test result would indicate that the points are clustered and non-randomly distributed, suggesting the presence of spatial processes such as spatial contagion, spatial dependence, or spatial interaction.

```{r}
chi_lep <- quadrat.test(lepto_kel.ppp2, nx= 30, ny=15)
chi_lep16 <- quadrat.test(loc_lep16.ppp2, nx= 30, ny=15)
chi_lep17 <- quadrat.test(loc_lep17.ppp2, nx= 30, ny=15)
chi_lep18 <- quadrat.test(loc_lep18.ppp2, nx= 30, ny=15)
chi_lep19 <- quadrat.test(loc_lep19.ppp2, nx= 30, ny=15)
chi_lep20 <- quadrat.test(loc_lep20.ppp2, nx= 30, ny=15)
chi_lep21 <- quadrat.test(loc_lep21.ppp2, nx= 30, ny=15)
chi_lep22 <- quadrat.test(loc_lep22.ppp2, nx= 30, ny=15)
```

result chi square for enteric fever

```{r}
chi_lep.df <- data.frame(Dataset = "lepto_kel.ppp2",
                       TestStatistic = chi_lep$statistic,
                       PValue = chi_lep$p.value)
chi_lep16.df <- data.frame(Dataset = "loc_lep16.ppp2",
                       TestStatistic = chi_lep16$statistic,
                       PValue = chi_lep16$p.value)
chi_lep17.df <- data.frame(Dataset = "loc_lep17.ppp2",
                       TestStatistic = chi_lep17$statistic,
                       PValue = chi_lep17$p.value)
chi_lep18.df <- data.frame(Dataset = "loc_lep18.ppp2",
                       TestStatistic = chi_lep18$statistic,
                       PValue = chi_lep18$p.value)
chi_lep19.df <- data.frame(Dataset = "loc_lep19.ppp2",
                       TestStatistic = chi_lep19$statistic,
                       PValue = chi_lep19$p.value)
chi_lep20.df <- data.frame(Dataset = "loc_lep20.ppp2",
                       TestStatistic = chi_lep20$statistic,
                       PValue = chi_lep20$p.value)
chi_lep21.df <- data.frame(Dataset = "loc_lep21.ppp2",
                       TestStatistic = chi_lep21$statistic,
                       PValue = chi_lep21$p.value)
chi_lep22.df <- data.frame(Dataset = "loc_lep22.ppp2",
                       TestStatistic = chi_lep22$statistic,
                       PValue = chi_lep22$p.value)
chi_quadLEP <- bind_rows(
  data.frame(Dataset = "2016", chi_lep16.df), 
  data.frame(Dataset = "2017", chi_lep17.df), 
  data.frame(Dataset = "2018", chi_lep18.df), 
  data.frame(Dataset = "2019", chi_lep19.df), 
  data.frame(Dataset = "2020", chi_lep20.df), 
  data.frame(Dataset = "2021", chi_lep21.df), 
  data.frame(Dataset = "2022", chi_lep22.df), 
  data.frame(Dataset = "2016-2022", chi_lep.df))
chi_quadLEP

```


Monte Carlo permutation for spatial randomnes - leptospirosis (chi square assumption not met)

```{r}
# calculate monte carlo
#| cache = TRUE
mc_qlep <- quadrat.test(lepto_kel.ppp2, nx= 30, ny=15, method = "MonteCarlo")
mc_qlep16 <- quadrat.test(loc_lep16.ppp2, nx= 30, ny=15, method = "MonteCarlo")
mc_qlep17 <- quadrat.test(loc_lep17.ppp2, nx= 30, ny=15, method = "MonteCarlo")
mc_qlep18 <- quadrat.test(loc_lep18.ppp2, nx= 30, ny=15, method = "MonteCarlo")
mc_qlep19 <- quadrat.test(loc_lep19.ppp2, nx= 30, ny=15, method = "MonteCarlo")
mc_qlep20 <- quadrat.test(loc_lep20.ppp2, nx= 30, ny=15, method = "MonteCarlo")
mc_qlep21 <- quadrat.test(loc_lep21.ppp2, nx= 30, ny=15, method = "MonteCarlo")
mc_qlep22 <- quadrat.test(loc_lep22.ppp2, nx= 30, ny=15, method = "MonteCarlo")
```


convert to data frame

```{r}
mc_qlep.df <- data.frame(Dataset = "lepto_kel.ppp2",
                       TestStatistic = mc_qlep$statistic,
                       PValue = mc_qlep$p.value)
mc_qlep16.df <- data.frame(Dataset = "loc_lep16.ppp2",
                       TestStatistic = mc_qlep16$statistic,
                       PValue = mc_qlep16$p.value)
mc_qlep17.df <- data.frame(Dataset = "loc_lep17.ppp2",
                       TestStatistic = mc_qlep17$statistic,
                       PValue = mc_qlep17$p.value)
mc_qlep18.df <- data.frame(Dataset = "loc_lep18.ppp2",
                       TestStatistic = mc_qlep18$statistic,
                       PValue = mc_qlep18$p.value)
mc_qlep19.df <- data.frame(Dataset = "loc_lep19.ppp2",
                       TestStatistic = mc_qlep19$statistic,
                       PValue = mc_qlep19$p.value)
mc_qlep20.df <- data.frame(Dataset = "loc_lep20.ppp2",
                       TestStatistic = mc_qlep20$statistic,
                       PValue = mc_qlep20$p.value)
mc_qlep21.df <- data.frame(Dataset = "loc_lep21.ppp2",
                       TestStatistic = mc_qlep21$statistic,
                       PValue = mc_qlep21$p.value)
mc_qlep22.df <- data.frame(Dataset = "loc_lep22.ppp2",
                       TestStatistic = mc_qlep22$statistic,
                       PValue = mc_qlep22$p.value)
```


Display quadrat test with Monte Carlo permutation

```{r}
mc_qLEP <- bind_rows(
  data.frame(Dataset = "2016", mc_qlep16.df), 
  data.frame(Dataset = "2017", mc_qlep17.df), 
  data.frame(Dataset = "2018", mc_qlep18.df), 
  data.frame(Dataset = "2019", mc_qlep19.df), 
  data.frame(Dataset = "2020", mc_qlep20.df), 
  data.frame(Dataset = "2021", mc_qlep21.df), 
  data.frame(Dataset = "2022", mc_qlep22.df), 
  data.frame(Dataset = "2016-2022", mc_qlep.df))
mc_qLEP
```


# intensity analysis for leptospirosis

Plot intensity

```{r name-of-chunk17, fig.width=16, fig.height=8}
#| cache = TRUE
par( mfrow= c(2,4) )
inten_lep16 <-intensity(quadr_count_lep16)
plot(intensity(quadr_count_lep16, image = TRUE), main = 2016, las = 1)
plot(loc_lep16, pch = 20, cex = 0.6, add = TRUE)

inten_lep17 <-intensity(quadr_count_lep17)
plot(intensity(quadr_count_lep17, image = TRUE), main = 2017, las = 1)
plot(loc_lep17, pch = 20, cex = 0.6, add = TRUE)

inten_lep18 <-intensity(quadr_count_lep18)
plot(intensity(quadr_count_lep18, image = TRUE), main = 2018, las = 1)
plot(loc_lep18, pch = 20, cex = 0.6, add = TRUE)

inten_lep19 <-intensity(quadr_count_lep19)
plot(intensity(quadr_count_lep19, image = TRUE), main = 2019, las = 1)
plot(loc_lep19, pch = 20, cex = 0.6, add = TRUE)

inten_lep20 <-intensity(quadr_count_lep20)
plot(intensity(quadr_count_lep20, image = TRUE), main = 2020, las = 1)
plot(loc_lep20, pch = 20, cex = 0.6, add = TRUE)

inten_lep21 <-intensity(quadr_count_lep21)
plot(intensity(quadr_count_lep21, image = TRUE), main = 2021, las = 1)
plot(loc_lep21, pch = 20, cex = 0.6, add = TRUE)

inten_lep22 <-intensity(quadr_count_lep22)
plot(intensity(quadr_count_lep22, image = TRUE), main = 2022, las = 1)
plot(loc_lep22, pch = 20, cex = 0.6, add = TRUE)

inten_lep <-intensity(quadr_count_lep)
plot(intensity(quadr_count_lep, image = TRUE), main = "2016-2022" , las = 1)
plot(lepto_kel.ppp2, pch = 20, cex = 0.6, add = TRUE)

mtext("Intensity Maps of Leptospirosis Cases in Kelantan 2016-2022", side = 1, line = -1, cex = 1.5, outer = TRUE)
```

# Kernel density raster for leptospirosis

rescale to km

```{r}
#point
lepto_kel.ppp2.km <- rescale(lepto_kel.ppp2, 1000, 'km') #cumulative
loc_lep16.ppp2.km <- rescale(loc_lep16.ppp2, 1000, 'km') #2016
loc_lep17.ppp2.km <- rescale(loc_lep17.ppp2, 1000, 'km') #2017
loc_lep18.ppp2.km <- rescale(loc_lep18.ppp2, 1000, 'km') #2018
loc_lep19.ppp2.km <- rescale(loc_lep19.ppp2, 1000, 'km') #2019
loc_lep20.ppp2.km <- rescale(loc_lep20.ppp2, 1000, 'km') #2020
loc_lep21.ppp2.km <- rescale(loc_lep21.ppp2, 1000, 'km') #2021
loc_lep22.ppp2.km <- rescale(loc_lep22.ppp2, 1000, 'km') #2022

#map
kel_map.owin.km <- rescale(kel_map.owin, 1000, 'km') 
```


calculate kernel density raster (kde) for Leptospirosis 2016 - 2022

```{r}
#| cache = TRUE
kde.lep <- density(lepto_kel.ppp2.km) #2016 - 2022
kde.lep16 <- density(loc_lep16.ppp2.km) #2016
kde.lep17 <- density(loc_lep17.ppp2.km) #2017
kde.lep18 <- density(loc_lep18.ppp2.km) #2018
kde.lep19 <- density(loc_lep19.ppp2.km) #2019
kde.lep20 <- density(loc_lep20.ppp2.km) #2020
kde.lep21 <- density(loc_lep21.ppp2.km) #2021
kde.lep22 <- density(loc_lep22.ppp2.km) #2022
```


plot KDE leptospirosis

```{r name-of-chunk18, fig.width=20, fig.height=11}
#| cache = TRUE
par(mfrow=c(2,4))
plot(kde.lep16, main = 2016, las = 1) 
plot(kde.lep17, main = 2017, las = 1) 
plot(kde.lep18, main = 2018, las = 1) 
plot(kde.lep19, main = 2019, las = 1) 
plot(kde.lep20, main = 2020, las = 1) 
plot(kde.lep21, main = 2021, las = 1) 
plot(kde.lep22, main = 2022, las = 1) 
plot(kde.lep, main = "2016-2022", las = 1)
```

plot kde 20km bandwith leptospirosis

```{r name-of-chunk19, fig.width=20, fig.height=11}
#| cache = TRUE
kde.lep.20km <- density(lepto_kel.ppp2.km, sigma = 20) #2016 - 2022
kde.lep16.20km <- density(loc_lep16.ppp2.km, sigma = 20) #2016
kde.lep17.20km <- density(loc_lep17.ppp2.km, sigma = 20) #2017
kde.lep18.20km <- density(loc_lep18.ppp2.km, sigma = 20) #2018
kde.lep19.20km <- density(loc_lep19.ppp2.km, sigma = 20) #2019
kde.lep20.20km <- density(loc_lep20.ppp2.km, sigma = 20) #2020
kde.lep21.20km <- density(loc_lep21.ppp2.km, sigma = 20) #2021
kde.lep22.20km <- density(loc_lep22.ppp2.km, sigma = 20) #2022

#plot KDE leptospirosis 20km bandwith
par(mfrow=c(2,4))
#2016
plot(kde.lep16.20km, main = 2016, las = 1)
contour(kde.lep16.20km, add = TRUE)
#2017
plot(kde.lep17.20km, main = 2017, las = 1)
contour(kde.lep17.20km, add = TRUE)
#2018
plot(kde.lep18.20km, main = 2018, las = 1)
contour(kde.lep18.20km, add = TRUE)
#2019
plot(kde.lep19.20km, main = 2019, las = 1)
contour(kde.lep19.20km, add = TRUE)
#2020
plot(kde.lep20.20km, main = 2020, las = 1)
contour(kde.lep20.20km, add = TRUE)
#2021
plot(kde.lep21.20km, main = 2021, las = 1)
contour(kde.lep21.20km, add = TRUE)
#2022
plot(kde.lep22.20km, main = 2022, las = 1)
contour(kde.lep22.20km, add = TRUE)
#2016-2022
plot(kde.lep.20km, main = "2016-2022", las = 1)
contour(kde.lep.20km, add = TRUE)
```

kde 5km bandwith leptospirosis

```{r}
#| cache = TRUE
kde.lep.5km <- density(lepto_kel.ppp2.km, sigma = 5) #2016 - 2022
kde.lep16.5km <- density(loc_lep16.ppp2.km, sigma = 5) #2016
kde.lep17.5km <- density(loc_lep17.ppp2.km, sigma = 5) #2017
kde.lep18.5km <- density(loc_lep18.ppp2.km, sigma = 5) #2018
kde.lep19.5km <- density(loc_lep19.ppp2.km, sigma = 5) #2019
kde.lep20.5km <- density(loc_lep20.ppp2.km, sigma = 5) #2020
kde.lep21.5km <- density(loc_lep21.ppp2.km, sigma = 5) #2021
kde.lep22.5km <- density(loc_lep22.ppp2.km, sigma = 5) #2022
```


plot kde 5km bandwith leptospirosis

```{r name-of-chunk20, fig.width=20, fig.height=11}
par( mfrow= c(2,4) )
plot(kde.lep16.5km, main = "2016", cex.main = 1.5, font.main = 2, las = 1)
contour(kde.lep16.5km, add = TRUE)
plot(kde.lep17.5km, main = "2017", cex.main = 1.5, font.main = 2, las = 1)
contour(kde.lep17.5km, add = TRUE)
plot(kde.lep18.5km, main = "2018", cex.main = 1.5, font.main = 2, las = 1)
contour(kde.lep18.5km, add = TRUE)
plot(kde.lep19.5km, main = "2019", cex.main = 1.5, font.main = 2, las = 1)
contour(kde.lep19.5km, add = TRUE)
plot(kde.lep20.5km, main = "2020", cex.main = 1.5, font.main = 2, las = 1)
contour(kde.lep20.5km, add = TRUE)
plot(kde.lep21.5km, main = "2021", cex.main = 1.5, font.main = 2, las = 1)
contour(kde.lep21.5km, add = TRUE)
plot(kde.lep22.5km, main = "2022", cex.main = 1.5, font.main = 2, las = 1)
contour(kde.lep22.5km, add = TRUE)
plot(kde.lep.5km, main = "2016-2022", cex.main = 1.5, font.main = 2, las = 1)
contour(kde.lep.5km, add = TRUE)

#add title to the plot
mtext("Kernel Density Estimate (KDE) Heatmaps of Leptospirosis Cases in Kelantan 2016-2022", side = 1, line = -1, cex = 2, outer = TRUE)
```


# distance-based analysis for Enteric fever

ANN and intensity for Enteric fever 2016-2022

```{r}
# extracts the intensity - average number of enteric fever per square kilometer)
ENTIntensity <- round(summary(Enteric_kel.ppp2.km)$intensity, 3) 

# calculates the average nearest neighbor distance  in km
ENTANNkm <- round(mean(nndist(Enteric_kel.ppp2.km)), 2) 

# calculates the average nearest neighbor distance  in m
ENTANN <- round(mean(nndist(Enteric_kel.ppp2)), 2) 
```


Intensity by year

```{r}
ENT16_Intensity <- round(summary(loc_ent16.ppp2.km)$intensity, 3) #2016
ENT17_Intensity <- round(summary(loc_ent17.ppp2.km)$intensity, 3) #2017
ENT18_Intensity <- round(summary(loc_ent18.ppp2.km)$intensity, 3) #2018
ENT19_Intensity <- round(summary(loc_ent19.ppp2.km)$intensity, 3) #2019
ENT20_Intensity <- round(summary(loc_ent20.ppp2.km)$intensity, 3) #2020
ENT21_Intensity <- round(summary(loc_ent21.ppp2.km)$intensity, 3) #2021
ENT22_Intensity <- round(summary(loc_ent22.ppp2.km)$intensity, 3) #2022
```


average ANN in meter by year

```{r}
ENTANN16 <- round(mean(nndist(loc_ent16.ppp2)), 2)
ENTANN17 <- round(mean(nndist(loc_ent17.ppp2)), 2)
ENTANN18 <- round(mean(nndist(loc_ent18.ppp2)), 2)
ENTANN19 <- round(mean(nndist(loc_ent19.ppp2)), 2)
ENTANN20 <- round(mean(nndist(loc_ent20.ppp2)), 2)
ENTANN21 <- round(mean(nndist(loc_ent21.ppp2)), 2)
ENTANN22 <- round(mean(nndist(loc_ent22.ppp2)), 2)
```


interpret spatial intensity

```{r}
print(paste0("The average spatial intensity for Enteric Fever cases 2016-2022 is ", ENTIntensity, " cases per sq. km."))
print(paste0("The average spatial intensity for Enteric Fever cases 2016 is ", ENT16_Intensity, " cases per sq. km."))
print(paste0("The average spatial intensity for Enteric Fever cases 2017 is ", ENT17_Intensity, " cases per sq. km."))
print(paste0("The average spatial intensity for Enteric Fever cases 2018 is ", ENT18_Intensity, " cases per sq. km."))
print(paste0("The average spatial intensity for Enteric Fever cases 2019 is ", ENT19_Intensity, " cases per sq. km."))
print(paste0("The average spatial intensity for Enteric Fever cases 2020 is ", ENT20_Intensity, " cases per sq. km."))
print(paste0("The average spatial intensity for Enteric Fever cases 2021 is ", ENT21_Intensity, " cases per sq. km."))
print(paste0("The average spatial intensity for Enteric Fever cases 2022 is ", ENT22_Intensity, " cases per sq. km."))
```



Calculate nearest neigbour index (NNI) for enteric fever
NNI < 1 clustered, >1 dispersed
```{r}
#interpret ANN
print(paste0("The mean nearest neighbor distance for Enteric Fever cases 2016-2022 is ", ENTANN, " m.")) 
print(paste0("The mean nearest neighbor distance for Enteric Fever cases 2016 is ", ENTANN16, " m.")) 
print(paste0("The mean nearest neighbor distance for Enteric Fever cases 2016 is ", ENTANN17, " m.")) 
print(paste0("The mean nearest neighbor distance for Enteric Fever cases 2016 is ", ENTANN18, " m.")) 
print(paste0("The mean nearest neighbor distance for Enteric Fever cases 2016 is ", ENTANN19, " m.")) 
print(paste0("The mean nearest neighbor distance for Enteric Fever cases 2016 is ", ENTANN20, " m.")) 
print(paste0("The mean nearest neighbor distance for Enteric Fever cases 2016 is ", ENTANN21, " m.")) 
print(paste0("The mean nearest neighbor distance for Enteric Fever cases 2016 is ", ENTANN22, " m.")) 
```

```{r}
#| cache = TRUE
nniENT <- nni(Enteric_kel)
nniENT16 <- nni(loc_ent16)
nniENT17 <-nni(loc_ent17)
nniENT18 <- nni(loc_ent18)
nniENT19 <-nni(loc_ent19)
nniENT20 <-nni(loc_ent20)
nniENT21 <-nni(loc_ent21)
nniENT22 <-nni(loc_ent22)

NNI_ENT <- bind_rows(
  data.frame(Dataset = "2016-2022", nniENT),
  data.frame(Dataset = "2016", nniENT16),
  data.frame(Dataset = "2017", nniENT17),
  data.frame(Dataset = "2018", nniENT18),
  data.frame(Dataset = "2019", nniENT19),
  data.frame(Dataset = "2020", nniENT20),
  data.frame(Dataset = "2021", nniENT21),
  data.frame(Dataset = "2022", nniENT22)
)
NNI_ENT
```


clark evans test for Enteric Fever
-R less than 1 indicates spatial clustering. The smaller the value of R, the greater the degree of clustering.

```{r}
#| cache = TRUE
CE_ent <- clarkevans.test(Enteric_kel.ppp2)
CE_ent16 <- clarkevans.test(loc_ent16.ppp2)
CE_ent17 <- clarkevans.test(loc_ent17.ppp2)
CE_ent18 <- clarkevans.test(loc_ent18.ppp2)
CE_ent19 <- clarkevans.test(loc_ent19.ppp2)
CE_ent20 <- clarkevans.test(loc_ent20.ppp2)
CE_ent21 <- clarkevans.test(loc_ent21.ppp2)
CE_ent22 <- clarkevans.test(loc_ent22.ppp2)

# dataframe
CE_ent.df <- data.frame(Dataset = "Enteric_kel.ppp2",
                       TestStatistic = CE_ent$statistic,
                       PValue = CE_ent$p.value)
CE_ent16.df <- data.frame(Dataset = "loc_ent16.ppp2",
                       TestStatistic = CE_ent16$statistic,
                       PValue = CE_ent16$p.value)
CE_ent17.df <- data.frame(Dataset = "loc_ent17.ppp2",
                       TestStatistic = CE_ent17$statistic,
                       PValue = CE_ent17$p.value)
CE_ent18.df <- data.frame(Dataset = "loc_ent18.ppp2",
                       TestStatistic = CE_ent18$statistic,
                       PValue = CE_ent18$p.value)
CE_ent19.df <- data.frame(Dataset = "loc_ent19.ppp2",
                       TestStatistic = CE_ent19$statistic,
                       PValue = CE_ent19$p.value)
CE_ent20.df <- data.frame(Dataset = "loc_ent20.ppp2",
                       TestStatistic = CE_ent20$statistic,
                       PValue = CE_ent20$p.value)
CE_ent21.df <- data.frame(Dataset = "loc_ent21.ppp2",
                       TestStatistic = CE_ent21$statistic,
                       PValue = CE_ent21$p.value)
CE_ent22.df <- data.frame(Dataset = "loc_ent22.ppp2",
                       TestStatistic = CE_ent22$statistic,
                       PValue = CE_ent22$p.value)

#combine rows
CE_ENT <- bind_rows(
  data.frame(Dataset = "2016-2022", CE_ent.df),
  data.frame(Dataset = "2016", CE_ent16.df),
  data.frame(Dataset = "2017", CE_ent17.df),
  data.frame(Dataset = "2018", CE_ent18.df),
  data.frame(Dataset = "2019", CE_ent19.df),
  data.frame(Dataset = "2020", CE_ent20.df),
  data.frame(Dataset = "2021", CE_ent21.df),
  data.frame(Dataset = "2022", CE_ent22.df)
)
CE_ENT
```


mean centre and standard distance for Enteric fever

```{r}
#calculate mean center (m) for enteric fever
entxy <- coords(Enteric_kel.ppp2)
mcent <- summarize(entxy, xmean = mean(x), ymean = mean(y))
mcent

#calculate standard distance (m) for enteric fever
sdent <- sqrt(sum((entxy[,1] - mcent[1])^2 + (entxy[,2] - mcent[2])^2) / nrow(entxy))
sdent
```


Median Absolute Deviation test with 99 simulations of Complete Spatial Randomness (CSR) for K function - enteric fever

```{r}
#| cache = TRUE
ENTMADTest <- mad.test(Enteric_kel.ppp2.km, fun = Kest, nsim = 99, verbose = FALSE) # 
```


```{r}
#clustering for Enteric fever - hypthesis testing
if(ENTMADTest$p.value < 0.05){ # if the MAD test has a p value less than 0.05, this prints a statement indicating statistical significance
  print(paste0("These Enteric Fever cases in 2016-2022 are significantly more clustered than expected by Complete Spatial Randomness (CSR) - Median absolute deviation p value = ", round(ENTMADTest$p.value, 3), "."))
} else { # if the p value is greater than or equal to 0.05, this prints a statement indicating non-significance
  print(paste0("These Enteric Fever cases are not significantly more clustered than expected by Complete Spatial Randomness (CSR) - Median absolute deviation p value = ", round(sightMADTest$p.value, 3), "."))
}
```

```{r}
#| cache = TRUE
ENTMADTest16 <- mad.test(loc_ent16.ppp2.km, fun = Kest, nsim = 99, verbose = FALSE)
ENTMADTest17 <- mad.test(loc_ent17.ppp2.km, fun = Kest, nsim = 99, verbose = FALSE)
ENTMADTest18 <- mad.test(loc_ent18.ppp2.km, fun = Kest, nsim = 99, verbose = FALSE)
ENTMADTest19 <- mad.test(loc_ent19.ppp2.km, fun = Kest, nsim = 99, verbose = FALSE)
ENTMADTest20 <- mad.test(loc_ent20.ppp2.km, fun = Kest, nsim = 99, verbose = FALSE)
ENTMADTest21 <- mad.test(loc_ent21.ppp2.km, fun = Kest, nsim = 99, verbose = FALSE)
ENTMADTest22 <- mad.test(loc_ent22.ppp2.km, fun = Kest, nsim = 99, verbose = FALSE)
```


result MAD test Enteric Fever for K function

```{r}
MAD_ent.df <- data.frame(Dataset = "Enteric_kel.ppp2",
                       TestStatistic = ENTMADTest$statistic,
                       PValue = ENTMADTest16$p.value)
MAD_ent16.df <- data.frame(Dataset = "loc_ent16.ppp2",
                       TestStatistic = ENTMADTest16$statistic,
                       PValue = ENTMADTest16$p.value)
MAD_ent17.df <- data.frame(Dataset = "loc_ent17.ppp2",
                       TestStatistic = ENTMADTest17$statistic,
                       PValue = ENTMADTest17$p.value)
MAD_ent18.df <- data.frame(Dataset = "loc_ent18.ppp2",
                       TestStatistic = ENTMADTest18$statistic,
                       PValue = ENTMADTest18$p.value)
MAD_ent19.df <- data.frame(Dataset = "loc_ent19.ppp2",
                       TestStatistic = ENTMADTest19$statistic,
                       PValue = ENTMADTest19$p.value)
MAD_ent20.df <- data.frame(Dataset = "loc_ent20.ppp2",
                       TestStatistic = ENTMADTest20$statistic,
                       PValue = ENTMADTest20$p.value)
MAD_ent21.df <- data.frame(Dataset = "loc_ent21.ppp2",
                       TestStatistic = ENTMADTest21$statistic,
                       PValue = ENTMADTest21$p.value)
MAD_ent22.df <- data.frame(Dataset = "loc_ent22.ppp2",
                       TestStatistic = ENTMADTest22$statistic,
                       PValue = ENTMADTest22$p.value)

#combine rows
MAD_ENT <- bind_rows(
  data.frame(Dataset = "2016-2022", MAD_ent.df),
  data.frame(Dataset = "2016", MAD_ent16.df),
  data.frame(Dataset = "2017", MAD_ent17.df),
  data.frame(Dataset = "2018", MAD_ent18.df),
  data.frame(Dataset = "2019", MAD_ent19.df),
  data.frame(Dataset = "2020", MAD_ent20.df),
  data.frame(Dataset = "2021", MAD_ent21.df),
  data.frame(Dataset = "2022", MAD_ent22.df)
)
MAD_ENT
```

K function enteric fever

```{r}
#| cache = TRUE
Kest_ent <- Kest(Enteric_kel.ppp2.km)
Kest_ent16 <- Kest(loc_ent16.ppp2.km)
Kest_ent17 <- Kest(loc_ent17.ppp2.km)
Kest_ent18 <- Kest(loc_ent18.ppp2.km)
Kest_ent19 <- Kest(loc_ent19.ppp2.km)
Kest_ent20 <- Kest(loc_ent20.ppp2.km)
Kest_ent21 <- Kest(loc_ent21.ppp2.km)
Kest_ent22 <- Kest(loc_ent22.ppp2.km)
```


Plot K test
- Empirical values greater than theoretical values suggest clustering.
- plotting envelope takes time, significance already tested using MAD test + MC permutation

```{r name-of-chunk21, fig.width=25, fig.height=8}
par( mfrow= c(2,4) )
plot(Kest_ent16, main = "2016", cex.main = 3, cex.axis = 1.5, cex.lab = 1.5)
plot(Kest_ent17, main = "2017", cex.main = 3, cex.axis = 1.5, cex.lab = 1.5)
plot(Kest_ent18, main = "2018", cex.main = 3, cex.axis = 1.5, cex.lab = 1.5)
plot(Kest_ent19, main = "2019", cex.main = 3, cex.axis = 1.5, cex.lab = 1.5)
plot(Kest_ent20, main = "2020", cex.main = 3, cex.axis = 1.5, cex.lab = 1.5)
plot(Kest_ent21, main = "2021", cex.main = 3, cex.axis = 1.5, cex.lab = 1.5)
plot(Kest_ent22, main = "2022", cex.main = 3, cex.axis = 1.5, cex.lab = 1.5)
plot(Kest_ent, main = "2016-2022", cex.main = 3, cex.axis = 1.5, cex.lab = 1.5)
```

G function enteric fever
- empirical function is greater than the theoretical (blue) suggest clustered pattern

```{r}
#| cache = TRUE
gestENT <- Gest(Enteric_kel.ppp2.km)
gestENT16 <- Gest(loc_ent16.ppp2.km)
gestENT17 <- Gest(loc_ent17.ppp2.km)
gestENT18 <- Gest(loc_ent18.ppp2.km)
gestENT19 <- Gest(loc_ent19.ppp2.km)
gestENT20 <- Gest(loc_ent20.ppp2.km)
gestENT21 <- Gest(loc_ent21.ppp2.km)
gestENT22 <- Gest(loc_ent22.ppp2.km)
```

```{r name-of-chunk22, fig.width=25, fig.height=8}
#| cache = TRUE
par( mfrow= c(2,4) )
plot(gestENT16, main = "2016", cex.main = 3, cex.lab = 1.5, cex.axis = 1.5)
plot(gestENT17, main = "2017", cex.main = 3, cex.lab = 1.5, cex.axis = 1.5)
plot(gestENT18, main = "2018", cex.main = 3, cex.lab = 1.5, cex.axis = 1.5)
plot(gestENT19, main = "2019", cex.main = 3, cex.lab = 1.5, cex.axis = 1.5)
plot(gestENT20, main = "2020", cex.main = 3, cex.lab = 1.5, cex.axis = 1.5)
plot(gestENT21, main = "2021", cex.main = 3, cex.lab = 1.5, cex.axis = 1.5)
plot(gestENT22, main = "2022", cex.main = 3, cex.lab = 1.5, cex.axis = 1.5)
plot(gestENT, main = "2016-2022", cex.main = 3, cex.lab = 1.5, cex.axis = 1.5)
```


enveloped G function enteric fever 2016-2022

```{r name-of-chunk23a, fig.width=25, fig.height=8}
#| cache = TRUE
# G function enteric fever per year
par( mfrow= c(2,4) )
GestENT16 <- plot(envelope(loc_ent16.ppp2.km, Gest, nsim = 99, verbose = FALSE), main = "2016",xlab = "Distance (KM)", ylab = "Proportion of Enteric Fever Cases 2016", cex.main = 2, cex.axis = 1, cex.lab = 1)
GestENT17 <- plot(envelope(loc_ent17.ppp2.km, Gest, nsim = 99, verbose = FALSE), main = "2017",xlab = "Distance (KM)", ylab = "Proportion of Enteric Fever Cases 2017", cex.main = 2, cex.axis = 1, cex.lab = 1)
GestENT18 <- plot(envelope(loc_ent18.ppp2.km, Gest, nsim = 99, verbose = FALSE), main = "2018",xlab = "Distance (KM)", ylab = "Proportion of Enteric Fever Cases 2018", cex.main = 2, cex.axis = 1, cex.lab = 1)
GestENT19 <- plot(envelope(loc_ent19.ppp2.km, Gest, nsim = 99, verbose = FALSE), main = "2019",xlab = "Distance (KM)", ylab = "Proportion of Enteric Fever Cases 2019", cex.main = 2, cex.axis = 1, cex.lab = 1)
GestENT20 <- plot(envelope(loc_ent20.ppp2.km, Gest, nsim = 99, verbose = FALSE), main = "2020",xlab = "Distance (KM)", ylab = "Proportion of Enteric Fever Cases 2020", cex.main = 2, cex.axis = 1, cex.lab = 1)
GestENT21 <- plot(envelope(loc_ent21.ppp2.km, Gest, nsim = 99, verbose = FALSE), main = "2021",xlab = "Distance (KM)", ylab = "Proportion of Enteric Fever Cases 2021", cex.main = 2, cex.axis = 1, cex.lab = 1)
GestENT22 <- plot(envelope(loc_ent22.ppp2.km, Gest, nsim = 99, verbose = FALSE), main = "2022",xlab = "Distance (KM)", ylab = "Proportion of Enteric Fever Cases 2022", cex.main = 2, cex.axis = 1, cex.lab = 1)
GestENT <- plot(envelope(Enteric_kel.ppp2.km, Gest, nsim = 99, verbose = FALSE), main = "2016-2022",xlab = "Distance (KM)", ylab = "Proportion of Enteric Fever Cases 2016-2022", cex.main = 2, cex.axis = 1, cex.lab = 1)

mtext("Enteric Fever Pattern in Kelantan 2016-2022", side = 3, line = -1, cex = 1.5, outer = TRUE)
```


F test for enteric fever
- empirical  values below theoretical (blue) suggest clustering

```{r name-of-chunk24, fig.width=28, fig.height=14}
#| cache = TRUE
par( mfrow= c(2,4) )
plot(Fest(loc_ent16.ppp2.km), main = "2016", cex.main = 3, cex.axis = 1.5, cex.lab = 1.5)
plot(Fest(loc_ent17.ppp2.km), main = "2017", cex.main = 3, cex.axis = 1.5, cex.lab = 1.5)
plot(Fest(loc_ent18.ppp2.km), main = "2018", cex.main = 3, cex.axis = 1.5, cex.lab = 1.5)
plot(Fest(loc_ent19.ppp2.km), main = "2019", cex.main = 3, cex.axis = 1.5, cex.lab = 1.5)
plot(Fest(loc_ent20.ppp2.km), main = "2020", cex.main = 3, cex.axis = 1.5, cex.lab = 1.5)
plot(Fest(loc_ent21.ppp2.km), main = "2021", cex.main = 3, cex.axis = 1.5, cex.lab = 1.5)
plot(Fest(loc_ent22.ppp2.km), main = "2022", cex.main = 3, cex.axis = 1.5, cex.lab = 1.5)
plot(Fest(Enteric_kel.ppp2.km), main = "2016-2022", cex.main = 3, cex.axis = 1.5, cex.lab = 1.5)

mtext("Enteric Fever Pattern in Kelantan 2016-2022", side = 1, line = -1, cex = 1.5, outer = TRUE)
```

# distance-based analysis for Leptospirosis

ANN and intensity for Leptospirosis 2016-2022

```{r}
#| cache = TRUE
LEPIntensity <- round(summary(lepto_kel.ppp2.km)$intensity, 3) 
LEPANNkm <- round(mean(nndist(lepto_kel.ppp2.km)), 3) 
LEPANN <- round(mean(nndist(lepto_kel.ppp2)), 3) 

#Intensity by year
LEP16_Intensity <- round(summary(loc_lep16.ppp2.km)$intensity, 3) #2016
LEP17_Intensity <- round(summary(loc_lep17.ppp2.km)$intensity, 3) #2017
LEP18_Intensity <- round(summary(loc_lep18.ppp2.km)$intensity, 3) #2018
LEP19_Intensity <- round(summary(loc_lep19.ppp2.km)$intensity, 3) #2019
LEP20_Intensity <- round(summary(loc_lep20.ppp2.km)$intensity, 3) #2020
LEP21_Intensity <- round(summary(loc_lep21.ppp2.km)$intensity, 3) #2021
LEP22_Intensity <- round(summary(loc_lep22.ppp2.km)$intensity, 3) #2022

#ANN in meter by year
LEPANN16 <- round(mean(nndist(loc_lep16.ppp2)), 2)
LEPANN17 <- round(mean(nndist(loc_lep17.ppp2)), 2)
LEPANN18 <- round(mean(nndist(loc_lep18.ppp2)), 2)
LEPANN19 <- round(mean(nndist(loc_lep19.ppp2)), 2)
LEPANN20 <- round(mean(nndist(loc_lep20.ppp2)), 2)
LEPANN21 <- round(mean(nndist(loc_lep21.ppp2)), 2)
LEPANN22 <- round(mean(nndist(loc_lep22.ppp2)), 2)

#interpret spatial intensity
print(paste0("The average spatial intensity for Leptospirosis cases 2016-2022 is ", LEPIntensity, " cases per sq. km."))
print(paste0("The average spatial intensity for Leptospirosis cases 2016 is ", LEP16_Intensity, " cases per sq. km."))
print(paste0("The average spatial intensity for Leptospirosis cases 2017 is ", LEP17_Intensity, " cases per sq. km."))
print(paste0("The average spatial intensity for Leptospirosis cases 2018 is ", LEP18_Intensity, " cases per sq. km."))
print(paste0("The average spatial intensity for Leptospirosis cases 2019 is ", LEP19_Intensity, " cases per sq. km."))
print(paste0("The average spatial intensity for Leptospirosis cases 2020 is ", LEP20_Intensity, " cases per sq. km."))
print(paste0("The average spatial intensity for Leptospirosis cases 2021 is ", LEP21_Intensity, " cases per sq. km."))
print(paste0("The average spatial intensity for Leptospirosis cases 2022 is ", LEP22_Intensity, " cases per sq. km."))

#interpret ANN
print(paste0("The mean nearest neighbor distance for Leptospirosis cases 2016-2022 is ", LEPANN, " m.")) 
print(paste0("The mean nearest neighbor distance for Leptospirosis cases 2016 is ", LEPANN16, " m.")) 
print(paste0("The mean nearest neighbor distance for Leptospirosis cases 2017 is ", LEPANN17, " m.")) 
print(paste0("The mean nearest neighbor distance for Leptospirosis cases 2018 is ", LEPANN18, " m.")) 
print(paste0("The mean nearest neighbor distance for Leptospirosis cases 2019 is ", LEPANN19, " m.")) 
print(paste0("The mean nearest neighbor distance for Leptospirosis cases 2020 is ", LEPANN20, " m.")) 
print(paste0("The mean nearest neighbor distance for Leptospirosis cases 2021 is ", LEPANN21, " m.")) 
print(paste0("The mean nearest neighbor distance for Leptospirosis cases 2022 is ", LEPANN22, " m.")) 

```


result ANN +intensity for leptospirosis

```{r}
ANN_LEPr <- bind_rows(
  data.frame(Dataset = "2016-2022", ANN.m = LEPANN, Intensity.sq.km = LEPIntensity), 
  data.frame(Dataset = "2016", ANN.m = LEPANN16, Intensity.sq.km = LEP16_Intensity), 
  data.frame(Dataset = "2017", ANN.m = LEPANN17, Intensity.sq.km = LEP17_Intensity),
  data.frame(Dataset = "2018", ANN.m = LEPANN18, Intensity.sq.km = LEP18_Intensity),
  data.frame(Dataset = "2019", ANN.m = LEPANN19, Intensity.sq.km = LEP19_Intensity),
  data.frame(Dataset = "2020", ANN.m = LEPANN20, Intensity.sq.km = LEP20_Intensity),
  data.frame(Dataset = "2021", ANN.m = LEPANN21, Intensity.sq.km = LEP21_Intensity),
  data.frame(Dataset = "2022", ANN.m = LEPANN22, Intensity.sq.km = LEP22_Intensity)
)
ANN_LEPr # ANN+intensity in table
```


Calculate nearest neigbour index (NNI) for leptospirosis
- NNI < 1 clustered, >1 dispersed

```{r}
#| cache = TRUE
nniLEP <- nni(lepto_kel)
nniLEP16 <- nni(loc_lep16)
nniLEP17 <- nni(loc_lep17)
nniLEP18 <- nni(loc_lep18)
nniLEP19 <- nni(loc_lep19)
nniLEP20 <- nni(loc_lep20)
nniLEP21 <- nni(loc_lep21)
nniLEP22 <- nni(loc_lep22)

NNI_LEP <- bind_rows(
  data.frame(Dataset = "lepto_kel", nniLEP),
  data.frame(Dataset = "loc_lep16", nniLEP16),
  data.frame(Dataset = "loc_lep17", nniLEP17),
  data.frame(Dataset = "loc_lep18", nniLEP18),
  data.frame(Dataset = "loc_lep19", nniLEP19),
  data.frame(Dataset = "loc_lep20", nniLEP20),
  data.frame(Dataset = "loc_lep21", nniLEP21),
  data.frame(Dataset = "loc_lep22", nniLEP22)
)
NNI_LEP

# NNI < 1 clustered, >1 dispersed under CSR
```


clark evans test for Enteric Fever
-R less than 1 indicates spatial clustering. The smaller the value of R, the greater the degree of clustering.
mean centre and standard distance for Enteric fever

```{r}
#| cache = TRUE
CElep <- clarkevans.test(lepto_kel.ppp2)
CElep16 <- clarkevans.test(loc_lep16.ppp2)
CElep17 <-clarkevans.test(loc_lep17.ppp2)
CElep18 <-clarkevans.test(loc_lep18.ppp2)
CElep19 <-clarkevans.test(loc_lep19.ppp2)
CElep20 <-clarkevans.test(loc_lep20.ppp2)
CElep21 <-clarkevans.test(loc_lep21.ppp2)
CElep22 <-clarkevans.test(loc_lep22.ppp2)


tCElep <- data.frame(Dataset = "lepto_kel.ppp2",
                       TestStatistic = CElep$statistic,
                       PValue = CElep$p.value)
tCElep16 <- data.frame(Dataset = "loc_lep16.ppp2",
                       TestStatistic = CElep16$statistic,
                       PValue = CElep16$p.value)
tCElep17 <- data.frame(Dataset = "loc_lep17.ppp2",
                       TestStatistic = CElep17$statistic,
                       PValue = CElep17$p.value)
tCElep18 <- data.frame(Dataset = "loc_lep18.ppp2",
                       TestStatistic = CElep18$statistic,
                       PValue = CElep18$p.value)
tCElep19 <- data.frame(Dataset = "loc_lep19.ppp2",
                       TestStatistic = CElep19$statistic,
                       PValue = CElep19$p.value)
tCElep20 <- data.frame(Dataset = "loc_lep20.ppp2",
                       TestStatistic = CElep20$statistic,
                       PValue = CElep20$p.value)
tCElep21 <- data.frame(Dataset = "loc_lep21.ppp2",
                       TestStatistic = CElep21$statistic,
                       PValue = CElep21$p.value)
tCElep22 <- data.frame(Dataset = "loc_lep22.ppp2",
                       TestStatistic = CElep22$statistic,
                       PValue = CElep22$p.value)

CE_LEPr <- bind_rows(
  data.frame(Dataset = "lepto_kel", tCElep),
  data.frame(Dataset = "loc_lep16", tCElep16),
  data.frame(Dataset = "loc_lep17", tCElep17),
  data.frame(Dataset = "loc_lep18", tCElep18),
  data.frame(Dataset = "loc_lep19", tCElep19),
  data.frame(Dataset = "loc_lep20", tCElep20),
  data.frame(Dataset = "loc_lep21", tCElep21),
  data.frame(Dataset = "loc_lep22", tCElep22)
)
CE_LEPr

# R less than 1 indicates spatial clustering. The smaller the value of R, the greater the degree of clustering.
```

mean center and standard distance for leptospirosis
```{r}
#calculate mean center (m) for leptospirosis
lepxy <- coords(lepto_kel.ppp2)
mclep <- summarize(lepxy, xmean = mean(x), ymean = mean(y))
mclep

#calculate standard distance (m) for leptospirosis
sdlep <- sqrt(sum((lepxy[,1] - mclep[1])^2 + (lepxy[,2] - mclep[2])^2) / nrow(lepxy))
sdlep
```

Median Absolute Deviation test with 99 simulations of Complete Spatial Randomness (CSR) for K function - enteric fever

```{r}
#| cache = TRUE
LEPMADTest16 <- mad.test(loc_lep16.ppp2.km, fun = Kest, nsim = 99, verbose = FALSE)
```

```{r}
#| cache = TRUE
LEPMADTest17 <- mad.test(loc_lep17.ppp2.km, fun = Kest, nsim = 99, verbose = FALSE)
```

```{r}
#| cache = TRUE
LEPMADTest18 <- mad.test(loc_lep18.ppp2.km, fun = Kest, nsim = 99, verbose = FALSE)
```

```{r}
#| cache = TRUE
LEPMADTest19 <- mad.test(loc_lep19.ppp2.km, fun = Kest, nsim = 99, verbose = FALSE)
```

```{r}
#| cache = TRUE
LEPMADTest20 <- mad.test(loc_lep20.ppp2.km, fun = Kest, nsim = 99, verbose = FALSE)
```

```{r}
#| cache = TRUE
LEPMADTest21 <- mad.test(loc_lep21.ppp2.km, fun = Kest, nsim = 99, verbose = FALSE)
```

```{r}
#| cache = TRUE
LEPMADTest22 <- mad.test(loc_lep21.ppp2.km, fun = Kest, nsim = 99, verbose = FALSE)
```

```{r}
#| cache = TRUE
LEPMADTest <- mad.test(lepto_kel.ppp2.km, fun = Kest, nsim = 99, verbose = FALSE)
```


Result MAD for K test leptospirosis

```{r}
#| cache = TRUE
MAD_lep16.df <- data.frame(Dataset = "loc_lep16.ppp2",
                       TestStatistic = LEPMADTest16$statistic,
                       PValue = LEPMADTest16$p.value)
MAD_lep17.df <- data.frame(Dataset = "loc_lep17.ppp2",
                       TestStatistic = LEPMADTest17$statistic,
                       PValue = LEPMADTest17$p.value)
MAD_lep18.df <- data.frame(Dataset = "loc_lep18.ppp2",
                       TestStatistic = LEPMADTest18$statistic,
                       PValue = LEPMADTest18$p.value)
MAD_lep19.df <- data.frame(Dataset = "loc_lep19.ppp2",
                       TestStatistic = LEPMADTest19$statistic,
                       PValue = LEPMADTest19$p.value)
MAD_lep20.df <- data.frame(Dataset = "loc_lep20.ppp2",
                       TestStatistic = LEPMADTest20$statistic,
                       PValue = LEPMADTest20$p.value)
MAD_lep21.df <- data.frame(Dataset = "loc_lep21.ppp2",
                       TestStatistic = LEPMADTest21$statistic,
                       PValue = LEPMADTest21$p.value)
MAD_lep22.df <- data.frame(Dataset = "loc_lep22.ppp2",
                       TestStatistic = LEPMADTest22$statistic,
                       PValue = LEPMADTest22$p.value)

#combine rows
MAD_LEP <- bind_rows(
  data.frame(Dataset = "2016", MAD_lep16.df),
  data.frame(Dataset = "2017", MAD_lep17.df),
  data.frame(Dataset = "2018", MAD_lep18.df),
  data.frame(Dataset = "2019", MAD_lep19.df),
  data.frame(Dataset = "2020", MAD_lep20.df),
  data.frame(Dataset = "2021", MAD_lep21.df),
  data.frame(Dataset = "2022", MAD_lep22.df)
)
MAD_LEP
```


K function leptospirosis

```{r}
#| cache = TRUE
KestLep16 <- Kest(loc_lep16.ppp2.km)
KestLep17 <- Kest(loc_lep17.ppp2.km)
KestLep18 <- Kest(loc_lep18.ppp2.km)
KestLep19 <- Kest(loc_lep19.ppp2.km)
KestLep20 <- Kest(loc_lep20.ppp2.km)
KestLep21 <- Kest(loc_lep21.ppp2.km)
KestLep22 <- Kest(loc_lep22.ppp2.km)
```

```{r}
#| cache = TRUE
KestLep <- Kest(lepto_kel.ppp2.km)
```


Plot K test
- Empirical values greater than theoretical values suggest clustering.
- significance already tested using MAD test + MC permutation

```{r name-of-chunk22a, fig.width=25, fig.height=8}
#| cache = TRUE
par( mfrow= c(2,4) )
plot(KestLep16, main = "2016", cex.main = 3, cex.axis = 1.5, cex.lab = 1.5)
plot(KestLep17, main = "2017", cex.main = 3, cex.axis = 1.5, cex.lab = 1.5)
plot(KestLep18, main = "2018", cex.main = 3, cex.axis = 1.5, cex.lab = 1.5)
plot(KestLep19, main = "2019", cex.main = 3, cex.axis = 1.5, cex.lab = 1.5)
plot(KestLep20, main = "2020", cex.main = 3, cex.axis = 1.5, cex.lab = 1.5)
plot(KestLep21, main = "2021", cex.main = 3, cex.axis = 1.5, cex.lab = 1.5)
plot(KestLep22, main = "2022", cex.main = 3, cex.axis = 1.5, cex.lab = 1.5)
plot(KestLep, main = "2016-2022", cex.main = 3, cex.axis = 1.5, cex.lab = 1.5)
```


calculate G function for Leptospirosis

```{r name-of-chunk23, fig.width=25, fig.height=8}
#| cache = TRUE
# G function leptospirosis
par( mfrow= c(2,4) )
plot(Gest(loc_lep16.ppp2.km), main = "2016", cex.main = 3, cex.lab = 1.5, cex.axis = 1.5)
plot(Gest(loc_lep17.ppp2.km), main = "2017", cex.main = 3, cex.lab = 1.5, cex.axis = 1.5)
plot(Gest(loc_lep18.ppp2.km), main = "2018", cex.main = 3, cex.lab = 1.5, cex.axis = 1.5)
plot(Gest(loc_lep19.ppp2.km), main = "2019", cex.main = 3, cex.lab = 1.5, cex.axis = 1.5)
plot(Gest(loc_lep20.ppp2.km), main = "2020", cex.main = 3, cex.lab = 1.5, cex.axis = 1.5)
plot(Gest(loc_lep21.ppp2.km), main = "2021", cex.main = 3, cex.lab = 1.5, cex.axis = 1.5)
plot(Gest(loc_lep22.ppp2.km), main = "2022", cex.main = 3, cex.lab = 1.5, cex.axis = 1.5)
plot(Gest(lepto_kel.ppp2.km), main = "2016-2022", cex.main = 3, cex.lab = 1.5, cex.axis = 1.5)
```


Enveloped G function

```{r name-of-chunk24a, fig.width=25, fig.height=10}
#| cache = TRUE
# G function leptospirosis fever per year
par( mfrow= c(2,4) )
GestLEP16 <- plot(envelope(loc_lep16.ppp2.km, Gest, nsim = 99, verbose = FALSE), main = "2016",xlab = "Distance (KM)", ylab = "Proportion of Leptospirosis Cases 2016", cex.main = 2, cex.axis = 1, cex.lab = 1)
GestLEP17 <- plot(envelope(loc_lep17.ppp2.km, Gest, nsim = 99, verbose = FALSE), main = "2017",xlab = "Distance (KM)", ylab = "Proportion of Leptospirosis Cases 2016", cex.main = 2, cex.axis = 1, cex.lab = 1)
GestLEP18 <- plot(envelope(loc_lep18.ppp2.km, Gest, nsim = 99, verbose = FALSE), main = "2018",xlab = "Distance (KM)", ylab = "Proportion of Leptospirosis Cases 2016", cex.main = 2, cex.axis = 1, cex.lab = 1)
GestLEP19 <- plot(envelope(loc_lep19.ppp2.km, Gest, nsim = 99, verbose = FALSE), main = "2019",xlab = "Distance (KM)", ylab = "Proportion of Leptospirosis Cases 2016", cex.main = 2, cex.axis = 1, cex.lab = 1)
GestLEP20 <- plot(envelope(loc_lep20.ppp2.km, Gest, nsim = 99, verbose = FALSE), main = "2020",xlab = "Distance (KM)", ylab = "Proportion of Leptospirosis Cases 2016", cex.main = 2, cex.axis = 1, cex.lab = 1)
GestLEP21 <- plot(envelope(loc_lep21.ppp2.km, Gest, nsim = 99, verbose = FALSE), main = "2021",xlab = "Distance (KM)", ylab = "Proportion of Leptospirosis Cases 2016", cex.main = 2, cex.axis = 1, cex.lab = 1)
GestLEP22 <- plot(envelope(loc_lep22.ppp2.km, Gest, nsim = 99, verbose = FALSE), main = "2022",xlab = "Distance (KM)", ylab = "Proportion of Leptospirosis Cases 2016", cex.main = 2, cex.axis = 1, cex.lab = 1)
GestLEP <- plot(envelope(lepto_kel.ppp2.km, Gest, nsim = 99, verbose = FALSE), main = "2016-2022",xlab = "Distance (KM)", ylab = "Proportion of Leptospirosis Cases 2016-2022", cex.main = 2, cex.axis = 1, cex.lab = 1)

mtext("Enteric Fever Pattern in Kelantan 2016-2022", side = 3, line = -1, cex = 1.5, outer = TRUE)
```


F test of leptospirosis cases 2016-2022
- empirical  values below theoretical (blue) suggest clustering

```{r name-of-chunk25, fig.width=28, fig.height=14}
#| cache = TRUE
par( mfrow= c(2,4) )
plot(Fest(loc_lep16.ppp2.km), main = "2016", cex.main = 3, cex.axis = 1.5, cex.lab = 1.5)
plot(Fest(loc_lep17.ppp2.km), main = "2017", cex.main = 3, cex.axis = 1.5, cex.lab = 1.5)
plot(Fest(loc_lep18.ppp2.km), main = "2018", cex.main = 3, cex.axis = 1.5, cex.lab = 1.5)
plot(Fest(loc_lep19.ppp2.km), main = "2019", cex.main = 3, cex.axis = 1.5, cex.lab = 1.5)
plot(Fest(loc_lep20.ppp2.km), main = "2020", cex.main = 3, cex.axis = 1.5, cex.lab = 1.5)
plot(Fest(loc_lep21.ppp2.km), main = "2021", cex.main = 3, cex.axis = 1.5, cex.lab = 1.5)
plot(Fest(loc_lep22.ppp2.km), main = "2022", cex.main = 3, cex.axis = 1.5, cex.lab = 1.5)
plot(Fest(lepto_kel.ppp2.km), main = "2016-2022", cex.main = 3, cex.axis = 1.5, cex.lab = 1.5)

mtext("Leptospirosis Pattern in Kelantan 2016-2022", side = 1, line = -1, cex = 1.5, outer = TRUE)
```


##Spatial autocorrelation using point data

# spatial autocorrelation for enteric fever cases

extract enteric fever 2016-2022

```{r}
#| cache = TRUE
linelist2 <- read_xlsx(here ("linelist.xlsx"))
listENT_kel <- filter(linelist2, Diagnosis=="Enteric_fever")  
listENT_kel16 <- filter(listENT_kel, `Tahun Daftar` =="2016")
listENT_kel17 <- filter(listENT_kel, `Tahun Daftar` =="2017")
listENT_kel18 <- filter(listENT_kel, `Tahun Daftar` =="2018")
listENT_kel19 <- filter(listENT_kel, `Tahun Daftar` =="2019")
listENT_kel20 <- filter(listENT_kel, `Tahun Daftar` =="2020")
listENT_kel21 <- filter(listENT_kel, `Tahun Daftar` =="2021")
listENT_kel22 <- filter(listENT_kel, `Tahun Daftar` =="2022")
```


calculate global moran

```{r}
#| cache = TRUE
# join point to polygon
polyENT_kel <- merge(kel_map, listENT_kel, by = c("MUKIM"))

# count cases in mukim
count_ent <- polyENT_kel%>% 
  count(DAERAH, MUKIM, AVR) %>% 
  ungroup()

# set neigbouring queen
nb_ent <- poly2nb(count_ent, queen = TRUE) 

# identify area with no neighbour
nb_ent[[1]]

# assign weight
lw_ent <- nb2listw(nb_ent, style = "W" , zero.policy = TRUE)

# create lag function
ent_lag <- lag.listw(lw_ent, count_ent$n) 

# calculate global Moran
mt_ent <- moran.test(count_ent$n, lw_ent)

mt_ent
```


calculate local moran

```{r}
#| cache = TRUE
#local moran Enteric fever
Local_Moran_ent <- localmoran(count_ent$n,lw_ent)
ENT_poly <- cbind(count_ent, Local_Moran_ent)
#plot Local Moran
ggplot(data = ENT_poly) +
  geom_sf(aes(fill=Ii)) +
  theme_bw() +
  scale_fill_gradient2(low="#2c7bb6", mid="#ffffbf", high="#d7191c",
                       name="Local Moran's I") +
  labs(title="Local Moran's I statistic for Enteric Fever 2016-2022",
       subtitle="Kelantan")
```


Moran's I by year

```{r}
#| cache = TRUE
#Morans I by year 
#2016
polyENT_kel16 <- merge(kel_map, listENT_kel16, by = c("MUKIM"))
count_ent16 <- polyENT_kel16%>% 
  count(MUKIM, AVR, JUMLAH_2016) %>% 
  ungroup()
count_ent16s <-count_ent16[-13, ] #remove region with no link from polygon
nb_ent16 <- poly2nb(count_ent16, queen = TRUE) #set neigbouring queen
nb_ent16 # identify region with no link
nb_ent16s <-subset(nb_ent16, subset=card(nb_ent16) > 0) #remove region with no link
lw_ent16 <- nb2listw(nb_ent16s, style = "W" , zero.policy = TRUE) #assign weight
ent_lag16 <- lag.listw(lw_ent16, count_ent16s$n) #create lag function
gmoran16 <- moran.test(count_ent16s$n, lw_ent16, randomisation=TRUE, zero.policy=NULL,
                       alternative="greater", rank = FALSE, na.action=na.fail, spChk=NULL,
                       adjust.n=TRUE, drop.EI2=FALSE)
gmoran16

#local moran Enteric fever
Local_Moran_ent16 <- localmoran(count_ent16s$n,lw_ent16)
ENT_poly16 <- cbind(count_ent16s, Local_Moran_ent16)
#plot Local Moran
ggplot(data = ENT_poly16) +
  geom_sf(aes(fill=Ii)) +
  theme_bw() +
  scale_fill_gradient2(low="#2c7bb6", mid="#ffffbf", high="#d7191c",
                       name="Local Moran's I") +
  labs(title="Local Moran's I Statistic for Enteric Fever Cases in 2016",
       subtitle="Kelantan")
```


```{r}
#| cache = TRUE
#Moran 2017
polyENT_kel17 <- merge(kel_map, listENT_kel17, by = c("MUKIM"))
count_ent17 <- polyENT_kel17%>% 
  count(DAERAH, MUKIM, AVR, JUMLAH_2017) %>% 
  ungroup()
nb_ent17 <- poly2nb(count_ent17, queen = TRUE) #set neigbouring queen
nb_ent17 # identify region with no
nb_ent17s <-subset(nb_ent17, subset=card(nb_ent17) > 0)
nb_ent17s
count_ent17s <- count_ent17[-c(5,6,14,15,16,17,18),]
count_ent17s
lw_ent17 <- nb2listw(nb_ent17s, style = "W" , zero.policy = TRUE) #assign weight
ent_lag17 <- lag.listw(lw_ent17, count_ent17s$n) #create lag function
gmoran17 <- moran.test(count_ent17s$n, lw_ent17)
gmoran17

#local moran Enteric fever
Local_Moran_ent17 <- localmoran(count_ent17s$n,lw_ent17)
ENT_poly17 <- cbind(count_ent17s, Local_Moran_ent17)
#plot Local Moran
ggplot(data = ENT_poly17) +
  geom_sf(aes(fill=Ii)) +
  theme_bw() +
  scale_fill_gradient2(low="#2c7bb6", mid="#ffffbf", high="#d7191c",
                       name="Local Moran's I") +
  labs(title="Local Moran's I Statistic for Enteric Fever Cases in 2017",
       subtitle="Kelantan")
```


```{r}
#| cache = TRUE
#Moran 2018
polyENT_kel18 <- merge(kel_map, listENT_kel18, by = c("MUKIM"))
count_ent18 <- polyENT_kel18%>% 
  count(DAERAH, MUKIM, AVR, JUMLAH_2018) %>% 
  ungroup()
nb_ent18 <- poly2nb(count_ent18, queen = TRUE) #set neigbouring queen
nb_ent18 #identify region with no link
nb_ent18s <-subset(nb_ent18, subset=card(nb_ent18) > 0)
count_ent18s <-count_ent18[-c(1,4,18),]
lw_ent18 <- nb2listw(nb_ent18s, style = "W" , zero.policy = TRUE) #assign weight
ent_lag18 <- lag.listw(lw_ent18, count_ent18s$n) #create lag function
gmoran18 <- moran.test(Count_ent18s$n, lw_ent18)
gmoran18

#local moran Enteric fever
Local_Moran_ent18 <- localmoran(count_ent18s$n,lw_ent18)
ENT_poly18 <- cbind(count_ent18s, Local_Moran_ent18)
#plot Local Moran
ggplot(data = ENT_poly18) +
  geom_sf(aes(fill=Ii)) +
  theme_bw() +
  scale_fill_gradient2(low="#2c7bb6", mid="#ffffbf", high="#d7191c",
                       name="Local Moran's I") +
  labs(title="Local Moran's I Statistic for Enteric Fever Cases in 2018",
       subtitle="Kelantan")
```


```{r}
#| cache = TRUE
#Moran 2019
polyENT_kel19 <- merge(kel_map, listENT_kel19, by = c("MUKIM"))
count_ent19 <- polyENT_kel19%>% 
  count(DAERAH, MUKIM, AVR, JUMLAH_2019) %>% 
  ungroup()
nb_ent19 <- poly2nb(count_ent19, queen = TRUE) #set neigbouring queen
nb_ent19
nb_ent19s <-subset(nb_ent19, subset=card(nb_ent19) > 0)
count_ent19s <-count_ent19[-c(4,9,10,11,12),]
lw_ent19 <- nb2listw(nb_ent19s, style = "W" , zero.policy = TRUE) #assign weight
ent_lag19 <- lag.listw(lw_ent19, count_ent19s$n) #create lag function
gmoran19 <- moran.test(count_ent19s$n, lw_ent19)
gmoran19

#local moran Enteric fever
Local_Moran_ent19 <- localmoran(count_ent19s$n,lw_ent19)
ENT_poly19 <- cbind(count_ent19s, Local_Moran_ent19)
#plot Local Moran
ggplot(data = ENT_poly19) +
  geom_sf(aes(fill=Ii)) +
  theme_bw() +
  scale_fill_gradient2(low="#2c7bb6", mid="#ffffbf", high="#d7191c",
                       name="Local Moran's I") +
  labs(title="Local Moran's I Statistic for Enteric Fever Cases in 2019",
       subtitle="Kelantan")
```


```{r}
#| cache = TRUE
#Moran 2020
polyENT_kel20 <- merge(kel_map, listENT_kel20, by = c("MUKIM"))
count_ent20 <- polyENT_kel20%>% 
  count(DAERAH, MUKIM, AVR, JUMLAH_2020) %>% 
  ungroup()
nb_ent20 <- poly2nb(count_ent20, queen = TRUE) #set neigbouring queen
nb_ent20 #identify region with no link
nb_ent20s <-subset(nb_ent20, subset=card(nb_ent20) > 0)
count_ent20s <-count_ent20[-c(3,9),]
lw_ent20 <- nb2listw(nb_ent20s, style = "W" , zero.policy = TRUE) #assign weight
ent_lag20 <- lag.listw(lw_ent20, count_ent20s$n) #create lag function
gmoran20 <-moran.test(count_ent20s$n, lw_ent20)
gmoran20

#local moran Enteric fever
Local_Moran_ent20 <- localmoran(count_ent20s$n,lw_ent20)
ENT_poly20 <- cbind(count_ent20s, Local_Moran_ent20)
#plot Local Moran
ggplot(data = ENT_poly20) +
  geom_sf(aes(fill=Ii)) +
  theme_bw() +
  scale_fill_gradient2(low="#2c7bb6", mid="#ffffbf", high="#d7191c",
                       name="Local Moran's I") +
  labs(title="Local Moran's I Statistic for Enteric Fever Cases in 2020",
       subtitle="Kelantan")
```


```{r, eval=FALSE}
#| cache = TRUE
#Moran 21
polyENT_kel21 <- merge(kel_map, listENT_kel21, by = c("MUKIM"))
count_ent21 <- polyENT_kel21%>% 
  count(DAERAH, MUKIM, AVR, JUMLAH_2021) %>% 
  ungroup()
nb_ent21 <- poly2nb(count_ent21, queen = TRUE) #set neigbouring queen
nb_ent21
nb_ent21s <-subset(nb_ent21, subset=card(nb_ent21) > 0)
count_ent21s <-count_ent21[-c(3),]
lw_ent21 <- nb2listw(nb_ent21s, style = "W" , zero.policy = TRUE) #assign weight
ent_lag21 <- lag.listw(lw_ent21, count_ent21s$n) #create lag function
gmoran21 <- moran.test(count_ent21s$n, lw_ent21)
gmoran21

#local moran Enteric fever
Local_Moran_ent21 <- localmoran(count_ent21s$n,lw_ent21)
ENT_poly21 <- cbind(count_ent21s, Local_Moran_ent21)
#plot Local Moran
ggplot(data = ENT_poly21) +
  geom_sf(aes(fill=Ii)) +
  theme_bw() +
  scale_fill_gradient2(low="#2c7bb6", mid="#ffffbf", high="#d7191c",
                       name="Local Moran's I") +
  labs(title="Local Moran's I Statistic for Enteric Fever Cases in 2021",
       subtitle="Kelantan")
```


```{r}
#| cache = TRUE
#Moran 2022
polyENT_kel22 <- merge(kel_map, listENT_kel22, by = c("MUKIM"))
count_ent22 <- polyENT_kel22%>% 
  count(DAERAH, MUKIM, AVR, JUMLAH_2022) %>% 
  ungroup()
nb_ent22 <- poly2nb(count_ent22, queen = TRUE) #set neigbouring queen
nb_ent22
nb_ent22s <-subset(nb_ent22, subset=card(nb_ent22) > 0)
count_ent22s <-count_ent22[-c(1,3,8,10),]
lw_ent22 <- nb2listw(nb_ent22s, style = "W" , zero.policy = TRUE) #assign weight
ent_lag22 <- lag.listw(lw_ent22, count_ent22s$n) #create lag function
gmoran22 <- moran.test(count_ent22s$n, lw_ent22)
gmoran22


#local moran Enteric fever
Local_Moran_ent22 <- localmoran(count_ent22s$n,lw_ent22)
ENT_poly22 <- cbind(count_ent22s, Local_Moran_ent22)
#plot Local Moran
ggplot(data = ENT_poly22) +
  geom_sf(aes(fill=Ii)) +
  theme_bw() +
  scale_fill_gradient2(low="#2c7bb6", mid="#ffffbf", high="#d7191c",
                       name="Local Moran's I") +
  labs(title="Local Moran's I Statistic for Enteric Fever Cases in 2022",
       subtitle="Kelantan")
```


# spatial autocorrelation for leptospirosis cases

prepare linelisting

```{r}
#| cache = TRUE
#extract leptospirosis 2016-2022
linelist2 <- read_xlsx(here ("linelist.xlsx"))
listLEP_kel <- filter(linelist2, Diagnosis=="Leptospirosis")
LEP_kel16 <- filter(linelist2, `Tahun Daftar` =="2016")
LEP_kel17 <- filter(linelist2, `Tahun Daftar` =="2017")
LEP_kel18 <- filter(linelist2, `Tahun Daftar` =="2018")
LEP_kel19 <- filter(linelist2, `Tahun Daftar` =="2019")
LEP_kel20 <- filter(linelist2, `Tahun Daftar` =="2020")
LEP_kel21 <- filter(linelist2, `Tahun Daftar` =="2021")
LEP_kel22 <- filter(linelist2, `Tahun Daftar` =="2022")
```


Global Moran

```{r}
#| cache = TRUE
#Moran's I for leptospirosis 2016-2022
polyLEP_kel <- merge(kel_map, listLEP_kel, by = c("MUKIM"))
count_lep <- polyLEP_kel%>% 
  count(DAERAH, MUKIM, AVR) %>% 
  ungroup()
nb_lep <- poly2nb(count_lep, queen = TRUE) #set neigbouring queen
nb_lep[[1]]
lw_lep <- nb2listw(nb_lep, style = "W" , zero.policy = TRUE) #assign weight
lep_lag <- lag.listw(lw_lep, count_lep$n) #create lag function
moran.test(count_lep$n, lw_lep)

```

```{r}
#| cache = TRUE
#Moran's I for leptospirosis 2016
polyLEP_kel16 <- merge(kel_map, LEP_kel16, by = c("MUKIM"))
count_lep16 <- polyLEP_kel16%>% 
  count(DAERAH, MUKIM, AVR, JUMLAH_2016) %>% 
  ungroup()
nb_lep16 <- poly2nb(count_lep16, queen = TRUE) #set neigbouring queen
nb_lep16[[1]]
lw_lep16 <- nb2listw(nb_lep16, style = "W" , zero.policy = TRUE) #assign weight
lep16_lag <- lag.listw(lw_lep16, count_lep16$n) #create lag function
moran.test(count_lep16$n, lw_lep16)
```

```{r}
#| cache = TRUE
#Moran's I for leptospirosis 2017
polyLEP_kel17 <- merge(kel_map, LEP_kel17, by = c("MUKIM"))
count_lep17 <- polyLEP_kel17%>% 
  count(DAERAH, MUKIM, AVR, JUMLAH_2017) %>% 
  ungroup()
nb_lep17 <- poly2nb(count_lep17, queen = TRUE) #set neigbouring queen
nb_lep17[[1]]
lw_lep17 <- nb2listw(nb_lep17, style = "W" , zero.policy = TRUE) #assign weight
lep17_lag <- lag.listw(lw_lep17, count_lep17$n) #create lag function
moran.test(count_lep17$n, lw_lep17)
```

```{r}
#| cache = TRUE
#Moran's I for leptospirosis 2018
polyLEP_kel18 <- merge(kel_map, LEP_kel18, by = c("MUKIM"))
count_lep18 <- polyLEP_kel18%>% 
  count(DAERAH, MUKIM, AVR, JUMLAH_2018) %>% 
  ungroup()
nb_lep18 <- poly2nb(count_lep18, queen = TRUE) #set neigbouring queen
nb_lep18[[1]]
lw_lep18 <- nb2listw(nb_lep18, style = "W" , zero.policy = TRUE) #assign weight
lep18_lag <- lag.listw(lw_lep18, count_lep18$n) #create lag function
moran.test(count_lep18$n, lw_lep18)
```

```{r}
#| cache = TRUE
#Moran's I for leptospirosis 2019
polyLEP_kel19 <- merge(kel_map, LEP_kel19, by = c("MUKIM"))
count_lep19 <- polyLEP_kel19%>% 
  count(DAERAH, MUKIM, AVR, JUMLAH_2019) %>% 
  ungroup()
nb_lep19 <- poly2nb(count_lep19, queen = TRUE) #set neigbouring queen
nb_lep19[[1]]
lw_lep19 <- nb2listw(nb_lep19, style = "W" , zero.policy = TRUE) #assign weight
lep19_lag <- lag.listw(lw_lep19, count_lep19$n) #create lag function
moran.test(count_lep19$n, lw_lep19)
```

```{r}
#| cache = TRUE
#Moran's I for leptospirosis 2020
polyLEP_kel20 <- merge(kel_map, LEP_kel20, by = c("MUKIM"))
count_lep20 <- polyLEP_kel20%>% 
  count(DAERAH, MUKIM, AVR, JUMLAH_2020) %>% 
  ungroup()
nb_lep20 <- poly2nb(count_lep20, queen = TRUE) #set neigbouring queen
nb_lep20[[1]]
lw_lep20 <- nb2listw(nb_lep20, style = "W" , zero.policy = TRUE) #assign weight
lep20_lag <- lag.listw(lw_lep20, count_lep20$n) #create lag function
moran.test(count_lep20$n, lw_lep20)
```

```{r}
#| cache = TRUE
#Moran's I for leptospirosis 2021
polyLEP_kel21 <- merge(kel_map, LEP_kel21, by = c("MUKIM"))
count_lep21 <- polyLEP_kel21%>% 
  count(DAERAH, MUKIM, AVR, JUMLAH_2021) %>% 
  ungroup()
nb_lep21 <- poly2nb(count_lep21, queen = TRUE) #set neigbouring queen
nb_lep21[[1]]
lw_lep21 <- nb2listw(nb_lep21, style = "W" , zero.policy = TRUE) #assign weight
lep21_lag <- lag.listw(lw_lep21, count_lep21$n) #create lag function
moran.test(count_lep21$n, lw_lep21)
```

```{r}
#| cache = TRUE
#Moran's I for leptospirosis 2022
polyLEP_kel22 <- merge(kel_map, LEP_kel22, by = c("MUKIM"))
count_lep22 <- polyLEP_kel22%>% 
  count(DAERAH, MUKIM, AVR, JUMLAH_2022) %>% 
  ungroup()
nb_lep22 <- poly2nb(count_lep22, queen = TRUE) #set neigbouring queen
nb_lep22[[1]]
lw_lep22 <- nb2listw(nb_lep22, style = "W" , zero.policy = TRUE) #assign weight
lep22_lag <- lag.listw(lw_lep22, count_lep22$n) #create lag function
moran.test(count_lep22$n, lw_lep22)
```

Local Moran

```{r}
#| cache = TRUE
#local moran leptospirosis 2016-2022
Local_Moran_lep <- localmoran(count_lep$n,lw_lep)
LEP_poly <- cbind(count_lep, Local_Moran_lep)
#plot Local Moran
ggplot(data = LEP_poly) +
  geom_sf(aes(fill=Ii)) +
  theme_bw() +
  scale_fill_gradient2(low="#2c7bb6", mid="#ffffbf", high="#d7191c",
                       name="Local Moran's I") +
  labs(title="Local Moran's I statistic for Leptospirosis 2016-2022",
       subtitle="Kelantan")
```


```{r}
#| cache = TRUE
#2016
Local_Moran_lep16 <- localmoran(count_lep16$n,lw_lep16)
LEP16_poly <- cbind(count_lep16, Local_Moran_lep16)
#plot Local Moran
plotMlep16 <- ggplot(data = LEP16_poly) +
  geom_sf(aes(fill=Ii)) +
  theme_bw() +
  scale_fill_gradient2(low="#2c7bb6", mid="#ffffbf", high="#d7191c",
                       name="Local Moran's I") +
  labs(title="2016")

#2017
Local_Moran_lep17 <- localmoran(count_lep17$n,lw_lep17)
LEP17_poly <- cbind(count_lep17, Local_Moran_lep17)
#plot Local Moran
plotMlep17 <- ggplot(data = LEP17_poly) +
  geom_sf(aes(fill=Ii)) +
  theme_bw() +
  scale_fill_gradient2(low="#2c7bb6", mid="#ffffbf", high="#d7191c",
                       name="Local Moran's I") +
  labs(title="2017")

#2018
Local_Moran_lep18 <- localmoran(count_lep18$n,lw_lep18)
LEP18_poly <- cbind(count_lep18, Local_Moran_lep18)
#plot Local Moran
plotMlep18 <- ggplot(data = LEP18_poly) +
  geom_sf(aes(fill=Ii)) +
  theme_bw() +
  scale_fill_gradient2(low="#2c7bb6", mid="#ffffbf", high="#d7191c",
                       name="Local Moran's I") +
  labs(title="2018")

#2019
Local_Moran_lep19 <- localmoran(count_lep19$n,lw_lep19)
LEP19_poly <- cbind(count_lep19, Local_Moran_lep19)
#plot Local Moran
plotMlep19 <-ggplot(data = LEP19_poly) +
  geom_sf(aes(fill=Ii)) +
  theme_bw() +
  scale_fill_gradient2(low="#2c7bb6", mid="#ffffbf", high="#d7191c",
                       name="Local Moran's I") +
  labs(title="2019")

#2020 - Global Moran's not significant
#2021 - Global Moran's not significant

#2022
Local_Moran_lep22 <- localmoran(count_lep22$n,lw_lep22)
LEP22_poly <- cbind(count_lep22, Local_Moran_lep22)
#plot Local Moran
plotMlep22 <-ggplot(data = LEP22_poly) +
  geom_sf(aes(fill=Ii)) +
  theme_bw() +
  scale_fill_gradient2(low="#2c7bb6", mid="#ffffbf", high="#d7191c",
                       name="Local Moran's I") +
  labs(title="2022")

#2016-2022
plotMlep <-ggplot(data = LEP_poly) +
  geom_sf(aes(fill=Ii)) +
  theme_bw() +
  scale_fill_gradient2(low="#2c7bb6", mid="#ffffbf", high="#d7191c",
                       name="Local Moran's I") +
  labs(title="2016-2022")
```


local Moran's I p-values map for leptospirosis

```{r}
#| cache = TRUE
LMp16 <- tm_shape(LEP16_poly) +
  tm_fill(col = "Pr.z....E.Ii..",
          breaks=c(-Inf, 0.001, 0.01, 0.05, 0.1, Inf),
          palette = "-Purples",
          title = "2016")+
  tm_borders(alpha = 0.5)

LMp17 <- tm_shape(LEP17_poly) +
  tm_fill(col = "Pr.z....E.Ii..",
          breaks=c(-Inf, 0.001, 0.01, 0.05, 0.1, Inf),
          palette = "-Purples",
          title = "2017")+
  tm_borders(alpha = 0.5)

LMp18 <- tm_shape(LEP18_poly) +
  tm_fill(col = "Pr.z....E.Ii..",
          breaks=c(-Inf, 0.001, 0.01, 0.05, 0.1, Inf),
          palette = "-Purples",
          title = "2018")+
  tm_borders(alpha = 0.5)

LMp19 <- tm_shape(LEP19_poly) +
  tm_fill(col = "Pr.z....E.Ii..",
          breaks=c(-Inf, 0.001, 0.01, 0.05, 0.1, Inf),
          palette = "-Purples",
          title = "2019")+
  tm_borders(alpha = 0.5)

LMp22 <- tm_shape(LEP22_poly) +
  tm_fill(col = "Pr.z....E.Ii..",
          breaks=c(-Inf, 0.001, 0.01, 0.05, 0.1, Inf),
          palette = "-Purples",
          title = "2022")+
  tm_borders(alpha = 0.5)

LMp <- tm_shape(LEP_poly) +
  tm_fill(col = "Pr.z....E.Ii..",
          breaks=c(-Inf, 0.001, 0.01, 0.05, 0.1, Inf),
          palette = "-Purples",
          title = "2016-2022")+
  tm_borders(alpha = 0.5)
```


combine all local moran plot for leptospirosis  in one view

```{r name-of-chunk26, fig.width=14, fig.height=9}
#| cache = TRUE
grid.arrange(plotMlep16, plotMlep17, plotMlep18, plotMlep19, plotMlep22,plotMlep, nrow=2, top=textGrob("Local Moran's I Map of Leptospirosis in Kelantan 2016-2022", gp = gpar(fontsize = 20)))
```


combine all local moran pvalue plot for leptospirosis  in one view

```{r name-of-chunk27, fig.width=16, fig.height=9}
#| cache = TRUE
tmap_arrange(LMp16, LMp17, LMp18, LMp19, LMp22, LMp, ncol = 3) 
```


LISA map for Leptospirosis

```{r}
#| cache = TRUE
#LISA 2016-2022
m_lep <- count_lep$n - mean(count_lep$n) #Center the variable of interest around its mean
m_local_lep <- Local_Moran_lep[,1]- mean(Local_Moran_lep[,1]) #Center the local Moran around the mean
signif <- 0.05 #significant threshold
quadrant_lep <- vector(mode="numeric",length = nrow(Local_Moran_lep))
#set classification - built data quadrant
quadrant_lep [m_lep>0 & m_local_lep>0] <- 4
quadrant_lep [m_lep<0 & m_local_lep<0] <- 1
quadrant_lep [m_lep<0 & m_local_lep>0] <- 2
quadrant_lep [m_lep>0 & m_local_lep<0] <- 3
quadrant_lep [Local_Moran_lep[,5]] <- 0
#plot LISA
brks <- c(0,1,2,3,4)
colors <- c("white", "blue",rgb(0,0,1,alpha=0.4),rgb(1,0,0,alpha = 0.4),"red")
plot(count_lep[,5],main = "LISA Map of Leptospirosis Cases 2016-2022", las = 1, border="lightgray",col=colors[findInterval(quadrant_lep,brks,all.inside = FALSE)])
legend("bottomright",legend = c("insignificant","low-low","low-high","high-low","high-high"),
       fill=colors,bty="n")

#2016
m_lep16 <- count_lep16$n - mean(count_lep16$n) 
m_local_lep16 <- Local_Moran_lep16[,1]- mean(Local_Moran_lep16[,1]) 
signif <- 0.05 #significant threshold
quadrant_lep16 <- vector(mode="numeric",length = nrow(Local_Moran_lep16))
quadrant_lep16 [m_lep16>0 & m_local_lep16>0] <- 4
quadrant_lep16 [m_lep16<0 & m_local_lep16<0] <- 1
quadrant_lep16 [m_lep16<0 & m_local_lep16>0] <- 2
quadrant_lep16 [m_lep16>0 & m_local_lep16<0] <- 3
quadrant_lep16 [Local_Moran_lep16[,5]] <- 0
#plot LISA
brks <- c(0,1,2,3,4)
colors <- c("white", "blue",rgb(0,0,1,alpha=0.4),rgb(1,0,0,alpha = 0.4),"red")
plot(count_lep16[,5],main = "LISA Map of Leptospirosis Cases 2016", las = 1, border="lightgray",col=colors[findInterval(quadrant_lep16,brks,all.inside = FALSE)])
legend("bottomright",legend = c("insignificant","low-low","low-high","high-low","high-high"),
       fill=colors,bty="n")

#2017
m_lep17 <- count_lep17$n - mean(count_lep17$n) 
m_local_lep17 <- Local_Moran_lep16[,1]- mean(Local_Moran_lep17[,1]) 
signif <- 0.05
quadrant_lep17 <- vector(mode="numeric",length = nrow(Local_Moran_lep17))
quadrant_lep17 [m_lep17>0 & m_local_lep17>0] <- 4
quadrant_lep17 [m_lep17<0 & m_local_lep17<0] <- 1
quadrant_lep17 [m_lep17<0 & m_local_lep17>0] <- 2
quadrant_lep17 [m_lep17>0 & m_local_lep17<0] <- 3
quadrant_lep17 [Local_Moran_lep17[,5]] <- 0
#plot LISA
brks <- c(0,1,2,3,4)
colors <- c("white", "blue",rgb(0,0,1,alpha=0.4),rgb(1,0,0,alpha = 0.4),"red")
plot(count_lep17[,5],main = "LISA Map of Leptospirosis Cases 2017", las = 1, border="lightgray",col=colors[findInterval(quadrant_lep17,brks,all.inside = FALSE)])
legend("bottomright",legend = c("insignificant","low-low","low-high","high-low","high-high"),
       fill=colors,bty="n")

#2018
m_lep18 <- count_lep18$n - mean(count_lep18$n) 
m_local_lep18 <- Local_Moran_lep18[,1]- mean(Local_Moran_lep18[,1]) 
signif <- 0.05
quadrant_lep18 <- vector(mode="numeric",length = nrow(Local_Moran_lep17))
quadrant_lep18 [m_lep18>0 & m_local_lep18>0] <- 4
quadrant_lep18 [m_lep18<0 & m_local_lep18<0] <- 1
quadrant_lep18 [m_lep18<0 & m_local_lep18>0] <- 2
quadrant_lep18 [m_lep18>0 & m_local_lep18<0] <- 3
quadrant_lep18 [Local_Moran_lep18[,5]] <- 0
#plot LISA
brks <- c(0,1,2,3,4)
colors <- c("white", "blue",rgb(0,0,1,alpha=0.4),rgb(1,0,0,alpha = 0.4),"red")
plot(count_lep18[,5],main = "LISA Map of Leptospirosis Cases 2018", las = 1, border="lightgray",col=colors[findInterval(quadrant_lep18,brks,all.inside = FALSE)])
legend("bottomright",legend = c("insignificant","low-low","low-high","high-low","high-high"),
       fill=colors,bty="n")

#2019
m_lep19 <- count_lep19$n - mean(count_lep19$n) 
m_local_lep19 <- Local_Moran_lep19[,1]- mean(Local_Moran_lep19[,1]) 
signif <- 0.05
quadrant_lep19 <- vector(mode="numeric",length = nrow(Local_Moran_lep17))
quadrant_lep19 [m_lep19>0 & m_local_lep19>0] <- 4
quadrant_lep19 [m_lep19<0 & m_local_lep19<0] <- 1
quadrant_lep19 [m_lep19<0 & m_local_lep19>0] <- 2
quadrant_lep19 [m_lep19>0 & m_local_lep19<0] <- 3
quadrant_lep19 [Local_Moran_lep19[,5]] <- 0
#plot LISA
brks <- c(0,1,2,3,4)
colors <- c("white", "blue",rgb(0,0,1,alpha=0.4),rgb(1,0,0,alpha = 0.4),"red")
plot(count_lep19[,5],main = "LISA Map of Leptospirosis Cases 2019", las = 1, border="lightgray",col=colors[findInterval(quadrant_lep19,brks,all.inside = FALSE)])
legend("bottomright",legend = c("insignificant","low-low","low-high","high-low","high-high"),
       fill=colors,bty="n")

#2022
m_lep22 <- count_lep22$n - mean(count_lep22$n) 
m_local_lep22 <- Local_Moran_lep22[,1]- mean(Local_Moran_lep22[,1]) 
signif <- 0.05
quadrant_lep22 <- vector(mode="numeric",length = nrow(Local_Moran_lep17))
quadrant_lep22 [m_lep19>0 & m_local_lep22>0] <- 4
quadrant_lep22 [m_lep19<0 & m_local_lep22<0] <- 1
quadrant_lep22 [m_lep19<0 & m_local_lep22>0] <- 2
quadrant_lep22 [m_lep19>0 & m_local_lep22<0] <- 3
quadrant_lep22 [Local_Moran_lep22[,5]] <- 0

#plot LISA
brks <- c(0,1,2,3,4)
colors <- c("white", "blue",rgb(0,0,1,alpha=0.4),rgb(1,0,0,alpha = 0.4),"red")
plot(count_lep22[,5],main = "LISA Map of Leptospirosis Cases 2022", las = 1, border="lightgray",col=colors[findInterval(quadrant_lep22,brks,all.inside = FALSE)])
legend("bottomright",legend = c("insignificant","low-low","low-high","high-low","high-high"),
       fill=colors,bty="n")
```

calculate Getis ord i* statistic for leptospirosis cases

```{r}
#| cache = TRUE
#2016
getis_ordlep16 <- localG(count_lep16$n, lw_lep16)

# join results to sf data
count_lep16$getis_ord <- getis_ordlep16

# plot map
GiLEP16 <- ggplot(data=count_lep16) +
  geom_sf(aes(fill=getis_ord)) +
  theme_bw() +
  scale_fill_gradient2(low="#2c7bb6", mid="#ffffbf", high="#d7191c",
                       name="Gi*") +
  labs(title="2016")

#2017
getis_ordlep17 <- localG(count_lep17$n, lw_lep17)
count_lep17$getis_ord <- getis_ordlep17
GiLEP17 <- ggplot(data=count_lep17) +
  geom_sf(aes(fill=getis_ord)) +
  theme_bw() +
  scale_fill_gradient2(low="#2c7bb6", mid="#ffffbf", high="#d7191c",
                       name="Gi*") +
  labs(title="2017")

#2018
getis_ordlep18 <- localG(count_lep18$n, lw_lep18)
count_lep18$getis_ord <- getis_ordlep18
GiLEP18 <- ggplot(data=count_lep18) +
  geom_sf(aes(fill=getis_ord)) +
  theme_bw() +
  scale_fill_gradient2(low="#2c7bb6", mid="#ffffbf", high="#d7191c",
                       name="Gi*") +
  labs(title="2018")

#2019
getis_ordlep19 <- localG(count_lep19$n, lw_lep19)
count_lep19$getis_ord <- getis_ordlep19
GiLEP19 <- ggplot(data=count_lep19) +
  geom_sf(aes(fill=getis_ord)) +
  theme_bw() +
  scale_fill_gradient2(low="#2c7bb6", mid="#ffffbf", high="#d7191c",
                       name="Gi*") +
  labs(title="2019")

#2020
getis_ordlep20 <- localG(count_lep20$n, lw_lep20)
count_lep20$getis_ord <- getis_ordlep20
GiLEP20 <- ggplot(data=count_lep20) +
  geom_sf(aes(fill=getis_ord)) +
  theme_bw() +
  scale_fill_gradient2(low="#2c7bb6", mid="#ffffbf", high="#d7191c",
                       name="Gi*") +
  labs(title="2020")

#2021
getis_ordlep21 <- localG(count_lep21$n, lw_lep21)
count_lep21$getis_ord <- getis_ordlep21
GiLEP21 <- ggplot(data=count_lep21) +
  geom_sf(aes(fill=getis_ord)) +
  theme_bw() +
  scale_fill_gradient2(low="#2c7bb6", mid="#ffffbf", high="#d7191c",
                       name="Gi*") +
  labs(title="2021")

#2022
getis_ordlep22 <- localG(count_lep22$n, lw_lep22)
count_lep22$getis_ord <- getis_ordlep22
GiLEP22 <- ggplot(data=count_lep22) +
  geom_sf(aes(fill=getis_ord)) +
  theme_bw() +
  scale_fill_gradient2(low="#2c7bb6", mid="#ffffbf", high="#d7191c",
                       name="Gi*") +
  labs(title="2022")

#2016-2022
getis_ordlep <- localG(count_lep$n, lw_lep)
count_lep$getis_ord <- getis_ordlep
GiLEP <- ggplot(data=count_lep) +
  geom_sf(aes(fill=getis_ord)) +
  theme_bw() +
  scale_fill_gradient2(low="#2c7bb6", mid="#ffffbf", high="#d7191c",
                       name="Gi*") +
  labs(title="2016-2022")
```


plot Getis ord Map for leptospirosis

```{r name-of-chunk28, fig.width=16, fig.height=9}
#| cache = TRUE
grid.arrange(GiLEP16, GiLEP17, GiLEP18, GiLEP19, GiLEP20, GiLEP21, GiLEP22, GiLEP, nrow=2, top=textGrob("Getis Ord Gi* Statistic Map of Leptospirosis Cases in Kelantan 2016-2022", gp = gpar(fontsize = 20)))
```

## Spatial autocorrelation using incidence data

# enteric fever incidence 

Global Moran for enteric fever incidence 

```{r}
#| cache = TRUE
# 2016
inc_ent_kel16 <- count_ent16 %>% 
  mutate(inc_1000 = (n/JUMLAH_2016)*1000)
nb_ent_inc16 <- poly2nb(inc_ent_kel16, queen = TRUE) #set neigbouring queen
nb_ent_inc16 
nb_ent_inc16s <-subset(nb_ent_inc16, subset=card(nb_ent_inc16) > 0)
inc_ent_kel16s <-inc_ent_kel16[-c(13),]
lw_ent_inc16 <- nb2listw(nb_ent_inc16s, style = "W" , zero.policy = TRUE) #assign weight
lep_ent_lag16 <- lag.listw(lw_ent_inc16, inc_ent_kel16s$inc_1000) #create lag function
moran.test(inc_ent_kel16s$inc_1000, lw_ent_inc16)

# 2017
inc_ent_kel17 <- count_ent17 %>% 
  mutate(inc_1000 = (n/JUMLAH_2017)*1000)
nb_ent_inc17 <- poly2nb(inc_ent_kel17, queen = TRUE) #set neigbouring queen
nb_ent_inc17 
nb_ent_inc17s <-subset(nb_ent_inc17, subset=card(nb_ent_inc17) > 0)
inc_ent_kel17s <-inc_ent_kel17[-c(5, 6, 14, 15, 16, 17, 18),]
lw_ent_inc17 <- nb2listw(nb_ent_inc17s, style = "W" , zero.policy = TRUE) #assign weight
lep_ent_lag17 <- lag.listw(lw_ent_inc17, inc_ent_kel17s$inc_1000) #create lag function
moran.test(inc_ent_kel17s$inc_1000, lw_ent_inc17)

# 2018
inc_ent_kel18 <- count_ent18 %>% 
  mutate(inc_1000 = (n/JUMLAH_2018)*1000)
nb_ent_inc18 <- poly2nb(inc_ent_kel18, queen = TRUE) #set neigbouring queen
nb_ent_inc18 
nb_ent_inc18s <-subset(nb_ent_inc18, subset=card(nb_ent_inc18) > 0)
inc_ent_kel18s <-inc_ent_kel18[-c(1, 4, 18),]
lw_ent_inc18 <- nb2listw(nb_ent_inc18s, style = "W" , zero.policy = TRUE) #assign weight
lep_ent_lag18 <- lag.listw(lw_ent_inc18, inc_ent_kel18s$inc_1000) #create lag function
moran.test(inc_ent_kel18s$inc_1000, lw_ent_inc18)

# 2019
inc_ent_kel19 <- count_ent19 %>% 
  mutate(inc_1000 = (n/JUMLAH_2019)*1000)
nb_ent_inc19 <- poly2nb(inc_ent_kel19, queen = TRUE) #set neigbouring queen
nb_ent_inc19 
nb_ent_inc19s <-subset(nb_ent_inc19, subset=card(nb_ent_inc19) > 0)
inc_ent_kel19s <-inc_ent_kel19[-c(4, 9, 10, 11, 12),]
lw_ent_inc19 <- nb2listw(nb_ent_inc19s, style = "W" , zero.policy = TRUE) #assign weight
lep_ent_lag19 <- lag.listw(lw_ent_inc19, inc_ent_kel19s$inc_1000) #create lag function
moran.test(inc_ent_kel19s$inc_1000, lw_ent_inc19)

# 2020
inc_ent_kel20 <- count_ent20 %>% 
  mutate(inc_1000 = (n/JUMLAH_2020)*1000)
nb_ent_inc20 <- poly2nb(inc_ent_kel20, queen = TRUE) #set neigbouring queen
nb_ent_inc20 
nb_ent_inc20s <-subset(nb_ent_inc20, subset=card(nb_ent_inc20) > 0)
inc_ent_kel20s <-inc_ent_kel20[-c(3, 9),]
lw_ent_inc20 <- nb2listw(nb_ent_inc20s, style = "W" , zero.policy = TRUE) #assign weight
lep_ent_lag20 <- lag.listw(lw_ent_inc20, inc_ent_kel20s$inc_1000) #create lag function
moran.test(inc_ent_kel20s$inc_1000, lw_ent_inc20)

# 2021
inc_ent_kel21 <- count_ent21 %>% 
  mutate(inc_1000 = (n/JUMLAH_2021)*1000)
nb_ent_inc21 <- poly2nb(inc_ent_kel21, queen = TRUE) #set neigbouring queen
nb_ent_inc21 
nb_ent_inc21s <-subset(nb_ent_inc21, subset=card(nb_ent_inc21) > 0)
inc_ent_kel21s <-inc_ent_kel21[-c(3),]
lw_ent_inc21 <- nb2listw(nb_ent_inc21s, style = "W" , zero.policy = TRUE) #assign weight
lep_ent_lag21 <- lag.listw(lw_ent_inc21, inc_ent_kel21s$inc_1000) #create lag function
moran.test(inc_ent_kel21s$inc_1000, lw_ent_inc21)

# 2022
inc_ent_kel22 <- count_ent22 %>% 
  mutate(inc_1000 = (n/JUMLAH_2022)*1000)
nb_ent_inc22 <- poly2nb(inc_ent_kel22, queen = TRUE) #set neigbouring queen
nb_ent_inc22 
nb_ent_inc22s <-subset(nb_ent_inc22, subset=card(nb_ent_inc22) > 0)
inc_ent_kel22s <-inc_ent_kel22[-c(1, 3, 8, 10),]
lw_ent_inc22 <- nb2listw(nb_ent_inc22s, style = "W" , zero.policy = TRUE) #assign weight
lep_ent_lag22 <- lag.listw(lw_ent_inc22, inc_ent_kel22s$inc_1000) #create lag function
moran.test(inc_ent_kel22s$inc_1000, lw_ent_inc22)

# 2016-2022
inc_ent_kel <- count_ent %>% 
  mutate(inc_1000 = (n/AVR)*1000)
nb_ent_inc <- poly2nb(inc_ent_kel, queen = TRUE) #set neigbouring queen
lw_ent_inc <- nb2listw(nb_ent_inc, style = "W" , zero.policy = TRUE) #assign weight
lep_ent_lag <- lag.listw(lw_ent_inc, inc_ent_kel$inc_1000) #create lag function
moran.test(inc_ent_kel$inc_1000, lw_ent_inc)
```


# Leptospirosis incidence 

Global Moran for leptospirosis incidence 

```{r}
#| cache = TRUE
# 2016
inc_lep_kel16 <- count_lep16 %>% 
  mutate(inc_1000 = (n/JUMLAH_2016)*1000)
nb_lep_inc16 <- poly2nb(inc_lep_kel16, queen = TRUE) #set neigbouring queen
lw_lep_inc16 <- nb2listw(nb_lep_inc16, style = "W" , zero.policy = TRUE) #assign weight
lep_inc_lag16 <- lag.listw(lw_lep_inc16, inc_lep_kel16$inc_1000) #create lag function
moran.test(inc_lep_kel16$inc_1000, lw_lep_inc16)

# 2017
inc_lep_kel17 <- count_lep17 %>% 
  mutate(inc_1000 = (n/JUMLAH_2017)*1000)
nb_lep_inc17 <- poly2nb(inc_lep_kel17, queen = TRUE) #set neigbouring queen
lw_lep_inc17 <- nb2listw(nb_lep_inc17, style = "W" , zero.policy = TRUE) #assign weight
lep_inc_lag17 <- lag.listw(lw_lep_inc17, inc_lep_kel17$inc_1000) #create lag function
moran.test(inc_lep_kel17$inc_1000, lw_lep_inc17)

# 2018
inc_lep_kel18 <- count_lep18 %>% 
  mutate(inc_1000 = (n/JUMLAH_2018)*1000)
nb_lep_inc18 <- poly2nb(inc_lep_kel18, queen = TRUE) #set neigbouring queen
lw_lep_inc18 <- nb2listw(nb_lep_inc18, style = "W" , zero.policy = TRUE) #assign weight
lep_inc_lag18 <- lag.listw(lw_lep_inc18, inc_lep_kel18$inc_1000) #create lag function
moran.test(inc_lep_kel18$inc_1000, lw_lep_inc18)

# 2019
inc_lep_kel19 <- count_lep19 %>% 
  mutate(inc_1000 = (n/JUMLAH_2019)*1000)
nb_lep_inc19 <- poly2nb(inc_lep_kel19, queen = TRUE) #set neigbouring queen
lw_lep_inc19 <- nb2listw(nb_lep_inc19, style = "W" , zero.policy = TRUE) #assign weight
lep_inc_lag19 <- lag.listw(lw_lep_inc19, inc_lep_kel19$inc_1000) #create lag function
moran.test(inc_lep_kel19$inc_1000, lw_lep_inc19)

# 2020
inc_lep_kel20 <- count_lep20 %>% 
  mutate(inc_1000 = (n/JUMLAH_2020)*1000)
nb_lep_inc20 <- poly2nb(inc_lep_kel20, queen = TRUE) #set neigbouring queen
lw_lep_inc20 <- nb2listw(nb_lep_inc20, style = "W" , zero.policy = TRUE) #assign weight
lep_inc_lag20 <- lag.listw(lw_lep_inc20, inc_lep_kel20$inc_1000) #create lag function
moran.test(inc_lep_kel20$inc_1000, lw_lep_inc20)

# 2021
inc_lep_kel21 <- count_lep21 %>% 
  mutate(inc_1000 = (n/JUMLAH_2021)*1000)
nb_lep_inc21 <- poly2nb(inc_lep_kel21, queen = TRUE) #set neigbouring queen
lw_lep_inc21 <- nb2listw(nb_lep_inc21, style = "W" , zero.policy = TRUE) #assign weight
lep_inc_lag21 <- lag.listw(lw_lep_inc21, inc_lep_kel21$inc_1000) #create lag function
moran.test(inc_lep_kel21$inc_1000, lw_lep_inc21)

# 2022
inc_lep_kel22 <- count_lep22 %>% 
  mutate(inc_1000 = (n/JUMLAH_2022)*1000)
nb_lep_inc22 <- poly2nb(inc_lep_kel22, queen = TRUE) #set neigbouring queen
lw_lep_inc22 <- nb2listw(nb_lep_inc22, style = "W" , zero.policy = TRUE) #assign weight
lep_inc_lag22 <- lag.listw(lw_lep_inc22, inc_lep_kel22$inc_1000) #create lag function
moran.test(inc_lep_kel22$inc_1000, lw_lep_inc22)

# 2016-2022
inc_lep_kel <- count_lep %>% 
  mutate(inc_1000 = (n/AVR)*1000)
nb_lep_inc <- poly2nb(inc_lep_kel, queen = TRUE) #set neigbouring queen
lw_lep_inc <- nb2listw(nb_lep_inc, style = "W" , zero.policy = TRUE) #assign weight
lep_inc_lag <- lag.listw(lw_lep_inc, inc_lep_kel$inc_1000) #create lag function
moran.test(inc_lep_kel$inc_1000, lw_lep_inc)
```


local Moran for Leptospirosis incidence

```{r}
#| cache = TRUE
#2016
Local_Moran_lep16inc <- localmoran(inc_lep_kel16$inc_1000,lw_lep_inc16)
LEP16inc_poly <- cbind(inc_lep_kel16, Local_Moran_lep16inc)
#plot Local Moran
plotMlep16inc <- ggplot(data = LEP16inc_poly) +
  geom_sf(aes(fill=Ii)) +
  theme_bw() +
  scale_fill_gradient2(low="#2c7bb6", mid="#ffffbf", high="#d7191c",
                       name="Local Moran's I") +
  labs(title="2016")
plot(plotMlep16inc) 

#2017
Local_Moran_lep17inc <- localmoran(inc_lep_kel17$inc_1000,lw_lep_inc17)
LEP17inc_poly <- cbind(inc_lep_kel17, Local_Moran_lep17inc)
#plot Local Moran
plotMlep17inc <- ggplot(data = LEP17inc_poly) +
  geom_sf(aes(fill=Ii)) +
  theme_bw() +
  scale_fill_gradient2(low="#2c7bb6", mid="#ffffbf", high="#d7191c",
                       name="Local Moran's I") +
  labs(title="2017")
plot(plotMlep17inc) 

#2018
Local_Moran_lep18inc <- localmoran(inc_lep_kel18$inc_1000,lw_lep_inc18)
LEP18inc_poly <- cbind(inc_lep_kel18, Local_Moran_lep18inc)
#plot Local Moran
plotMlep18inc <- ggplot(data = LEP18inc_poly) +
  geom_sf(aes(fill=Ii)) +
  theme_bw() +
  scale_fill_gradient2(low="#2c7bb6", mid="#ffffbf", high="#d7191c",
                       name="Local Moran's I") +
  labs(title="2018")
plot(plotMlep18inc) 

#2019
Local_Moran_lep19inc <- localmoran(inc_lep_kel19$inc_1000,lw_lep_inc19)
LEP19inc_poly <- cbind(inc_lep_kel19, Local_Moran_lep19inc)
#plot Local Moran
plotMlep19inc <- ggplot(data = LEP19inc_poly) +
  geom_sf(aes(fill=Ii)) +
  theme_bw() +
  scale_fill_gradient2(low="#2c7bb6", mid="#ffffbf", high="#d7191c",
                       name="Local Moran's I") +
  labs(title="2019")
plot(plotMlep19inc)

#2020
Local_Moran_lep20inc <- localmoran(inc_lep_kel20$inc_1000,lw_lep_inc20)
LEP20inc_poly <- cbind(inc_lep_kel20, Local_Moran_lep20inc)
#plot Local Moran
plotMlep20inc <- ggplot(data = LEP20inc_poly) +
  geom_sf(aes(fill=Ii)) +
  theme_bw() +
  scale_fill_gradient2(low="#2c7bb6", mid="#ffffbf", high="#d7191c",
                       name="Local Moran's I") +
  labs(title="2020")
plot(plotMlep20inc)

#2021
Local_Moran_lep21inc <- localmoran(inc_lep_kel21$inc_1000,lw_lep_inc21)
LEP21inc_poly <- cbind(inc_lep_kel21, Local_Moran_lep21inc)
#plot Local Moran
plotMlep21inc <- ggplot(data = LEP21inc_poly) +
  geom_sf(aes(fill=Ii)) +
  theme_bw() +
  scale_fill_gradient2(low="#2c7bb6", mid="#ffffbf", high="#d7191c",
                       name="Local Moran's I") +
  labs(title="2021")
plot(plotMlep21inc)

#2022
Local_Moran_lep22inc <- localmoran(inc_lep_kel22$inc_1000,lw_lep_inc22)
LEP22inc_poly <- cbind(inc_lep_kel22, Local_Moran_lep22inc)
#plot Local Moran
plotMlep22inc <- ggplot(data = LEP22inc_poly) +
  geom_sf(aes(fill=Ii)) +
  theme_bw() +
  scale_fill_gradient2(low="#2c7bb6", mid="#ffffbf", high="#d7191c",
                       name="Local Moran's I") +
  labs(title="2022")
plot(plotMlep22inc)

#2016-2022
Local_Moran_lepinc <- localmoran(inc_lep_kel$inc_1000,lw_lep_inc)
LEPinc_poly <- cbind(inc_lep_kel, Local_Moran_lepinc)
#plot Local Moran
plotMlepinc <- ggplot(data = LEPinc_poly) +
  geom_sf(aes(fill=Ii)) +
  theme_bw() +
  scale_fill_gradient2(low="#2c7bb6", mid="#ffffbf", high="#d7191c",
                       name="Local Moran's I") +
  labs(title="2016-2020")
plot(plotMlepinc)
```


Plot Local Moran for Leptospirosis incidence
- 2020-2021 excluded (not significant)

```{r name-of-chunk29, fig.width=16, fig.height=9}
#| cache = TRUE
grid.arrange(plotMlep16inc, plotMlep17inc, plotMlep18inc, plotMlep19inc, plotMlep22inc, plotMlepinc,  nrow=2, top=textGrob("Local Moran's I Map of Leptospirosis Incidence in Kelantan 2016-2022", gp = gpar(fontsize = 20)))
```


Local Moran's I p-values leptospirosis incidence

```{r}
#| cache = TRUE
LMp16i <- tm_shape(LEP16inc_poly) +
  tm_fill(col = "Pr.z....E.Ii..",
          breaks=c(-Inf, 0.001, 0.01, 0.05, 0.1, Inf),
          palette = "-Blues",
          title = "2016")+
  tm_borders(alpha = 0.5)

LMp17i <- tm_shape(LEP17inc_poly) +
  tm_fill(col = "Pr.z....E.Ii..",
          breaks=c(-Inf, 0.001, 0.01, 0.05, 0.1, Inf),
          palette = "-Blues",
          title = "2017")+
  tm_borders(alpha = 0.5)

LMp18i <- tm_shape(LEP18inc_poly) +
  tm_fill(col = "Pr.z....E.Ii..",
          breaks=c(-Inf, 0.001, 0.01, 0.05, 0.1, Inf),
          palette = "-Blues",
          title = "2018")+
  tm_borders(alpha = 0.5)

LMp19i <- tm_shape(LEP19inc_poly) +
  tm_fill(col = "Pr.z....E.Ii..",
          breaks=c(-Inf, 0.001, 0.01, 0.05, 0.1, Inf),
          palette = "-Blues",
          title = "2019")+
  tm_borders(alpha = 0.5)

LMp22i <- tm_shape(LEP22inc_poly) +
  tm_fill(col = "Pr.z....E.Ii..",
          breaks=c(-Inf, 0.001, 0.01, 0.05, 0.1, Inf),
          palette = "-Blues",
          title = "2022")+
  tm_borders(alpha = 0.5)

LMpi <- tm_shape(LEP_poly) +
  tm_fill(col = "Pr.z....E.Ii..",
          breaks=c(-Inf, 0.001, 0.01, 0.05, 0.1, Inf),
          palette = "-Blues",
          title = "2016-2022")+
  tm_borders(alpha = 0.5)
```

plot Local Moran's I p-values leptospirosis incidence

```{r name-of-chunk30, fig.width=16, fig.height=9}
#| cache = TRUE
# combine all local moran pvalue plot for leptospirosis  in one view
tmap_arrange(LMp16i, LMp17i, LMp18i, LMp19i, LMp22i, LMpi, ncol = 3) 
```


LISA map for leptospirosis incidence

```{r}
#| cache = TRUE
#LISA 2016-2022 leptospirosis incidence
m_lepinc <- inc_lep_kel$inc_1000 - mean(inc_lep_kel$inc_1000) #Center the variable of interest around its mean
m_local_lepinc <- Local_Moran_lepinc[,1]- mean(Local_Moran_lepinc[,1]) #Center the local Moran around the mean
signif <- 0.05 #significant threshold
quadrant_lepinc <- vector(mode="numeric",length = nrow(Local_Moran_lepinc))
#set classification - built data quadrant
quadrant_lep [m_lepinc>0 & m_local_lepinc>0] <- 4
quadrant_lep [m_lepinc<0 & m_local_lepinc<0] <- 1
quadrant_lep [m_lepinc<0 & m_local_lepinc>0] <- 2
quadrant_lep [m_lepinc>0 & m_local_lepinc<0] <- 3
quadrant_lep [Local_Moran_lepinc[,5]] <- 0
#plot LISA
brks <- c(0,1,2,3,4)
colors <- c("white", "blue",rgb(0,0,1,alpha=0.4),rgb(1,0,0,alpha = 0.4),"red")
plot(inc_lep_kel[,5],main = "LISA Map of Leptospirosis Incidence 2016-2022", las = 1, border="lightgray",col=colors[findInterval(quadrant_lep,brks,all.inside = FALSE)])
legend("bottomright",legend = c("insignificant","low-low","low-high","high-low","high-high"),
       fill=colors,bty="n")

#LISA 2016leptospirosis incidence
m_lepinc16 <- inc_lep_kel16$inc_1000 - mean(inc_lep_kel16$inc_1000) 
m_local_lepinc16 <- Local_Moran_lep16inc[,1]- mean(Local_Moran_lep16inc[,1]) 
signif <- 0.05 
quadrant_lep16inc <- vector(mode="numeric",length = nrow(Local_Moran_lep16inc))
quadrant_lep [m_lepinc16>0 & m_local_lepinc16>0] <- 4
quadrant_lep [m_lepinc16<0 & m_local_lepinc16<0] <- 1
quadrant_lep [m_lepinc16<0 & m_local_lepinc16>0] <- 2
quadrant_lep [m_lepinc16>0 & m_local_lepinc16<0] <- 3
quadrant_lep [Local_Moran_lep16inc[,5]] <- 0
brks <- c(0,1,2,3,4)
colors <- c("white", "blue",rgb(0,0,1,alpha=0.4),rgb(1,0,0,alpha = 0.4),"red")
plot(inc_lep_kel16[,5],main = "LISA Map of Leptospirosis Incidence 2016", las = 1, border="lightgray",col=colors[findInterval(quadrant_lep,brks,all.inside = FALSE)])
legend("bottomright",legend = c("insignificant","low-low","low-high","high-low","high-high"),
       fill=colors,bty="n")

#LISA 2017leptospirosis incidence
m_lepinc17 <- inc_lep_kel17$inc_1000 - mean(inc_lep_kel17$inc_1000) 
m_local_lepinc17 <- Local_Moran_lep17inc[,1]- mean(Local_Moran_lep17inc[,1]) 
signif <- 0.05 
quadrant_lep17inc <- vector(mode="numeric",length = nrow(Local_Moran_lep17inc))
quadrant_lep [m_lepinc17>0 & m_local_lepinc17>0] <- 4
quadrant_lep [m_lepinc17<0 & m_local_lepinc17<0] <- 1
quadrant_lep [m_lepinc17<0 & m_local_lepinc17>0] <- 2
quadrant_lep [m_lepinc17>0 & m_local_lepinc17<0] <- 3
quadrant_lep [Local_Moran_lep17inc[,5]] <- 0
brks <- c(0,1,2,3,4)
colors <- c("white", "blue",rgb(0,0,1,alpha=0.4),rgb(1,0,0,alpha = 0.4),"red")
plot(inc_lep_kel17[,5],main = "LISA Map of Leptospirosis Incidence 2017", las = 1, border="lightgray",col=colors[findInterval(quadrant_lep,brks,all.inside = FALSE)])
legend("bottomright",legend = c("insignificant","low-low","low-high","high-low","high-high"),
       fill=colors,bty="n")

#LISA 2018leptospirosis incidence
m_lepinc18 <- inc_lep_kel18$inc_1000 - mean(inc_lep_kel18$inc_1000) 
m_local_lepinc18 <- Local_Moran_lep18inc[,1]- mean(Local_Moran_lep18inc[,1]) 
signif <- 0.05 
quadrant_lep18inc <- vector(mode="numeric",length = nrow(Local_Moran_lep18inc))
quadrant_lep [m_lepinc18>0 & m_local_lepinc18>0] <- 4
quadrant_lep [m_lepinc18<0 & m_local_lepinc18<0] <- 1
quadrant_lep [m_lepinc18<0 & m_local_lepinc18>0] <- 2
quadrant_lep [m_lepinc18>0 & m_local_lepinc18<0] <- 3
quadrant_lep [Local_Moran_lep18inc[,5]] <- 0
brks <- c(0,1,2,3,4)
colors <- c("white", "blue",rgb(0,0,1,alpha=0.4),rgb(1,0,0,alpha = 0.4),"red")
plot(inc_lep_kel18[,5],main = "LISA Map of Leptospirosis Incidence 2018", las = 1, border="lightgray",col=colors[findInterval(quadrant_lep,brks,all.inside = FALSE)])
legend("bottomright",legend = c("insignificant","low-low","low-high","high-low","high-high"),
       fill=colors,bty="n")

#LISA 2019leptospirosis incidence
m_lepinc19 <- inc_lep_kel19$inc_1000 - mean(inc_lep_kel19$inc_1000) 
m_local_lepinc19 <- Local_Moran_lep19inc[,1]- mean(Local_Moran_lep19inc[,1]) 
signif <- 0.05 
quadrant_lep19inc <- vector(mode="numeric",length = nrow(Local_Moran_lep19inc))
quadrant_lep [m_lepinc19>0 & m_local_lepinc19>0] <- 4
quadrant_lep [m_lepinc19<0 & m_local_lepinc19<0] <- 1
quadrant_lep [m_lepinc19<0 & m_local_lepinc19>0] <- 2
quadrant_lep [m_lepinc19>0 & m_local_lepinc19<0] <- 3
quadrant_lep [Local_Moran_lep19inc[,5]] <- 0
brks <- c(0,1,2,3,4)
colors <- c("white", "blue",rgb(0,0,1,alpha=0.4),rgb(1,0,0,alpha = 0.4),"red")
plot(inc_lep_kel19[,5],main = "LISA Map of Leptospirosis Incidence 2019", las = 1, border="lightgray",col=colors[findInterval(quadrant_lep,brks,all.inside = FALSE)])
legend("bottomright",legend = c("insignificant","low-low","low-high","high-low","high-high"),
       fill=colors,bty="n")

#LISA 2022leptospirosis incidence
m_lepinc22 <- inc_lep_kel22$inc_1000 - mean(inc_lep_kel22$inc_1000) 
m_local_lepinc22 <- Local_Moran_lep22inc[,1]- mean(Local_Moran_lep22inc[,1]) 
signif <- 0.05 
quadrant_lep22inc <- vector(mode="numeric",length = nrow(Local_Moran_lep22inc))
quadrant_lep [m_lepinc22>0 & m_local_lepinc22>0] <- 4
quadrant_lep [m_lepinc22<0 & m_local_lepinc22<0] <- 1
quadrant_lep [m_lepinc22<0 & m_local_lepinc22>0] <- 2
quadrant_lep [m_lepinc22>0 & m_local_lepinc22<0] <- 3
quadrant_lep [Local_Moran_lep22inc[,5]] <- 0
brks <- c(0,1,2,3,4)
colors <- c("white", "blue",rgb(0,0,1,alpha=0.4),rgb(1,0,0,alpha = 0.4),"red")
plot(inc_lep_kel22[,5],main = "LISA Map of Leptospirosis Incidence 2022", las = 1, border="lightgray",col=colors[findInterval(quadrant_lep,brks,all.inside = FALSE)])
legend("bottomright",legend = c("insignificant","low-low","low-high","high-low","high-high"),
       fill=colors,bty="n")
```

# corelation between cases abd population density

spatial relation of enteric fever cases to population by Lee's L test
- if not significant, the pattern of cases and population are not related to one another, the spatial pattern of cases is not strictly a result of population density in high-risk areas

```{r}
#| cache = TRUE
# 2016-2022
lee_testent <- lee.test(x=count_ent$n, y=count_ent$AVR, listw=lw_ent)            
lee_testent

# 2016
lee_testent16 <- lee.test(x=count_ent16s$n, y=count_ent16s$JUMLAH_2016, listw=lw_ent16)
lee_testent16

# 2017
lee_testent17 <- lee.test(x=count_ent17s$n, y=count_ent17s$JUMLAH_2017, listw=lw_ent17)     
lee_testent17

# 2018
lee_testent18 <- lee.test(x=count_ent18s$n, y=count_ent18s$JUMLAH_2018, listw=lw_ent18)      
lee_testent18

# 2019
lee_testent19 <- lee.test(x=count_ent19s$n, y=count_ent19s$JUMLAH_2019, listw=lw_ent19)     
lee_testent19

# 2020
lee_testent20 <- lee.test(x=count_ent20s$n, y=count_ent20s$JUMLAH_2020, listw=lw_ent20)     
lee_testent20

# 2021
lee_testent21 <- lee.test(x=count_ent21s$n, y=count_ent21s$JUMLAH_2021, listw=lw_ent21)     
lee_testent21

# 2022
lee_testent22 <- lee.test(x=count_ent22s$n, y=count_ent22s$JUMLAH_2022, listw=lw_ent22)     
lee_testent22

```

spatial relation of leptospirosis cases to population by Lee's L test
- if not significant, the pattern of cases and population are not related to one another, the spatial pattern of cases is not strictly a result of population density in high-risk areas

```{r}
#| cache = TRUE
# 2016-2022
lee_testlep <- lee.test(x=inc_lep_kel$n, y=inc_lep_kel$AVR, listw=lw_lep)            
lee_testlep

# 2016
lee_testlep16 <- lee.test(x=inc_lep_kel16$n, y=inc_lep_kel16$JUMLAH_2016, listw=lw_lep16)
lee_testlep16

# 2017
lee_testlep17 <- lee.test(x=inc_lep_kel17$n, y=inc_lep_kel17$JUMLAH_2017, listw=lw_lep17)
lee_testlep17

# 2018
lee_testlep18 <- lee.test(x=inc_lep_kel18$n, y=inc_lep_kel18$JUMLAH_2018, listw=lw_lep18)
lee_testlep18

# 2019
lee_testlep19 <- lee.test(x=inc_lep_kel19$n, y=inc_lep_kel19$JUMLAH_2019, listw=lw_lep19)
lee_testlep19

# 2020
lee_testlep20 <- lee.test(x=inc_lep_kel20$n, y=inc_lep_kel20$JUMLAH_2020, listw=lw_lep20)
lee_testlep20

# 2021
lee_testlep21 <- lee.test(x=inc_lep_kel21$n, y=inc_lep_kel21$JUMLAH_2021, listw=lw_lep21)
lee_testlep21

# 2022
lee_testlep22 <- lee.test(x=inc_lep_kel22$n, y=inc_lep_kel22$JUMLAH_2022, listw=lw_lep22)
lee_testlep22
```

## comparison between leptospirosis and enteric fever (bivariate analysis)

# prepare data

mark data with diagnosis (factor)

```{r}
#| cache = TRUE
# transform to sp
loc_both <- st_as_sf(linelist2, 
                     coords = c("Longitude (WGS)", "Latitude (WGS)"), 
                     crs = 4326)
loc_both <-st_transform(loc_both, 3168)
loc_both.sp <- as(loc_both, "Spatial")

# transform data sp to ppp - save categorical data as factor
all_diag.ppp <- as(loc_both.sp, "ppp")
marks(all_diag.ppp) <- as.factor(loc_both.sp$Diagnosis)
Window(all_diag.ppp) <- kel_map.owin
```

```{r}
#summary
summary(all_diag.ppp)
```


plot split data

```{r}
#| cache = TRUE
plot(split(all_diag.ppp))
plot(density(split(all_diag.ppp)), ribbon = FALSE)
```

# spatial relative risk (sparr package)

cumulative cases

```{r}
#| cache = TRUE
# split data
diag_lep <- split(all_diag.ppp)$Leptospirosis
diag_ent <- split(all_diag.ppp)$Enteric_fever

# Compute a symmetric (pooled) adaptive relative risk estimate
# with tolerance contours
h0 <- OS(all_diag.ppp, nstar="geometric")
diag_rr <- risk(diag_lep, diag_ent, h0=h0, adapt=TRUE, tolerate=TRUE,
                hp=OS(all_diag.ppp)/2, pilot.symmetry="pooled", davies.baddeley=0.05)

plot(diag_rr)
```

prepare file for spatial relative risk by year

```{r}
#| cache = TRUE
# extract list contain both diagnosis by year
list2016 <- linelist2 %>% 
  filter(`Tahun Daftar` == "2016")
list2017 <- linelist2 %>% 
  filter(`Tahun Daftar` == "2017")
list2018 <- linelist2 %>% 
  filter(`Tahun Daftar` == "2018")
list2019 <- linelist2 %>% 
  filter(`Tahun Daftar` == "2019")
list2020 <- linelist2 %>% 
  filter(`Tahun Daftar` == "2020")
list2021 <- linelist2 %>% 
  filter(`Tahun Daftar` == "2021")
list2022 <- linelist2 %>% 
  filter(`Tahun Daftar` == "2022")

# transform to sf
list2016.sf <- st_as_sf(list2017, 
                     coords = c("Longitude (WGS)", "Latitude (WGS)"), 
                     crs = 4326)
list2017.sf <- st_as_sf(list2017, 
                     coords = c("Longitude (WGS)", "Latitude (WGS)"), 
                     crs = 4326)
list2018.sf <- st_as_sf(list2018, 
                     coords = c("Longitude (WGS)", "Latitude (WGS)"), 
                     crs = 4326)
list2019.sf <- st_as_sf(list2019, 
                     coords = c("Longitude (WGS)", "Latitude (WGS)"), 
                     crs = 4326)
list2020.sf <- st_as_sf(list2020, 
                     coords = c("Longitude (WGS)", "Latitude (WGS)"), 
                     crs = 4326)
list2021.sf <- st_as_sf(list2021, 
                     coords = c("Longitude (WGS)", "Latitude (WGS)"), 
                     crs = 4326)
list2022.sf <- st_as_sf(list2022, 
                     coords = c("Longitude (WGS)", "Latitude (WGS)"), 
                     crs = 4326)

# transform to RSO
loc_list2016.sf <-st_transform(list2016.sf, 3168)
loc_list2017.sf <-st_transform(list2017.sf, 3168)
loc_list2018.sf <-st_transform(list2018.sf, 3168)
loc_list2019.sf <-st_transform(list2019.sf, 3168)
loc_list2020.sf <-st_transform(list2020.sf, 3168)
loc_list2021.sf <-st_transform(list2021.sf, 3168)
loc_list2022.sf <-st_transform(list2022.sf, 3168)

# transform to sp
loc_list2016.sp <- as(loc_list2016.sf, "Spatial")
loc_list2017.sp <- as(loc_list2017.sf, "Spatial")
loc_list2018.sp <- as(loc_list2018.sf, "Spatial")
loc_list2019.sp <- as(loc_list2019.sf, "Spatial")
loc_list2020.sp <- as(loc_list2020.sf, "Spatial")
loc_list2021.sp <- as(loc_list2021.sf, "Spatial")
loc_list2022.sp <- as(loc_list2022.sf, "Spatial")

# transform data sp to ppp 
loc_list2016.ppp <- as(loc_list2016.sp, "ppp")
loc_list2017.ppp <- as(loc_list2017.sp, "ppp")
loc_list2018.ppp <- as(loc_list2018.sp, "ppp")
loc_list2019.ppp <- as(loc_list2019.sp, "ppp")
loc_list2020.ppp <- as(loc_list2020.sp, "ppp")
loc_list2021.ppp <- as(loc_list2021.sp, "ppp")
loc_list2022.ppp <- as(loc_list2022.sp, "ppp")

# save categorical data as factor (marked data)
marks(loc_list2016.ppp) <- as.factor(loc_list2016.sp$Diagnosis)
marks(loc_list2017.ppp) <- as.factor(loc_list2017.sp$Diagnosis)
marks(loc_list2018.ppp) <- as.factor(loc_list2018.sp$Diagnosis)
marks(loc_list2019.ppp) <- as.factor(loc_list2019.sp$Diagnosis)
marks(loc_list2020.ppp) <- as.factor(loc_list2020.sp$Diagnosis)
marks(loc_list2021.ppp) <- as.factor(loc_list2021.sp$Diagnosis)
marks(loc_list2022.ppp) <- as.factor(loc_list2022.sp$Diagnosis)

# set window
Window(loc_list2016.ppp) <- kel_map.owin
Window(loc_list2017.ppp) <- kel_map.owin
Window(loc_list2018.ppp) <- kel_map.owin
Window(loc_list2019.ppp) <- kel_map.owin
Window(loc_list2020.ppp) <- kel_map.owin
Window(loc_list2021.ppp) <- kel_map.owin
Window(loc_list2022.ppp) <- kel_map.owin
```


calculate spatial relative risk by year

```{r}
#| cache = TRUE
# 2016
# split into diagnosis
diag_lep16 <- split(loc_list2016.ppp)$Leptospirosis
diag_ent16 <- split(loc_list2016.ppp)$Enteric_fever

# Compute a symmetric (pooled) adaptive relative risk estimate
# with tolerance contours
h0_16 <- OS(loc_list2016.ppp, nstar="geometric")
diag_rr16 <- risk(diag_lep16, diag_ent16, h0=h0_16, adapt=TRUE, tolerate=TRUE,
                hp=OS(loc_list2016.ppp)/2, pilot.symmetry="pooled", davies.baddeley=0.05)

# 2017
diag_lep17 <- split(loc_list2017.ppp)$Leptospirosis
diag_ent17 <- split(loc_list2017.ppp)$Enteric_fever
h0_17 <- OS(loc_list2017.ppp, nstar="geometric")
diag_rr17 <- risk(diag_lep17, diag_ent17, h0=h0_17, adapt=TRUE, tolerate=TRUE,
                hp=OS(loc_list2017.ppp)/2, pilot.symmetry="pooled", davies.baddeley=0.05)

# 2018
diag_lep18 <- split(loc_list2018.ppp)$Leptospirosis
diag_ent18 <- split(loc_list2018.ppp)$Enteric_fever
h0_18 <- OS(loc_list2018.ppp, nstar="geometric")
diag_rr18 <- risk(diag_lep18, diag_ent18, h0=h0_18, adapt=TRUE, tolerate=TRUE,
                hp=OS(loc_list2018.ppp)/2, pilot.symmetry="pooled", davies.baddeley=0.05)

# 2019
diag_lep19 <- split(loc_list2019.ppp)$Leptospirosis
diag_ent19 <- split(loc_list2019.ppp)$Enteric_fever
h0_19 <- OS(loc_list2019.ppp, nstar="geometric")
diag_rr19 <- risk(diag_lep19, diag_ent19, h0=h0_19, adapt=TRUE, tolerate=TRUE,
                hp=OS(loc_list2019.ppp)/2, pilot.symmetry="pooled", davies.baddeley=0.05)

# 2020
diag_lep20 <- split(loc_list2020.ppp)$Leptospirosis
diag_ent20 <- split(loc_list2020.ppp)$Enteric_fever
h0_20 <- OS(loc_list2020.ppp, nstar="geometric")
diag_rr20 <- risk(diag_lep20, diag_ent20, h0=h0_20, adapt=TRUE, tolerate=TRUE,
                hp=OS(loc_list2020.ppp)/2, pilot.symmetry="pooled", davies.baddeley=0.05)

# 2021
diag_lep21 <- split(loc_list2021.ppp)$Leptospirosis
diag_ent21 <- split(loc_list2021.ppp)$Enteric_fever
h0_21 <- OS(loc_list2021.ppp, nstar="geometric")
diag_rr21 <- risk(diag_lep21, diag_ent21, h0=h0_21, adapt=TRUE, tolerate=TRUE,
                hp=OS(loc_list2021.ppp)/2, pilot.symmetry="pooled", davies.baddeley=0.05)

# 2022
diag_lep22 <- split(loc_list2022.ppp)$Leptospirosis
diag_ent22 <- split(loc_list2022.ppp)$Enteric_fever
h0_22 <- OS(loc_list2022.ppp, nstar="geometric")
diag_rr22 <- risk(diag_lep22, diag_ent22, h0=h0_22, adapt=TRUE, tolerate=TRUE,
                hp=OS(loc_list2022.ppp)/2, pilot.symmetry="pooled", davies.baddeley=0.05)
```


plot spatial relative risk by year

```{r name-of-chunk32, fig.width=20, fig.height=9}
#| cache = TRUE
par( mfrow= c(2,4) )
plot(diag_rr16, main = "2016", cex.main = 3, cex.lab = 1.5)
plot(diag_rr17, main = "2017", cex.main = 3, cex.lab = 1.5)
plot(diag_rr18, main = "2018", cex.main = 3, cex.lab = 1.5)
plot(diag_rr19, main = "2019", cex.main = 3, cex.lab = 1.5)
plot(diag_rr20, main = "2020", cex.main = 3, cex.lab = 1.5)
plot(diag_rr21, main = "2021", cex.main = 3, cex.lab = 1.5)
plot(diag_rr22, main = "2022", cex.main = 3, cex.lab = 1.5)
plot(diag_rr, main = "2016-2022", cex.main = 3, cex.lab = 1.5)
```


# bivariate L function 

calculate bivariate L function 

```{r}
#| cache = TRUE
Lboth <- alltypes(all_diag.ppp, "L")
```

```{r}
#| cache = TRUE
Lboth16 <- alltypes(loc_list2016.ppp, "L")
Lboth17 <- alltypes(loc_list2017.ppp, "L")
Lboth18 <- alltypes(loc_list2018.ppp, "L")
Lboth19 <- alltypes(loc_list2019.ppp, "L")
Lboth20 <- alltypes(loc_list2020.ppp, "L")
Lboth21 <- alltypes(loc_list2021.ppp, "L")
Lboth22 <- alltypes(loc_list2022.ppp, "L")
```


plot bivariate L graph by year

```{r name-of-chunk33, fig.width=20, fig.height=9}
plot(Lboth16, main = "2016", cex.main = 3, cex.lab = 1.5)
plot(Lboth17, main = "2017", cex.main = 3, cex.lab = 1.5)
plot(Lboth18, main = "2018", cex.main = 3, cex.lab = 1.5)
plot(Lboth19, main = "2019", cex.main = 3, cex.lab = 1.5)
plot(Lboth20, main = "2020", cex.main = 3, cex.lab = 1.5)
plot(Lboth21, main = "2021", cex.main = 3, cex.lab = 1.5)
plot(Lboth22, main = "2022", cex.main = 3, cex.lab = 1.5)
plot(Lboth, main = "2016-2022", cex.main = 3, cex.lab = 1.5)
```


```{r}
#| cache = TRUE
#plot estimates of the bivariate pair correlation functions
Lbothest <- plot(alltypes(all_diag.ppp, pcfcross))
```


```{r}
#| cache = TRUE
# bivariate G-functions
Gboth16 <- alltypes(loc_list2016.ppp, Gcross)
Gboth17 <- alltypes(loc_list2017.ppp, Gcross)
Gboth18 <- alltypes(loc_list2018.ppp, Gcross)
Gboth19 <- alltypes(loc_list2019.ppp, Gcross)
Gboth20 <- alltypes(loc_list2020.ppp, Gcross)
Gboth21 <- alltypes(loc_list2021.ppp, Gcross)
Gboth22 <- alltypes(loc_list2022.ppp, Gcross)
Gboth <- alltypes(all_diag.ppp, Gcross)
```


```{r name-of-chunk, fig.width=20, fig.height=9}
# plot bivariate G-functions
plot(Gboth16)
plot(Gboth17)
plot(Gboth18)
plot(Gboth19)
plot(Gboth20)
plot(Gboth21)
plot(Gboth22)
plot(Gboth)
```


# calculate Gdot function

```{r}
#| cache = TRUE
Gdotall <- alltypes(all_diag.ppp, Gdot)
Gdot16 <- alltypes(loc_list2016.ppp, Gdot)
Gdot17 <- alltypes(loc_list2017.ppp, Gdot)
Gdot18 <- alltypes(loc_list2018.ppp, Gdot)
Gdot19 <- alltypes(loc_list2019.ppp, Gdot)
Gdot20 <- alltypes(loc_list2020.ppp, Gdot)
Gdot21 <- alltypes(loc_list2021.ppp, Gdot)
Gdot22 <- alltypes(loc_list2022.ppp, Gdot)

```


# plot Gdot

```{r}
plot(Gdot16)
plot(Gdot17)
plot(Gdot18)
plot(Gdot19)
plot(Gdot20)
plot(Gdot21)
plot(Gdot22)
plot(Gdotall)
```


# Ripley's K cross function

calculate K cross

```{r}
#| cache = TRUE
K16 <- Kcross(loc_list2016.ppp, "Leptospirosis", "Enteric_fever") 
K17 <- Kcross(loc_list2017.ppp, "Leptospirosis", "Enteric_fever") 
K18 <- Kcross(loc_list2018.ppp, "Leptospirosis", "Enteric_fever") 
K19 <- Kcross(loc_list2019.ppp, "Leptospirosis", "Enteric_fever") 
K20 <- Kcross(loc_list2020.ppp, "Leptospirosis", "Enteric_fever") 
K21 <- Kcross(loc_list2021.ppp, "Leptospirosis", "Enteric_fever") 
K22 <- Kcross(loc_list2022.ppp, "Leptospirosis", "Enteric_fever") 
Kall <- Kcross(all_diag.ppp, "Leptospirosis", "Enteric_fever") 

```


plot Ripley's cross K function
- Departures from the reference line indicate clustering of the two types of points relative to each other
- When the observed K value is larger than the expected K value for a particular distance, the distribution is more clustered than a random distribution at that distance (scale of analysis). When the observed K value is smaller than the expected K value, the distribution is more dispersed than a random distribution at that distance

```{r name-of-chunk34, fig.width=20, fig.height=9}
par(mfrow=c(2,4))
plot(K16, main = 2016, cex.main = 3, cex.lab = 1.5)
plot(K17, main = 2017, cex.main = 3, cex.lab = 1.5)
plot(K18, main = 2018, cex.main = 3, cex.lab = 1.5)
plot(K19, main = 2019, cex.main = 3, cex.lab = 1.5)
plot(K20, main = 2020, cex.main = 3, cex.lab = 1.5)
plot(K21, main = 2021, cex.main = 3, cex.lab = 1.5)
plot(K22, main = 2022, cex.main = 3, cex.lab = 1.5)
plot(Kall, main = 2016-2022, cex.main = 3, cex.lab = 1.5)
```

# Kdot

Calculate Kdot

```{r}
#| cache = TRUE
Kdot16 <- Kdot(loc_list2016.ppp, "Leptospirosis")
Kdot17 <- Kdot(loc_list2017.ppp, "Leptospirosis")
Kdot18 <- Kdot(loc_list2018.ppp, "Leptospirosis")
Kdot19 <- Kdot(loc_list2019.ppp, "Leptospirosis")
Kdot20 <- Kdot(loc_list2020.ppp, "Leptospirosis")
Kdot21 <- Kdot(loc_list2021.ppp, "Leptospirosis")
Kdot22 <- Kdot(loc_list2022.ppp, "Leptospirosis")
Kdotall <- Kdot(all_diag.ppp, "Leptospirosis")
```


Plot Kdot
- Departures from the reference line indicate clustering or repulsion of the leptospirosis points relative to others

```{r name-of-chunk35, fig.width=20, fig.height=9}
par(mfrow=c(2,4))
plot(Kdot16, main = "2016", cex.main = 3, cex.lab = 1.5)
plot(Kdot17, main = "2017", cex.main = 3, cex.lab = 1.5)
plot(Kdot18, main = "2018", cex.main = 3, cex.lab = 1.5)
plot(Kdot19, main = "2019", cex.main = 3, cex.lab = 1.5)
plot(Kdot20, main = "2020", cex.main = 3, cex.lab = 1.5)
plot(Kdot21, main = "2021", cex.main = 3, cex.lab = 1.5)
plot(Kdot22, main = "2022", cex.main = 3, cex.lab = 1.5)
plot(Kdotall, main = "2016-2022", cex.main = 3, cex.lab = 1.5)
```


# Pair correlation function

pcf for K cross

```{r}
#| cache = TRUE
K16pcf<- pcf(K16, spar=1, method="b")
K17pcf<- pcf(K17, spar=1, method="b")
K18pcf<- pcf(K18, spar=1, method="b")
K19pcf<- pcf(K19, spar=1, method="b")
K20pcf<- pcf(K20, spar=1, method="b")
K21pcf<- pcf(K21, spar=1, method="b")
K22pcf<- pcf(K22, spar=1, method="b")
Kallpcf<- pcf(Kall, spar=1, method="b")
```



Plot pcf for K cross

```{r name-of-chunk36, fig.width=20, fig.height=9}
par(mfrow=c(2,4))
plot(K16pcf, main = "2016", cex.main = 3, cex.lab = 1.5)
plot(K17pcf, main = "2017", cex.main = 3, cex.lab = 1.5)
plot(K18pcf, main = "2018", cex.main = 3, cex.lab = 1.5)
plot(K19pcf, main = "2019", cex.main = 3, cex.lab = 1.5)
plot(K20pcf, main = "2020", cex.main = 3, cex.lab = 1.5)
plot(K21pcf, main = "2021", cex.main = 3, cex.lab = 1.5)
plot(K22pcf, main = "2022", cex.main = 3, cex.lab = 1.5)
plot(Kallpcf, main = "2016-2022", cex.main = 3, cex.lab = 1.5)
```

```{r}
#| cache = TRUE
# pcf cross
cross16pcf <- pcfcross(loc_list2016.ppp, "Leptospirosis", "Enteric_fever", stoyan=0.1)
cross17pcf <- pcfcross(loc_list2017.ppp, "Leptospirosis", "Enteric_fever", stoyan=0.1)
cross18pcf <- pcfcross(loc_list2018.ppp, "Leptospirosis", "Enteric_fever", stoyan=0.1)
cross19pcf <- pcfcross(loc_list2019.ppp, "Leptospirosis", "Enteric_fever", stoyan=0.1)
cross20pcf <- pcfcross(loc_list2020.ppp, "Leptospirosis", "Enteric_fever", stoyan=0.1)
cross21pcf <- pcfcross(loc_list2021.ppp, "Leptospirosis", "Enteric_fever", stoyan=0.1)
cross22pcf <- pcfcross(loc_list2022.ppp, "Leptospirosis", "Enteric_fever", stoyan=0.1)
crosspcf <- pcfcross(all_diag.ppp, "Leptospirosis", "Enteric_fever", stoyan=0.1)
```



plot cross pcf
- A value g(r) < 1 indicates that interpoint distances equal to r are less frequent than would be expected for a completely random process, so this suggests regularity. A value g(r) > 1 indicates that this interpoint distance is more frequent than expected for a completely random pattern, which suggests clustering.

- If the value of gr,d obs(r) is larger than the value of the gr,d high(r), then leptospirosis and enteric fever suggest a positive relationship. If the value of gr,d obs(r) is smaller than the value of the gr,d low(r), then the leptospirosis and enteric fever suggest a negative relationship. In the grey zone, we cannot reject the null hypothesis of independency of components. The variable r refers to distance from leptospirosis to enteric fever.

```{r name-of-chunk37, fig.width=20, fig.height=9}
#| cache = TRUE
par(mfrow=c(2,4))
plot(cross16pcf, main = "2016", cex.main = 3, cex.lab = 1.5)
plot(cross17pcf, main = "2017", cex.main = 3, cex.lab = 1.5)
plot(cross18pcf, main = "2018", cex.main = 3, cex.lab = 1.5)
plot(cross19pcf, main = "2019", cex.main = 3, cex.lab = 1.5)
plot(cross20pcf, main = "2020", cex.main = 3, cex.lab = 1.5)
plot(cross21pcf, main = "2021", cex.main = 3, cex.lab = 1.5)
plot(cross22pcf, main = "2022", cex.main = 3, cex.lab = 1.5)
plot(crosspcf, main = "2016-2022", cex.main = 3, cex.lab = 1.5)
```


enveloped ripley cross K
- When the observed K value is larger than the upper confidence envelope (HiConfEnv) value, spatial clustering for that distance is statistically significant. When the observed K value is smaller than the lower confidence envelope (LwConfEnv) value, spatial dispersion for that distance is statistically significant.

```{r}
#| cache = TRUE
env_K01 <- plot(envelope(all_diag.ppp, Kcross, nsim = 99))
```

```{r}
#| cache = TRUE
env_K16 <- envelope(loc_list2016.ppp, Kcross, nsim = 99)
```

```{r}
#| cache = TRUE
env_K17 <- envelope(loc_list2017.ppp, Kcross, nsim = 99)
env_K18 <- envelope(loc_list2018.ppp, Kcross, nsim = 99)
env_K19 <- envelope(loc_list2019.ppp, Kcross, nsim = 99)
env_K20 <- envelope(loc_list2020.ppp, Kcross, nsim = 99)
env_K21 <- envelope(loc_list2021.ppp, Kcross, nsim = 99)
env_K22 <- envelope(loc_list2022.ppp, Kcross, nsim = 99)

```


Plot enveloped Ripley's (cross) K by year

```{r name-of-chunk38, fig.width=20, fig.height=9}
par(mfrow=c(2,4))
plot(env_K16, main = "2016", cex.main = 3)
plot(env_K17, main = "2017", cex.main = 3)
plot(env_K18, main = "2018", cex.main = 3)
plot(env_K19, main = "2019", cex.main = 3)
plot(env_K20, main = "2020", cex.main = 3)
plot(env_K21, main = "2021", cex.main = 3)
plot(env_K22, main = "2022", cex.main = 3)
```




