---
title: "Spatial Distribution of Enteric Fever and Leptospirosis in Kelantan"
author: "Hazlienor"
date: "2023-04-12"
output: 
  html_document: 
    keep_md: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Prepare environment

Required packages 

```{r}
# install.packages("sf")
# install.packages("tidyverse")
# install.packages("here") #working directory
# install.packages("janitor")
# install.packages("gtsummary")
# install.packages("DT")
# install.packages("stringr")
# install.packages("readxl")
# install.packages("broom")
# install.packages("tmap")
# install.packages("mapview")
# install.packages("lubridate")
# install.packages("vctrs")
# install.packages("spatialECO") for NNI

```

Load packages 

```{r}
library(sf)
library(tidyverse)
library(here)
library(janitor)
library(gtsummary)
library(DT)
library(stringr)
library(readxl)
library(broom)
library(tmap)
library(mapview)
library(lubridate)
library(maptools)
library(spatstat)
library(spdep)
library(gridExtra)
library(sparr)
library(grid)
library(spatialEco)
```

## Load polygon data

read polygon data - kelantan map

```{r}
kel <- st_read(here("Map",
                    "kelantan.shp"))
```


```{r}
st_geometry(kel)
```

read population data per mukim per year

```{r}
kel_mukim <- read_xlsx(here ("mukim.xlsx"))
kel_mukim %>% datatable()
```

merge population data to polygon

```{r}
kel_map <- merge(kel,kel_mukim,by.x="MUKIM", by.y="MUKIM", all.x=T, sort=F)
dim(kel_map)
class(kel_map)
st_crs(kel_map)
```

Make plot

```{r}
st_geometry(kel_map)
plot(kel_map[,2]) # negeri
plot(kel_map[,3]) # daerah
plot(kel_map[,1]) # mukim
```


Adding facet 

```{r}
kel_map %>% ggplot() + geom_sf(aes(fill = DAERAH)) +ggtitle('Population by district') + theme_bw()
kel_map %>% ggplot() + geom_sf(aes(fill = MUKIM)) +ggtitle('Population by subdistrict') + theme_bw() + theme(legend.position = "none")
tm_shape(kel_map) + tm_polygons("AVR_FEMALE") + tm_layout(legend.width = 0.50)
tm_shape(kel_map) + tm_polygons("AVR")
tm_shape(kel_map) +
  tm_polygons(c("MUKIM", "DAERAH", "AVR_MALE")) +
  tm_facets(ncol = 3)
tm_shape(kel_map) +
  tm_polygons(c("MUKIM", "DAERAH", "AVR_MALE")) +
  tm_facets(nrow = 3)
```


## load disease data

read linelisting in .xlsx format

```{r}
linelist <- read_xlsx(here ("linelist.xlsx")) %>% clean_names()
glimpse(linelist)
```


remove na coordinate

```{r}
#disease data all - list all is all leptospirosis and enteric fever cases
listALL <- linelist %>% 
  filter(!is.na(latitude_wgs),
         !is.na(longitude_wgs))
glimpse(listALL)
```


## convert all disease data to spatial data

```{r}
loc_ALL <- st_as_sf(listALL, 
                    coords = c("longitude_wgs", "latitude_wgs"), 
                    crs = 4326)
loc_ALL %>% datatable()
```


confirm CRS is wgs84

```{r}
st_crs(loc_ALL)
```


plot map to see outlier

```{r}
ggplot() +
  geom_sf(data = loc_ALL) +
  ggtitle("Map of Leptospirosis and Enteric Fever") +
  theme_bw()
```


## convert shapefile to RSO

```{r}
loc_ALL2 <- st_transform(loc_ALL, 3168)
loc_ALL2 %>% datatable()
```


select point only in Kelantan (all_kel)

```{r}
all_kel <- loc_ALL2 %>% 
  mutate(within_kel_map = lengths(st_within(loc_ALL2, kel_map)))
all_kel2 <- all_kel %>% 
  filter(within_kel_map == 1)
```

glimpse polygon and point data

```{r}
glimpse(all_kel2)
glimpse (loc_ALL2)
```

## Descriptive 

Descriptive for all cases

```{r}
desc_all <- listALL%>%
  select(diagnosis, age, gender, race, nationality, klasifikasi_kes)
tbl_summary(desc_all)
```

Categorized age group

```{r}
desc_all$age_group <- cut(desc_all$age, breaks=c(0, 6, 9, 18, 60, Inf), 
                      labels=c("Under 5", "Children (6-9)", "Adolescent (10-19)" , "Adult (20-64)", "Elderly (65 and above)"))
```


Descriptive by diagnosis

```{r}
tbl_summary(desc_all, 
            by = diagnosis,
            label = list(age ~ "Age in years", age_group ~ "Age Category", gender ~ "Gender", race ~ "Race", nationality ~"Nationality", klasifikasi_kes ~ "Case Classification"), 
            digits = age ~ 1) %>%
  add_overall()
```


## overall plot for all cases

```{r name-of-chunk1, fig.width=18, fig.height=8}
all_plot <- ggplot() +
  geom_sf(data = kel) +
  geom_sf(data = all_kel2) +
  ggtitle("Map of Enteric Fever and Leptospirosis Cases in Kelantan for 2016-2022") +
  theme_bw()
all_plot + facet_grid(diagnosis ~ tahun_daftar) +
  theme(plot.title = element_text(size = 20),  strip.text = element_text(size = 20)) 
```


add facet

```{r}
# overall plot
all_plot + facet_wrap(~diagnosis)

# split data by diagnosis
Enteric_kel <- all_kel2 %>% 
  filter(diagnosis == "Enteric_fever") 
lepto_kel <- all_kel2 %>% 
  filter(diagnosis == "Leptospirosis") 
```


plot cumulative enteric fever cases


```{r}
Enteric_plot <- ggplot() +
  geom_sf(data = kel) +
  geom_sf(data = Enteric_kel) +
  ggtitle("Map of Enteric Fever Cases for 2016-2022") +
  theme_bw()
Enteric_plot
```


```{r name-of-chunk2, fig.width=14, fig.height=14}
Enteric_plot + 
  facet_wrap(~tahun_daftar) +
  geom_point(size=0.1) +
  theme(plot.title = element_text(size = 20),  strip.text = element_text(size = 20)) +
  ggtitle("Map of Enteric Fever Cases in Kelantan for 2016-2022")
```


plot cumulative leptospirosis cases


```{r}
lepto_kel <- all_kel2 %>% 
  filter(diagnosis == "Leptospirosis") 
lepto_plot <- ggplot() +
  geom_sf(data = kel) +
  geom_sf(data = lepto_kel) +
  ggtitle("Map of Leptospirosis Cases for 2016-2022") +
  theme_bw()
lepto_plot
```


```{r name-of-chunk3, fig.width=14, fig.height=14}
lepto_plot + 
  facet_wrap(~tahun_daftar) +
  geom_point(size=0.05) +
  theme(axis.text.x = element_text(angle = 0, hjust = 0.5, vjust = 0.5)) +
  theme(plot.title = element_text(size = 20),  strip.text = element_text(size = 20)) +
  ggtitle("Map of Leptospirosis Cases in Kelantan for 2016-2022")

```


split cumulative enteric fever cases by district

```{r name-of-chunk4, fig.width=16, fig.height=8}
# Cumulative Enteric fever cases by district
Enteric_plot + 
  facet_wrap(~DAERAH, nrow = 2) +
  theme(plot.title = element_text(size = 20),  strip.text = element_text(size = 18))
```


split cumulative leptospirosis cases by district

```{r name-of-chunk5, fig.width=16, fig.height=8}
lepto_plot + 
  facet_wrap(~DAERAH, nrow = 2) +
  theme(plot.title = element_text(size = 20),  strip.text = element_text(size = 18))
```


plot cumulative enteric fever cases by district using tmap

```{r}
tm_shape(kel) +
  tm_polygons("DAERAH", palette = rev(hcl.colors(7, "Teal"))) + 
  tm_shape(Enteric_kel) +
  tm_dots(size = 0.1) +
  tm_facets("tahun_daftar", nrow = 3, ncol = 3, as.layers = TRUE) +
  tm_layout(legend.outside = TRUE) +
  tm_layout(title = "Enteric Fever Cases in Kelantan")

```


plot cumulative leptospirosis cases by district using tmap

```{r}
#lepto plots by tmap
tm_shape(kel) +
  tm_polygons("DAERAH") + 
  tm_shape(lepto_kel) +
  tm_dots(size = 0.05) +
  tm_facets("tahun_daftar", nrow = 3, ncol = 3, as.layers = TRUE) +
  tm_layout(legend.outside = TRUE) +
  tm_layout(title = "Leptospirosis Cases in Kelantan")
```


plot combined enteric fever and leptospirosis case

```{r}
#all plot by disease
tm_shape(kel) +
  tm_polygons("NEGERI") +
  tm_facets('DAERAH') + 
  tm_shape(all_kel2) +
  tm_dots(shape = 'diagnosis', size = 0.1) 
```


## trend by year and epid week

Plot trend by year

```{r}
all_year <- all_kel2 %>% 
  group_by(diagnosis, tahun_daftar) %>% 
  count() %>% 
  print(n = 7)

ggplot(all_year, aes(x=tahun_daftar, y=n, group=diagnosis, color=as.factor(diagnosis))) +
  geom_line(size = 1) +
  labs(x="Year", y="Number of Cases", title="Enteric Fever & Leptospirosis Cases per by Year", color="Diagnosis") +
  scale_y_continuous(labels=function(x) format(x, digits=1, nsmall=0)) +
  scale_x_continuous(breaks=unique(all_year$tahun_daftar)) +
 theme_bw()
```


Plot line graph Enteric fever by epid week

```{r}
ent_week <- Enteric_kel %>% 
  group_by(tahun_daftar, epid_daftar) %>% 
  count() %>% 
  print(n = 52)

ggplot(ent_week, aes(x=epid_daftar, y=n, group=tahun_daftar, color=as.factor(tahun_daftar))) +
  geom_line(size = 1) +
  labs(x="Epid Week", y="Number of Cases", title="Enteric Fever Cases per Epid Week Separated by Year", color="Year") +
  scale_color_brewer(palette="Dark2") + 
  scale_y_continuous(labels=function(x) format(x, digits=1, nsmall=0)) +
 theme_bw()
  
```

```{r}
lep_week <- lepto_kel %>% 
  group_by(tahun_daftar, epid_daftar) %>% 
  count() %>% 
  print(n = 52)

ggplot(lep_week, aes(x=epid_daftar, y=n, group=tahun_daftar, color=as.factor(tahun_daftar))) +
  geom_line(size = 0.8) +
  labs(x="Epid Week", y="Number of Cases", title="Enteric Fever Cases per Epid Week Separated by Year", color="Year") +
  scale_color_brewer(palette="Set2") + 
  scale_y_continuous(labels=function(x) format(x, digits=1, nsmall=0)) +
 theme_bw()
```


## Density of cases

# Enteric fever density per subdistrict (mukim)

Joint point to polygon

```{r}
ent_in_muk <- st_join(Enteric_kel, kel_map, 
                      join = st_within)
glimpse(ent_in_muk)
```


count all enteric fever in mukim

```{r}
count_ent_mukim_yr <- ent_in_muk %>% 
  count(DAERAH, MUKIM, tahun_daftar, JUMLAH_2016, JUMLAH_2017, JUMLAH_2018, JUMLAH_2019, JUMLAH_2020, JUMLAH_2021, JUMLAH_2022, AVR) %>% 
  ungroup()
count_ent_mukim_yr %>% datatable()
```


Calculate incidence of enteric fever per 1000 population for mukim

```{r}
count_ent_muk_y_1000 <- count_ent_mukim_yr %>% 
  mutate(incidence_ent = (n/AVR)*1000)
count_ent_muk_y_1000 %>% datatable()
```


Calculate incidence of enteric fever per 1000 population for mukim per year +
join polygon to point

```{r}
# plot incidence enteric fever 
count_ent_mukim <- ent_in_muk %>% 
  count(DAERAH, MUKIM, AVR) %>% 
  ungroup()
count_ent_mukim_1000 <- count_ent_mukim %>% 
  mutate(inc_1000 = (n/AVR)*1000)
kelmap_with_ent <- st_join(kel_map, count_ent_mukim_1000)
glimpse(kelmap_with_ent)

# Incidence Enteric fever in 2016
count_ent_mukim_y16 <- subset(count_ent_mukim_yr, tahun_daftar=='2016') 
count_ent_mukim_y16_1000 <- count_ent_mukim_y16 %>% 
  mutate(inc_1000 = (n/JUMLAH_2016)*1000)
kelmap_with_ent_16 <- st_join(kel_map, count_ent_mukim_y16_1000)

# Incidence Enteric fever in 2017
count_ent_mukim_y17 <- subset(count_ent_mukim_yr, tahun_daftar=='2017') 
count_ent_mukim_y17_1000 <- count_ent_mukim_y17 %>% 
  mutate(inc_1000 = (n/JUMLAH_2017)*1000)
kelmap_with_ent_17 <- st_join(kel_map, count_ent_mukim_y17_1000)

# Incidence Enteric fever in 2018
count_ent_mukim_y18 <- subset(count_ent_mukim_yr, tahun_daftar=='2018') 
count_ent_mukim_y18_1000 <- count_ent_mukim_y18 %>% 
  mutate(inc_1000 = (n/JUMLAH_2018)*1000)
kelmap_with_ent_18 <- st_join(kel_map, count_ent_mukim_y18_1000)

# Incidence Enteric fever in 2019
count_ent_mukim_y19 <- subset(count_ent_mukim_yr, tahun_daftar=='2019') 
count_ent_mukim_y19_1000 <- count_ent_mukim_y19 %>% 
  mutate(inc_1000 = (n/JUMLAH_2019)*1000)
kelmap_with_ent_19 <- st_join(kel_map, count_ent_mukim_y19_1000)

# Incidence Enteric fever in 2020
count_ent_mukim_y20 <- subset(count_ent_mukim_yr, tahun_daftar=='2020') 
count_ent_mukim_y20_1000 <- count_ent_mukim_y20 %>% 
  mutate(inc_1000 = (n/JUMLAH_2020)*1000)
kelmap_with_ent_20 <- st_join(kel_map, count_ent_mukim_y20_1000)

# Incidence Enteric fever in 2021
count_ent_mukim_y21 <- subset(count_ent_mukim_yr, tahun_daftar=='2021') 
count_ent_mukim_y21_1000 <- count_ent_mukim_y21 %>% 
  mutate(inc_1000 = (n/JUMLAH_2021)*1000)
kelmap_with_ent_21 <- st_join(kel_map, count_ent_mukim_y21_1000)

# Incidence Enteric fever in 2022
count_ent_mukim_y22 <- subset(count_ent_mukim_yr, tahun_daftar=='2022') 
count_ent_mukim_y22_1000 <- count_ent_mukim_y22 %>% 
  mutate(inc_1000 = (n/JUMLAH_2021)*1000)
kelmap_with_ent_22 <- st_join(kel_map, count_ent_mukim_y22_1000)
```


plot incidence map Enteric fever 2016-2022

```{r}
incplot_ent <- ggplot() + 
  geom_sf(data = kelmap_with_ent, aes(fill = inc_1000)) +
  scale_fill_gradientn(colors = sf.colors(20)) +
  ggtitle('2016-2022') +
  theme_bw() 
incplot_ent

#plot incidence map Enteric fever 2016
incplot_ent16 <- ggplot() + 
  geom_sf(data = kelmap_with_ent_16, aes(fill = inc_1000)) +
  scale_fill_gradientn(colors = sf.colors(20)) +
  ggtitle('2016') +
  theme_bw() 
incplot_ent16

#plot incidence map Enteric fever 2017
incplot_ent17 <- ggplot() + 
  geom_sf(data = kelmap_with_ent_17, aes(fill = inc_1000)) +
  scale_fill_gradientn(colors = sf.colors(20)) +
  ggtitle('2017') +
  theme_bw()
incplot_ent17

#plot incidence map Enteric fever 2018
incplot_ent18 <- ggplot() + 
  geom_sf(data = kelmap_with_ent_18, aes(fill = inc_1000)) +
  scale_fill_gradientn(colors = sf.colors(20)) +
  ggtitle('2018') +
  theme_bw()
incplot_ent18

#plot incidence map Enteric fever 2019
incplot_ent19 <- ggplot() + 
  geom_sf(data = kelmap_with_ent_19, aes(fill = inc_1000)) +
  scale_fill_gradientn(colors = sf.colors(20)) +
  ggtitle('2019') +
  theme_bw()
incplot_ent19

#plot incidence map Enteric fever 2020
incplot_ent20 <- ggplot() + 
  geom_sf(data = kelmap_with_ent_20, aes(fill = inc_1000)) +
  scale_fill_gradientn(colors = sf.colors(20)) +
  ggtitle('2020') +
  theme_bw()
incplot_ent20

#plot incidence map Enteric fever 2021
incplot_ent21 <- ggplot() + 
  geom_sf(data = kelmap_with_ent_21, aes(fill = inc_1000)) +
  scale_fill_gradientn(colors = sf.colors(20)) +
  ggtitle('2021') +
  theme_bw()
incplot_ent21

#plot incidence map Enteric fever 2022
incplot_ent22 <- ggplot() + 
  geom_sf(data = kelmap_with_ent_22, aes(fill = inc_1000)) +
  scale_fill_gradientn(colors = sf.colors(20)) +
  ggtitle('2022') +
  theme_bw()
incplot_ent22
```


combine incidence map Enteric fever 2016-2022 in one view 

```{r name-of-chunk7, fig.width=16, fig.height=9}
grid.arrange(incplot_ent16, incplot_ent17, incplot_ent18, incplot_ent19, incplot_ent20, incplot_ent21, incplot_ent22,incplot_ent, nrow=2, top=textGrob("Incidence Map of Enteric Fever in Kelantan per 1000 population 2016-2022", gp = gpar(fontsize = 20)))
```


# Leptospirosis density per subdistrict (mukim)

Joint point to polygon
```{r}
#lepto density per mukim
lep_in_muk <- st_join(lepto_kel, kel_map, 
                      join = st_within)
glimpse(lep_in_muk)
```


count all lepto cases in mukim

```{r}
count_lep_mukim_yr <- lep_in_muk %>% 
  count(DAERAH, MUKIM, tahun_daftar, JUMLAH_2016, JUMLAH_2017, JUMLAH_2018, JUMLAH_2019, JUMLAH_2020, JUMLAH_2021, JUMLAH_2022, AVR) %>% 
  ungroup()
count_lep_mukim_yr %>% datatable()
```


Calculate incidence of leptospirosis per 1000 population for mukim+year

```{r}
count_lep_muk_y_1000 <- count_lep_mukim_yr %>% 
  mutate(incidence_lep = (n/AVR)*1000)
count_lep_muk_y_1000 %>% datatable()
```


Calculate incidence of enteric fever per 1000 population for mukim per year +
join polygon to point

```{r}
count_lep_mukim <- lep_in_muk %>% 
  count(DAERAH, MUKIM, tahun_daftar, JUMLAH_2016, JUMLAH_2017, JUMLAH_2018, JUMLAH_2019, JUMLAH_2020, JUMLAH_2021, JUMLAH_2022, AVR) %>% 
  ungroup()
count_lep_mukim_1000 <- count_lep_mukim %>% 
  mutate(inc_1000 = (n/AVR)*1000)
kelmap_with_lep <- st_join(kel_map, count_lep_mukim_1000)
glimpse(kelmap_with_lep)

# Incidence Leptospirosis in 2016
count_lep_mukim_y16 <- subset(count_lep_mukim_yr, tahun_daftar=='2016') 
count_lep_mukim_y16_1000 <- count_lep_mukim_y16 %>% 
  mutate(inc_1000 = (n/JUMLAH_2016)*1000)
kelmap_with_lep_16 <- st_join(kel_map, count_lep_mukim_y16_1000)

# Incidence Leptospirosis in 2017
count_lep_mukim_y17 <- subset(count_lep_mukim_yr, tahun_daftar=='2017') 
count_lep_mukim_y17_1000 <- count_lep_mukim_y17 %>% 
  mutate(inc_1000 = (n/JUMLAH_2017)*1000)
kelmap_with_lep_17 <- st_join(kel_map, count_lep_mukim_y17_1000)

# Incidence Leptospirosis in 2018
count_lep_mukim_y18 <- subset(count_lep_mukim_yr, tahun_daftar=='2018') 
count_lep_mukim_y18_1000 <- count_lep_mukim_y18 %>% 
  mutate(inc_1000 = (n/JUMLAH_2018)*1000)
kelmap_with_lep_18 <- st_join(kel_map, count_lep_mukim_y18_1000)

# Incidence Leptospirosis in 2019
count_lep_mukim_y19 <- subset(count_lep_mukim_yr, tahun_daftar=='2019') 
count_lep_mukim_y19_1000 <- count_lep_mukim_y19 %>% 
  mutate(inc_1000 = (n/JUMLAH_2019)*1000)
kelmap_with_lep_19 <- st_join(kel_map, count_lep_mukim_y19_1000)

# Incidence Leptospirosis in 2020
count_lep_mukim_y20 <- subset(count_lep_mukim_yr, tahun_daftar=='2020') 
count_lep_mukim_y20_1000 <- count_lep_mukim_y20 %>% 
  mutate(inc_1000 = (n/JUMLAH_2020)*1000)
kelmap_with_lep_20 <- st_join(kel_map, count_lep_mukim_y20_1000)

# Incidence Leptospirosis in 2021
count_lep_mukim_y21 <- subset(count_lep_mukim_yr, tahun_daftar=='2021') 
count_lep_mukim_y21_1000 <- count_lep_mukim_y21 %>% 
  mutate(inc_1000 = (n/JUMLAH_2021)*1000)
kelmap_with_lep_21 <- st_join(kel_map, count_lep_mukim_y21_1000)

# Incidence Leptospirosis in 2022
count_lep_mukim_y22 <- subset(count_lep_mukim_yr, tahun_daftar=='2022') 
count_lep_mukim_y22_1000 <- count_lep_mukim_y22 %>% 
  mutate(inc_1000 = (n/JUMLAH_2020)*1000)
kelmap_with_lep_22 <- st_join(kel_map, count_lep_mukim_y22_1000)
```


plot incidence map Leptospirosis 2016-2022

```{r}
# plot cumulative
incplot_lep <- ggplot() + 
  geom_sf(data = kelmap_with_lep, aes(fill = inc_1000)) +
  scale_fill_gradientn(colors = sf.colors(20)) +
  ggtitle('2016-2022') +
  theme_bw()
incplot_lep 

# plot incidence map Leptospirosis 2016
incplot_lep16 <- ggplot() + 
  geom_sf(data = kelmap_with_lep_16, aes(fill = inc_1000)) +
  scale_fill_gradientn(colors = sf.colors(20)) +
  ggtitle('2016') +
  theme_bw()
incplot_lep16

#plot incidence map Leptospirosis 2017
incplot_lep17 <- ggplot() + 
  geom_sf(data = kelmap_with_lep_17, aes(fill = inc_1000)) +
  scale_fill_gradientn(colors = sf.colors(20)) +
  ggtitle('2017') +
  theme_bw()
incplot_lep17

#plot incidence map Leptospirosis 2018
incplot_lep18 <- ggplot() + 
  geom_sf(data = kelmap_with_lep_18, aes(fill = inc_1000)) +
  scale_fill_gradientn(colors = sf.colors(20)) +
  ggtitle('2018') +
  theme_bw()
incplot_lep18

#plot incidence map Leptospirosis 2019
incplot_lep19 <- ggplot() + 
  geom_sf(data = kelmap_with_lep_19, aes(fill = inc_1000)) +
  scale_fill_gradientn(colors = sf.colors(20)) +
  ggtitle('2019') +
  theme_bw()
incplot_lep19

#plot incidence map Leptospirosis 2020
incplot_lep20 <- ggplot() + 
  geom_sf(data = kelmap_with_lep_20, aes(fill = inc_1000)) +
  scale_fill_gradientn(colors = sf.colors(20)) +
  ggtitle('2020') +
  theme_bw()
incplot_lep20

#plot incidence map Leptospirosis 2021
incplot_lep21 <- ggplot() + 
  geom_sf(data = kelmap_with_lep_21, aes(fill = inc_1000)) +
  scale_fill_gradientn(colors = sf.colors(20)) +
  ggtitle('2021') +
  theme_bw()
incplot_lep21

#plot incidence map Leptospirosis 2022
incplot_lep22 <- ggplot() + 
  geom_sf(data = kelmap_with_lep_22, aes(fill = inc_1000)) +
  scale_fill_gradientn(colors = sf.colors(20)) +
  ggtitle('2022') +
  theme_bw()
incplot_lep22
```


combine incidence maps Leptospirosis 2016-2022 in one view

```{r name-of-chunk8, fig.width=16, fig.height=9}
# combine all leptospirosis incidence plot  in one view
grid.arrange(incplot_lep16, incplot_lep17, incplot_lep18, incplot_lep19, incplot_lep20, incplot_lep21, incplot_lep22,incplot_lep, nrow=2, top=textGrob("Incidence Map of Leptospirosis in Kelantan per 1000 population 2016-2022", gp = gpar(fontsize = 20)))

```

